### 2023-12-26 main.tex
organisation

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Type theory in type theory using single substitutions} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
Type theory can be described as a generalised algebraic theory which
automatically gives a notion of model and the existence of the syntax
as the initial model given by a quotient inductive-inductive
type. Such descriptions include categories with families (CwF),
contextual categories, Awodey's natural models, C-systems and
B-systems. With the exception of B-systems, these theories are based on
a parallel substitution calculus where substitutions form a
category. In this paper we define a single substitution calculus (SSC)
for type theory and show that the SSC syntax and the CwF syntax are
isomorphic for a theory with dependent function space and universes.
SSC only includes single substitutions and single weakenings, and six
equations relating these. Four equations describe how to substitute
variables, and two more equations are needed to typecheck these four
equations. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. All the
results were formalised in Agda.
\end{abstract}

\input{inputs.tex}

\bibliography{b}

\end{document}

### 2023-12-26 main.tex
stuff

@@
 
-\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ambrus Kaposi}{Eötvös Loránd University, Budapest}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+\author{Szumi Xie}{Eötvös Loránd University, Budapest}{szumi@inf.elte.hu}{[orcid]}{}
 
-\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{A. Kaposi and Szumi Xie} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Ambrus Kaposi and Szumi Xie} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\ccsdesc[500]{Theory of computation~Logic~Type theory}
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{type theory, substitution calculus, category with families, generalised algebraic theory} %TODO mandatory; please add comma-separated list of keywords
 

### 2023-12-27 main.tex
stuff

@@
 \begin{abstract}
+TODO: rewrite the abstract to start with ``tutorial intro''.
+
 Type theory can be described as a generalised algebraic theory which

### 2023-12-27 main.tex
organise
stuff

@@
 
+\usepackage{agda}
+
 %\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
@@
 
-\author{Ambrus Kaposi}{Eötvös Loránd University, Budapest}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ambrus Kaposi}{Eötvös Loránd University, Budapest}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}
+       {The author was supported by project no.
+TKP2021-NVA-29 which has been implemented with the support provided by the Ministry of Culture and Innovation of Hungary from the National Research, Development and Innovation Fund, financed under the TKP2021-NVA funding scheme.}
+       %TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.

### 2024-03-05 types_2024/abstract.tex
stuff

% easychair.tex,v 3.5 2017/03/15

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{doc}
\usepackage{proof}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{newtx}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Type theory in type theory using single substitutions}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Ambrus Kaposi\inst{1}
\and
Szumi Xie\inst{1}
}

% Institutes for affiliations are also joined by \and,
\institute{
   Eötvös Loránd University,
   Budapest, Hungary, 
   \email{\{akaposi|szumi\}@inf.elte.hu}
}

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

\authorrunning{Kaposi, Xie}

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Type theory in type theory using single substitutions}

\begin{document}

\maketitle

\newcommand{\Ra}{\Rightarrow}
\newcommand{\ra}{\rightarrow}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\var}{\mathsf{var}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\renewcommand{\c}{\mathsf{c}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\oldapp}{\mathbin{\cdot}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Br}{\mathsf{Br}}
\newcommand{\aps}{\mathsf{aps}}
\newcommand{\ap}{\mathsf{ap}}
\newcommand{\mkBrPi}{\mathsf{mkBr}{\Pi}}
\newcommand{\Gel}{\mathsf{Gel}}
\newcommand{\gel}{\mathsf{gel}}
\newcommand{\ungel}{\mathsf{ungel}}
\renewcommand{\sp}{\hspace{1.5em}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\R}{\mathsf{R}}
\newcommand{\p}{\mathsf{p}}
\renewcommand{\S}{\mathsf{S}}
\newcommand{\vz}{\mathsf{vz}}
\newcommand{\vs}{\mathsf{vs}}

In this abstract we define type theory in a minimalistic way: we aim
for an intrinsic, that is, well-typed and quotiented definition of the
syntax, that is as small as possible in terms of needed operations and
equations for the substitution calculus. We only introduce operations
that are \emph{forced} upon us. We aim for a language with $\Pi$
types, and it turns out that all the essential ingredients are forced
upon us when we want to introduce the usual rules for $\Pi$ types
(abstraction, application, $\beta$ and $\eta$).

We need variables in our language, so we introduce sorts of contexts,
types and variables.
\[
 \Con : \Set \hspace{6em} \Ty : \Con\ra\Set \hspace{6em} \Var : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set  \\
\]
Contexts are either empty or extended with a type.
\[
\diamond : \Con \hspace{6em} \blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con
\]
We define variables as well-typed De Bruijn indices, but to express
these we need to weaken types: e.g.\ the zero De Bruijn index $\vz$
has a weakened type. We introduce a new sort for substitutions $\Sub$,
am instantiation operation $\blank[\blank]$ on types, and a weakening
substitution $\p$. For now, $\Sub$ seems like an overkill because we
are only using $\blank[\p]$, but it will come handy soon.
\[
\Sub: \Con\ra\Con\ra\Set \hspace{3em} \blank[\blank] : \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta \hspace{3em} \p : \Sub\,(\Gamma\ext A)\,\Gamma \hspace{3em}
\]
\[
\vz : \Var\,(\Gamma\ext A)\,(A[\p]) \hspace{3em} \vs : \Var\,\Gamma\,A\ra\Var\,(\Gamma\ext B)\,(A[\p])
\]
Now we introduce $\Pi$ types together with an equation on how
instantiation with $\p$ acts on them. This is tricky: as $\Pi$ binds a
new variable in its second argument, we need a new version of the
weakening substitution which leaves the last variable untouched. This
is why we introduce lifting of a substitution $\blank^{+}$, and now we
can state a general instantiation rule for $\Pi$ which works not only
for $\p$, but arbitrary substitutions.
\begin{alignat*}{10}
& \Pi:(A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \hspace{3em} && \blank^+ : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
& \Pi[] : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
\end{alignat*}
In addition to having variables, we need a sort of terms which
includes variables and lambda abstraction.
\[
\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Set \hspace{1.8em} \var : \Var\,\Gamma\,A\ra\Tm\,\Gamma\,A \hspace{1.8em} \lam : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B)  \\
\]
To express application, we need single substitutions as well because
the argument of the function appears in the return type. In addition
to $\p$ and $\blank^+$, $\langle\blank\rangle$ is the third and last
operation for creating substitutions.
\[
\langle\blank\rangle : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \hspace{5em} \blank\cdot\blank: \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle])
\]
Now we would like to express the $\beta$ computation rule, but for
this we also need to be able to instantiate terms (in addition to
types).
\[
\blank[\blank] : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \hspace{3em} \Pi\beta : \lam\,t\cdot a = t[\langle a\rangle] 
\]
Now that we introduced instantiation of terms, we have to revisit all
of our terms and provide rules on how to instantiate them: first of
all, we need instantiation rules for $\lam$ and
$\blank\cdot\blank$. The rule $\lam[]$ is well-typed because of
$\Pi[]$, however ${\cdot}[]$ is not well-typed on its own and requires
a new equation $[\langle\rangle]$.
\[
\lam[] : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \hspace{1em} [\langle\rangle] : A[\langle a\rangle][\gamma] = A[\gamma^+][\langle a[\gamma]\rangle] \hspace{1em} {\cdot}[] : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
\]
Then we need instantiation rules for variables, we list these by case
distinction on the substitution: weakening of a variable increases the
index by one. When instantiating with lifted substitutions and single
substitutions, we have to do case distinction on the De Bruijn index
whether it is zero or successor. For the latter two cases, we need
type equations (named $[\p][^+]$ and $[\p][\langle\rangle]$) to
typeckeck the term equations.\vspace{-1.5em}
\begin{alignat*}{10}
  & && \var\,x[\p] = \var\,(\vs\,x) \\
  & [\p][^+] : A[\p][\gamma^+] = A[\gamma][\p] \hspace{4em}&& \var\,\vz[\gamma^+] = \var\,\vz \hspace{4em} && \var\,(\vs\,x)[\gamma^+] = \var\,x[\gamma][\p] \\
  & [\p][\langle\rangle] : A[\p][\langle a\rangle] = A && \var\,\vz[\langle a\rangle] = a &&  \var\,(\vs\,x)[\langle a\rangle] = \var\,x
\end{alignat*}
Finally, to typecheck the $\Pi\eta$ rule, we need our last equation on
types.
\[
[^+][\langle\rangle] : A[p^+][\langle\var\,\vz\rangle] = A \hspace{6em}\Pi\eta : t = \lam\,(t[\p]\cdot\var\,\vz)
\]
This concludes all the rules for type theory with $\Pi$: we summarise
as follows: there are three kinds of substitutions (single weakening,
single substitution, lifted substitution), we have instantiation rule
for each operation, there are 5 equations describing how instantiation
acts on variables, and 4 equations which describe general properties
of instantiation on types. The rest of the rules are specific to our
single type former $\Pi$: the only extra requirement is that each
operation should come equipped with an instantiation rule ($\Pi[]$,
$\lam[]$, ${\cdot}[]$). Perhaps surprisingly, this is enough to define
the syntax: there is no need for $\Con$ and $\Sub$ to a form a
category, no need for parallel substitutions, empty substitution,
parallel weakenings, or combinations of these.

We can add more type formers...

Show how to derive CwF.

\label{sect:bib}
\bibliographystyle{plainurl}
%\bibliographystyle{alpha}
%\bibliographystyle{unsrt}
%\bibliographystyle{abbrv}
\bibliography{b}

%------------------------------------------------------------------------------
\end{document}

### 2024-03-05 types_2024/abstract.tex
stuff

@@
 \author{
-Ambrus Kaposi\inst{1}
+Ambrus Kaposi
 \and
-Szumi Xie\inst{1}
+Szumi Xie
 }
@@
 for an intrinsic, that is, well-typed and quotiented definition of the
-syntax, that is as small as possible in terms of needed operations and
-equations for the substitution calculus. We only introduce operations
-that are \emph{forced} upon us. We aim for a language with $\Pi$
-types, and it turns out that all the essential ingredients are forced
-upon us when we want to introduce the usual rules for $\Pi$ types
-(abstraction, application, $\beta$ and $\eta$).
+syntax, that is as small as possible in terms of number of operations
+and equations for the substitution calculus. We only introduce
+operations that are forced upon us. We aim for a language with $\Pi$
+types, and it turns out that we have to add all the ingredients of the
+substitution calculus while introducing the usual rules for $\Pi$
+types (abstraction, application, $\beta$ and $\eta$). When adding more
+type formers, we don't need to extend the substitution calculus
+anymore.
 
@@
 \]
-Contexts are either empty or extended with a type.
+Contexts are either empty or are built from a context extended with a
+type.
 \[
@@
 has a weakened type. We introduce a new sort for substitutions $\Sub$,
-am instantiation operation $\blank[\blank]$ on types, and a weakening
+an instantiation operation $\blank[\blank]$ on types, and a weakening
 substitution $\p$. For now, $\Sub$ seems like an overkill because we
@@
 can state a general instantiation rule for $\Pi$ which works not only
-for $\p$, but arbitrary substitutions.
+for $\p$, but arbitrary substitutions (including lifted ones).
 \begin{alignat*}{10}
@@
 \]
-Now that we introduced instantiation of terms, we have to revisit all
-of our terms and provide rules on how to instantiate them: first of
-all, we need instantiation rules for $\lam$ and
+Now that we have instantiation of terms, we need to revisit all
+operations producing terms and provide rules on how to instantiate
+them: first of all, we need instantiation rules for $\lam$ and
 $\blank\cdot\blank$. The rule $\lam[]$ is well-typed because of
@@
 \]
-Then we need instantiation rules for variables, we list these by case
-distinction on the substitution: weakening of a variable increases the
-index by one. When instantiating with lifted substitutions and single
+Then we need instantiation rules for variables, we list these for each
+possible substitution: weakening of a variable increases the index by
+one; when instantiating with lifted substitutions and single
 substitutions, we have to do case distinction on the De Bruijn index
@@
 type equations (named $[\p][^+]$ and $[\p][\langle\rangle]$) to
-typeckeck the term equations.\vspace{-1.5em}
+typeckeck the term equations.
 \begin{alignat*}{10}
@@
 \[
-[^+][\langle\rangle] : A[p^+][\langle\var\,\vz\rangle] = A \hspace{6em}\Pi\eta : t = \lam\,(t[\p]\cdot\var\,\vz)
+[\p^+][\langle\vz\rangle] : A[\p^+][\langle\var\,\vz\rangle] = A \hspace{6em}\Pi\eta : t = \lam\,(t[\p]\cdot\var\,\vz)
 \]
-This concludes all the rules for type theory with $\Pi$: we summarise
+This concludes all the rules for type theory with $\Pi$. We summarise
 as follows: there are three kinds of substitutions (single weakening,
-single substitution, lifted substitution), we have instantiation rule
-for each operation, there are 5 equations describing how instantiation
-acts on variables, and 4 equations which describe general properties
-of instantiation on types. The rest of the rules are specific to our
-single type former $\Pi$: the only extra requirement is that each
-operation should come equipped with an instantiation rule ($\Pi[]$,
-$\lam[]$, ${\cdot}[]$). Perhaps surprisingly, this is enough to define
-the syntax: there is no need for $\Con$ and $\Sub$ to a form a
-category, no need for parallel substitutions, empty substitution,
-parallel weakenings, or combinations of these.
+single substitution, lifted substitution), we have 5 equations
+describing how instantiation acts on variables, and 4 equations which
+describe general properties of instantiation on types. The rest of the
+rules are specific to our single type former $\Pi$: the only extra
+requirement is that each operation is equipped with an instantiation
+rule ($\Pi[]$, $\lam[]$, ${\cdot}[]$). Perhaps surprisingly, this is
+enough to define the syntax: there is no need for $\Con$ and $\Sub$ to
+a form a category, no need for parallel substitutions, empty
+substitution, parallel weakenings, or other combinations of these.
 
-We can add more type formers...
+When adding new type formers, we only need the rules for the type
+former, and an extra instantiation (naturality) rule for each
+operation. For example, a Coquand-universe can be added by $\U :
+\Ty\,\Gamma$, $\El:\Tm\,\Gamma\,\U\ra\Ty\,\Gamma$,
+$\c:\Ty\,\Gamma\ra\Tm\,\Gamma\,\U$, $\U\beta : \El\,(\c\,A) = A$,
+$\U\eta : \c\,(\El\,a) = a$, and three instantiation rules (note that
+we need indexing to avoid inconsistency). In \cite{sogat}, we showed
+that any second-order generalised algebraic theory (SOGAT) has a
+single substitution presentation.
 
-Show how to derive CwF.
+In the syntax (initial model, quotient inductive-inductive type) of
+the above theory, all the rules of categories with families (CwF
+\cite{DBLP:journals/corr/abs-1904-00827}) are admissible. That is, by
+induction on the single substitution syntax, we can define parallel
+substitutions (lists of terms) which are composable and form a
+category; we can define instantiation by parallel substitutions for
+types and terms, these have the usual universal property of
+comprehension. The main ingredient for this construction is the notion
+of $\alpha$-normal form: a kind of normal form which is still
+quotiented by $\Pi\beta$, $\Pi\eta$, but does not include explicit
+substitutions. If a type is in $\alpha$-normal form, we know whether
+it is $\Pi$, $\U$, or $\El$ of a term. If a term is in $\alpha$-normal
+form, we know whether it is a variable, a $\lam$, an application or a
+code for a type (note that any function on an $\alpha$-normal has to
+respect $\Pi\beta$, $\Pi\eta$, $\U\beta$, $\U\eta$). We prove
+$\alpha$-normalisation (every term has a unique $\alpha$-normal form)
+and then use induction on $\alpha$-normal forms to define parallel
+instantiation and prove its properties.
+
+We formalised the single substitution calculus with an infinite
+hierarchy of types closed under $\Pi$ and $\U$ as a quotient
+inductive-inductive type in Agda, proved $\alpha$-normalisation for
+it, and derived all the rules for the parallel substitution calculus
+(CwF equipped with $\Pi$ types and $\U$).
+
+It is clear that the rules for the single substitution calculus are
+all derivable from the CwF-rules. The other direction is however not
+true: there are more models of the single substitution calculus than
+the parallel substitution calculus. The situation is analogous to the
+relationship of combinatory logic and lambda calculus: the former has
+more models, but the syntaxes are isomorphic
+\cite{DBLP:conf/fscd/AltenkirchKSV23}. Our single substitution
+calculus is a minimalistic version of B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} (C-systems are their
+parallel substitution analogue): B-systems include telescopes and more
+equations than our single substitution calculus, making them
+equivalent to C-systems.
 

### 2024-03-07 types_2024/abstract.tex
finalise abstract

@@
 
-We formalised the single substitution calculus with an infinite
-hierarchy of types closed under $\Pi$ and $\U$ as a quotient
-inductive-inductive type in Agda, proved $\alpha$-normalisation for
-it, and derived all the rules for the parallel substitution calculus
-(CwF equipped with $\Pi$ types and $\U$).
+We formalised single substitution calculus with an infinite hierarchy
+of types closed under $\Pi$ and $\U$ as a quotient inductive-inductive
+type in Agda, proved $\alpha$-normalisation, and derived all the rules
+for parallel substitutions (CwF equipped with $\Pi$ types and $\U$):
+\url{https://bitbucket.org/akaposi/single}.
 

### 2024-05-07 types_2024/abstract.tex
typo
final

@@
 
-In this abstract we define type theory in a minimalistic way: we aim
-for an intrinsic, that is, well-typed and quotiented definition of the
-syntax, that is as small as possible in terms of number of operations
-and equations for the substitution calculus. We only introduce
-operations that are forced upon us. We aim for a language with $\Pi$
-types, and it turns out that we have to add all the ingredients of the
-substitution calculus while introducing the usual rules for $\Pi$
-types (abstraction, application, $\beta$ and $\eta$). When adding more
-type formers, we don't need to extend the substitution calculus
-anymore.
+What is the syntax of Martin-Löf type theory? In this abstract, we
+give a minimalistic answer to this question. We aim to define the
+syntax of type theory only using operations that are unavoidable. We
+would also like to eschew boilerplate by only defining well-typed terms
+which are quotiented by conversion
+\cite{DBLP:conf/popl/AltenkirchK16}. This means that type theory is a
+generalised algebraic theory (GAT), its syntax is a quotient
+inductive-inductive type (QIIT). Usual such definitions of type theory are
+based on categories: category with families \cite{DBLP:journals/corr/abs-1904-00827}, locally cartesian closed
+category \cite{Seely1984-SEELCC,DBLP:journals/mscs/ClairambaultD14}, contextual category \cite{DBLP:journals/apal/Cartmell86,initiality-agda}, C systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}, etc.
+In this abstract we
+show that categories are not necessary for defining the GAT of type
+theory. A category-free definition of type theory is B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which
+includes telescopes with complex operations and equations. In our
+definition we avoid these as well. In this abstract we give a
+tutorial-style introduction to our minimalistic definition of type
+theory, no prior knowledge of the metatheory of type theory is
+required. We introduce the operations in a naive, logical order.
 
 We need variables in our language, so we introduce sorts of contexts,
-types and variables.
+types (which depend on a context) and variables (which are in a context and have a type).
 \[
@@
 an instantiation operation $\blank[\blank]$ on types, and a weakening
-substitution $\p$. For now, $\Sub$ seems like an overkill because we
+substitution $\p$. For now, a separate sort $\Sub$ seems like an overkill because we
 are only using $\blank[\p]$, but it will come handy soon.
@@
 Then we need instantiation rules for variables, we list these for each
-possible substitution: weakening of a variable increases the index by
+possible substitution $\Sub$: weakening of a variable increases the index by
 one; when instantiating with lifted substitutions and single
@@
 \end{alignat*}
-Finally, to typecheck the $\Pi\eta$ rule, we need our last equation on
+Finally, to typecheck the $\Pi\eta$ rule, we need our last equations on
 types.
@@
 \]
-This concludes all the rules for type theory with $\Pi$. We summarise
+This concludes all the rules for type theory with $\Pi$ (this type theory is actually empty because there are no base types, but is enough to illustrate our method). We summarise
 as follows: there are three kinds of substitutions (single weakening,
@@
 a form a category, no need for parallel substitutions, empty
-substitution, parallel weakenings, or other combinations of these.
+substitution, parallel weakenings, telescopes, or combinations of these.
 
@@
 $\U\eta : \c\,(\El\,a) = a$, and three instantiation rules (note that
-we need indexing to avoid inconsistency). In \cite{sogat}, we showed
-that any second-order generalised algebraic theory (SOGAT) has a
+we would need indexing $\U$ and $\Ty$ by universe levels to avoid inconsistency). In \cite{sogat}, we showed
+that any second-order generalised algebraic theory has a
 single substitution presentation.
 
-In the syntax (initial model, quotient inductive-inductive type) of
+In the syntax (initial model, QIIT) of
 the above theory, all the rules of categories with families (CwF
@@
 form, we know whether it is a variable, a $\lam$, an application or a
-code for a type (note that any function on an $\alpha$-normal has to
+code for a type (note that any function on an $\alpha$-normal type/term has to
 respect $\Pi\beta$, $\Pi\eta$, $\U\beta$, $\U\eta$). We prove
@@
 
-We formalised single substitution calculus with an infinite hierarchy
-of types closed under $\Pi$ and $\U$ as a quotient inductive-inductive
-type in Agda, proved $\alpha$-normalisation, and derived all the rules
-for parallel substitutions (CwF equipped with $\Pi$ types and $\U$):
-\url{https://bitbucket.org/akaposi/single}.
+We formalised a single substitution calculus with an infinite hierarchy
+of types closed under $\Pi$ and $\U$ as a QIIT in Agda, proved
+$\alpha$-normalisation, and derived all the rules for parallel
+substitutions (CwF equipped with $\Pi$ types and $\U$):
+\url{https://bitbucket.org/akaposi/single}. In the formalisation, we
+use $\mathsf{sProp}$-valued equality and the syntax is postulated as a
+QIIT with rewrite rules for its $\beta$ laws.
 
@@
 true: there are more models of the single substitution calculus than
-the parallel substitution calculus. The situation is analogous to the
-relationship of combinatory logic and lambda calculus: the former has
-more models, but the syntaxes are isomorphic
-\cite{DBLP:conf/fscd/AltenkirchKSV23}. Our single substitution
-calculus is a minimalistic version of B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} (C-systems are their
-parallel substitution analogue): B-systems include telescopes and more
-equations than our single substitution calculus, making them
-equivalent to C-systems.
+the parallel substitution calculus, but the syntaxes are
+isomorphic. The situation is analogous to the relationship of
+combinatory logic and lambda calculus: their syntaxes are isomorphic,
+but the former has more models \cite{DBLP:conf/fscd/AltenkirchKSV23}.
 

### 2024-08-22 types_2024_postproc/p.tex
stuff

\documentclass[a4paper]{article}

\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}

\begin{document}

\title{Type theory in type theory with single substitutions}
\author{Ambrus Kaposi and Szumi Xie}
\date{\today}
\maketitle

\begin{abstract}
Type theory can be described as a generalised algebraic theory which
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Such descriptions include Ehrhard's definition of model,
categories with families (CwF), contextual categories, Awodey's
natural models, C-systems and B-systems. With the exception of
B-systems, these notions are based on a parallel substitution calculus
where substitutions form a category. In this paper we define a single
substitution calculus (SSC) for type theory and show that the SSC
syntax and the CwF syntax are isomorphic for a theory with dependent
function space and universe. SSC only includes single substitutions
and single weakenings, and six equations relating these. Four
equations describe how to substitute variables, and two more equations
are needed to typecheck these four equations. SSC provides a simple,
minimalistic alternative to parallel substitution calculi or B-systems
for defining type theory. All the results in this paper were
formalised in Agda.
\end{abstract}

\section{Introduction}

\cite{singleTypes}
\cite{coquandEhrhard}

What is type theory? This term can refer to formal systems (languages,
type systems, type theories) or to the field of study of type
theories. In this paper, we will give an answer to this question in
the form of a concrete formal system. We aim for simplicity and ease
of understanding: we only require a working knowledge of Agda, a
functional programming language which is an implementation of type
theory. Our metatheory in which this paper is written is thus type
theory, and everything is formalised in Agda.

The traditional way of defining a programming language formally
\cite{DBLP:books/daglib/0005958,DBLP:books/cu/Ha2016,Pierce:SF1,plfa22.08}
is comprised of three steps: (i) definition of the presyntax
(abstract syntax trees, ASTs) as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be plugged together in well-typed ways; (iii) terms
are quotiented by conversion rules, making convertible terms
equal. Inductive types which allow such mutual indexing and
quotienting are called quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}). As Agda does not support them
directly, we postulate their constructors and induction principle and add the
computation rules using rewrite rules \cite{DBLP:conf/types/Cockx19}.

A QIIT is also the initial algebra/model of a generalised algebraic
theory (GAT \cite{DBLP:journals/apal/Cartmell86}). The induction principle of
the QIIT is equivalent to having a unique morphism into any model
\cite{DBLP:journals/pacmpl/KaposiKA19}. There are several ways to
define type theory as a GAT: categories with families (CwFs
\cite{DBLP:journals/corr/abs-1904-00827}), contextual categories
\cite{DBLP:journals/apal/Cartmell86}, natural models
\cite{DBLP:journals/mscs/Awodey18}, C-systems and B-systems
\cite{ahrens_emmenegger_north_rijke_2023}. When extracting a QIIT from
a GAT, with the exception of B-systems, we obtain parallel
substitution calculi. Like B-systems, in this paper we define type
theory using \emph{single substitutions}. Our definition is simpler
than B-systems, but it is closely related to it. It is a minimalistic
presentation where every operation and equation is well-motivated (see
Section \ref{sec:SSC}). However, when we want to use our syntax (for
example, if we want to prove canonicity or normalisation for our
theory), it is much more convenient to use parallel substitution
calculi such as CwFs. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}). Then we show how to
derive the rules of a CwF from our syntax (Section
\ref{sec:cwf}). Finally, we show that our syntax is isomorphic to the
CwF-style syntax (Section \ref{sec:iso}). We list related work and
conclude in Section \ref{sec:conclusion}.

Our main contribution is the novel definition of a minimalistic single
substitution calculus, the $\alpha$-normalisation algorithm which can
be useful for other explicit substitution calculi, and finally our
proof of equivalence with the CwF-style syntax.

\subsection{Formalisation}

Notation and formalisation techniques.

\subsection{Structure of the paper}

\section{Single substitution syntax}
\label{sec:ssc}

Contents of the TYPES abstract, but with more details.

\section{The single substitution syntax is a category with families}

In this section we show that a category with families (CwF) equipped
with a hierarchy of Coquand-universes and $\Pi$ types can be derived
from the single substitution syntax defined in \autoref{sec:ssc}.

\section{Related work and conclusion}

In this paper we introduced a new single substitution calculus which
can be seen as a minimalistic variant of B-systems. We illustrate its
usage through the definition of a type theory with dependent function
space and a hierarchy of universes. We aim for accessibility: this
paper can be seen as a tutorial introduction to the syntax of type
theory inside type theory. All the operations and equations are
well-motivated and provide a minimalistic description of the syntax of
type theory. It is minimalistic in the sense that the syntax is
isomorphic to the CwF-syntax, but there are more models.

The relationship of our single substitution calculus and the CwF-based
theory is similar to that of combinatory logic and lambda calculus
\cite{DBLP:conf/fscd/AltenkirchKSV23}. Combinatory logic has more
models than lambda calculus, but the syntaxes are
isomorphic. Similarly, the single substitution calculus has more
models, but the syntaxes are isomorphic.

SSC was used in the 90s for simple type theories, ask Hugo for references.

A more abstract, higher level description of a language is simply
stating that a language is an algebraic theory: the terms of the
language are given by a sort of the theory, the different term formers
are operations, and the conversion relation is specified by the
equations. Such a description contains less information than the
traditional one, for example, conversion is not oriented, so it is not
clear how to turn it into a rewriting relation; we cannot choose how
much annotations we add to terms, they are always fully
annotated. this is similar to how asts contain less information than
strings, for example, the bracket-removal strategy is not encoded in
asts. the algebraic description automatically induces a notion of
model and the existence of an initial model
\cite{DBLP:journals/pacmpl/KaposiKA19} which we call the syntax. The
algebraic approach is especially natural for languages with dependent
types where typing and conversion have to be specified mutually
\cite{DBLP:conf/popl/AltenkirchK16}.

QIIT is nice because syntax and semantics are given together. The
notion of model/semantics is traditionally given separately, and work
has to be done to show that the above specified syntax generates a
model. For example, this was the goal of the initiality project for
homotopy type theory \cite{brunerie}. Maybe list work on type theory
using QIITs, see TYPES mailing list.

We can abstract over these strategies and define our language as a
second-order generalised algebraic theory
\cite{DBLP:journals/corr/abs-1904-04097}, but to state initiality of
our syntax, at some point we need to turn it into a first-order
algebraic theory \cite{DBLP:conf/fscd/BocquetKS23}.

We conjecture that any SOGAT can be turned into an SSC.

Maybe we can do without the extra type equations, by using assumptions.

Luksa, Thorsten thesis.

\bibliography{b}{}
\bibliographystyle{plain}

\end{document}

### 2024-09-07 types_2024_postproc/p.tex
stuff

@@
-\documentclass[a4paper]{article}
+\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 
-\usepackage{cite}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{hyperref}
+\newcommand{\Ty}{\mathsf{Ty}}
+\newcommand{\Tm}{\mathsf{Tm}}
+\newcommand{\p}{\mathsf{p}}
+\newcommand{\q}{\mathsf{q}}
+\newcommand{\ext}{\mathop{\triangleright}}
 
@@
 \date{\today}
-\maketitle
 
 \begin{abstract}
-Type theory can be described as a generalised algebraic theory which
+Type theory can be described as a generalised algebraic theory. This
 automatically gives a notion of model and the existence of the syntax
 as the initial model, which is a quotient inductive-inductive
-type. Such descriptions include Ehrhard's definition of model,
-categories with families (CwF), contextual categories, Awodey's
-natural models, C-systems and B-systems. With the exception of
-B-systems, these notions are based on a parallel substitution calculus
-where substitutions form a category. In this paper we define a single
-substitution calculus (SSC) for type theory and show that the SSC
-syntax and the CwF syntax are isomorphic for a theory with dependent
-function space and universe. SSC only includes single substitutions
-and single weakenings, and six equations relating these. Four
-equations describe how to substitute variables, and two more equations
-are needed to typecheck these four equations. SSC provides a simple,
-minimalistic alternative to parallel substitution calculi or B-systems
-for defining type theory. All the results in this paper were
-formalised in Agda.
+type. Algebraic definitions of type theory include Ehrhard's
+definition of model, categories with families (CwFs), contextual
+categories, Awodey's natural models, C-systems, B-systems. With the
+exception of B-systems, these notions are based on a parallel
+substitution calculus where substitutions form a category. In this
+paper we define a single substitution calculus (SSC) for type theory
+and show that the SSC syntax and the CwF syntax are isomorphic for a
+theory with dependent function space and a hierarchy of universes. SSC
+only includes single substitutions and single weakenings, and eight
+equations relating these: four equations describe how to substitute
+variables, and there are four equations on types which are needed to
+typecheck other equations. All the other equations are substitution
+(naturality) rules or computation rules for different type
+formers. SSC provides a simple, minimalistic alternative to parallel
+substitution calculi or B-systems for defining type theory. SSC
+relates to CwF as extensional combinatory calculus relates to lambda
+calculus. All the results in this paper were formalised in Agda.
 \end{abstract}
+
+\maketitle
 
@@
 
-\cite{singleTypes}
-\cite{coquandEhrhard}
-
-What is type theory? This term can refer to formal systems (languages,
-type systems, type theories) or to the field of study of type
-theories. In this paper, we will give an answer to this question in
-the form of a concrete formal system. We aim for simplicity and ease
-of understanding: we only require a working knowledge of Agda, a
-functional programming language which is an implementation of type
-theory. Our metatheory in which this paper is written is thus type
-theory, and everything is formalised in Agda.
+What is type theory? Here we refer to type theory as a particular
+formal system based on Martin-Löf's original definition
+\cite{martinlof73predicative}, and not to the field studying type
+systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When
+answering this question, we aim for simplicity and ease of
+understanding. We only require a working knowledge of an
+implementation of type theory such as Agda. Our metatheory in which
+this paper is written is thus type theory, and everything is
+formalised in Agda.
 
 The traditional way of defining a programming language formally
-\cite{DBLP:books/daglib/0005958,DBLP:books/cu/Ha2016,Pierce:SF1,plfa22.08}
-is comprised of three steps: (i) definition of the presyntax
-(abstract syntax trees, ASTs) as a formal grammar; (ii) specification of
+\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
+steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
+as a formal grammar; (ii) specification of
 well-formedness/well-typedness relations saying which ASTs are
@@
 grammar; (ii) terms are indexed by (object theoretic) types, thus only
-allowing ASTs to be plugged together in well-typed ways; (iii) terms
-are quotiented by conversion rules, making convertible terms
-equal. Inductive types which allow such mutual indexing and
-quotienting are called quotient inductive-inductive types (QIITs
-\cite{DBLP:journals/pacmpl/KaposiKA19}). As Agda does not support them
-directly, we postulate their constructors and induction principle and add the
-computation rules using rewrite rules \cite{DBLP:conf/types/Cockx19}.
+allowing ASTs to be built in well-typed ways; (iii) terms are
+quotiented by conversion rules, making convertible terms equal. The
+advantage of the intrinsic approach that it provides shorter, easier
+to understand, less technical definitions, and there is no separation
+between syntax and semantics. Type theory is a generalised algebraic
+theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the notion of model
+is direct, and the syntax is simply the initial model. GATs are
+generalisations of single-sorted algebraic theories (such as monoids,
+groups) to multiple sorts where later sorts can be indexed over
+previous sorts. A typical example is categories with a sorts of
+objects and a sort of morphisms double-indexed over objects. The
+initial model of a GAT is a quotient inductive-inductive type (QIIT
+\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
+inductive types where multiple types are defined mutually, later types
+can be indexed over previous ones, and equality constructors are also
+allowed, generalising quotients. The elimination principle of the QIIT
+corresponds to a structural induction principle for the QIIT and is
+equivalent to the statement that the syntax is the initial model
+\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
+morphism from the syntax to any other model.
 
-A QIIT is also the initial algebra/model of a generalised algebraic
-theory (GAT \cite{DBLP:journals/apal/Cartmell86}). The induction principle of
-the QIIT is equivalent to having a unique morphism into any model
-\cite{DBLP:journals/pacmpl/KaposiKA19}. There are several ways to
-define type theory as a GAT: categories with families (CwFs
-\cite{DBLP:journals/corr/abs-1904-00827}), contextual categories
-\cite{DBLP:journals/apal/Cartmell86}, natural models
-\cite{DBLP:journals/mscs/Awodey18}, C-systems and B-systems
-\cite{ahrens_emmenegger_north_rijke_2023}. When extracting a QIIT from
-a GAT, with the exception of B-systems, we obtain parallel
-substitution calculi. Like B-systems, in this paper we define type
-theory using \emph{single substitutions}. Our definition is simpler
-than B-systems, but it is closely related to it. It is a minimalistic
-presentation where every operation and equation is well-motivated (see
-Section \ref{sec:SSC}). However, when we want to use our syntax (for
-example, if we want to prove canonicity or normalisation for our
-theory), it is much more convenient to use parallel substitution
-calculi such as CwFs. We first show how to get rid of explicit
-substitutions from the syntax using the technique of
-$\alpha$-normalisation (Section \ref{sec:alpha}). Then we show how to
-derive the rules of a CwF from our syntax (Section
-\ref{sec:cwf}). Finally, we show that our syntax is isomorphic to the
-CwF-style syntax (Section \ref{sec:iso}). We list related work and
-conclude in Section \ref{sec:conclusion}.
+There are several ways to define type theory as a GAT. For example,
+Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
+categories with families (CwFs
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
+or C-systems
+\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
+natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
+basic substitution calculus part of the theory, and individual type
+formers (such as function space or product types) have to be specified
+separately. With the exception of B-systems, the listed definitions
+are all based on parallel substitution calculi, where contexts and
+substitutions form a category. Like B-systems, in this paper we define
+type theory using \emph{single substitutions}. Single substitutions
+don't compose, so they do not form a category. Our definition is
+simpler than B-systems, but it is closely related to it. It is a
+minimalistic presentation where every operation and equation is
+well-motivated (see Section \ref{sec:ssc}). In addition to its
+pedagogical value due to its simplicity, our definition might
+contribute in the future to a possible coherent syntax for type
+theory, see Section \ref{sec:conclusion}.
+
+Our theory is in fact a bit too minimalistic to be workable: for
+example, consider a term $b$ which depends on a context $\Gamma$ and
+an extra variable of type $A$, that is, $b : \Tm\,(\Gamma\ext
+A)\,B$. We can weaken $b$ so that we add an extra $A$ variable
+\emph{before} the last one obtaining $b[\p^+] : Tm\,(\Gamma\ext A\ext
+A[\p])\,(B[\p])$ where $\p$ is the weakening substitution. Then, we
+instantiate the last $A[\p]$ variable in the context by the previous
+$A$ obtaining $b[\p^+][\langle\q\rangle]$ where $\q$ is De Bruijn
+index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an
+arbitrary model, but this equation is admissible in the \emph{syntax}.
+% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
+% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
+This is analogous to e.g.\ parametricity results
+\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
+arbitrary model, but they hold for syntactic terms. In fact, all the
+operations and equations of the parallel substitution calculus (CwF)
+are admissible in our syntax, moreover the set of single substitution
+based syntactic terms are isomorphic to the set of CwF-based
+terms. Every CwF-based model is also a model of our single
+substitution calculus, but not the other way: there are more
+SSC-models than CwF-models. The situation is analogous to that of
+extensional combinatory calculus and lambda calculus
+\cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic terms form
+isomorphic sets, but there are more combinatory models than lambda
+calculus models.
+
+TODO: rewrite from here. Add: discussion of type formers, SOGATs and say that we only do this
+for an example t.t. and not for all of them. We conjecture that for
+any SOGAT, the SSC and CwF-based syntaxes are equivalent.
+
+The summary of the above discussion: we define a new minimal syntax
+for type theory which is equivalent to the original syntax, but
+
+We first show how to get rid of explicit substitutions from the syntax
+using the technique of $\alpha$-normalisation (Section
+\ref{sec:alpha}). Then we show how to derive the rules of a CwF from
+our syntax (Section \ref{sec:cwf}). Finally, we show that our syntax
+is isomorphic to the CwF-style syntax (Section \ref{sec:iso}). We list
+related work and conclude in Section \ref{sec:conclusion}.
 
@@
 
+% TODO: say that this is the paper version of the \cite{singleTypes}
+% paper
+
 \subsection{Formalisation}
+
+As Agda does not support them directly, we postulate their
+constructors and induction principle and add the computation rules
+using rewrite rules \cite{DBLP:conf/types/Cockx19}.
 
@@
 
-\section{Related work and conclusion}
+\section{Conclusions}
+\label{sec:conclusion}
 

### 2024-09-07 types_2024_postproc/p.tex
stuff

@@
 
+\newcommand{\ra}{\rightarrow}
+\newcommand{\Set}{\mathsf{Set}}
 \newcommand{\Ty}{\mathsf{Ty}}
@@
 \newcommand{\ext}{\mathop{\triangleright}}
+\newcommand{\N}{\mathbb{N}}
+\newcommand{\lam}{\mathsf{lam}}
+\newcommand{\app}{\mathsf{app}}
+\newcommand{\U}{\mathsf{U}}
+\newcommand{\El}{\mathsf{El}}
+\newcommand{\cd}{\mathsf{c}}
 
@@
 \cite{martinlof73predicative}, and not to the field studying type
-systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When
-answering this question, we aim for simplicity and ease of
-understanding. We only require a working knowledge of an
-implementation of type theory such as Agda. Our metatheory in which
-this paper is written is thus type theory, and everything is
-formalised in Agda.
+systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When answering this
+question, we aim for simplicity and ease of understanding.
 
@@
 quotiented by conversion rules, making convertible terms equal. The
-advantage of the intrinsic approach that it provides shorter, easier
-to understand, less technical definitions, and there is no separation
-between syntax and semantics. Type theory is a generalised algebraic
-theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the notion of model
-is direct, and the syntax is simply the initial model. GATs are
-generalisations of single-sorted algebraic theories (such as monoids,
-groups) to multiple sorts where later sorts can be indexed over
-previous sorts. A typical example is categories with a sorts of
-objects and a sort of morphisms double-indexed over objects. The
-initial model of a GAT is a quotient inductive-inductive type (QIIT
+advantage of the intrinsic approach is that it provides shorter,
+easier to understand, less technical definitions, and there is no
+mismatch between syntax and semantics. Type theory is a generalised
+algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the
+notion of model is direct, and the syntax is simply the initial
+model. GATs are generalisations of single-sorted algebraic theories
+(such as monoids, groups) to multiple sorts where later sorts can be
+indexed over previous sorts. An example is categories with a sort of
+objects and a sort of morphisms double-indexed over objects. In a
+type-theoretic metatheory, the initial model of a GAT is a quotient
+inductive-inductive type (QIIT
 \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
@@
 allowed, generalising quotients. The elimination principle of the QIIT
-corresponds to a structural induction principle for the QIIT and is
+corresponds to a structural induction principle and is
 equivalent to the statement that the syntax is the initial model
@@
 separately. With the exception of B-systems, the listed definitions
-are all based on parallel substitution calculi, where contexts and
+all define a parallel substitution calculus, where contexts and
 substitutions form a category. Like B-systems, in this paper we define
-type theory using \emph{single substitutions}. Single substitutions
-don't compose, so they do not form a category. Our definition is
-simpler than B-systems, but it is closely related to it. It is a
-minimalistic presentation where every operation and equation is
-well-motivated (see Section \ref{sec:ssc}). In addition to its
-pedagogical value due to its simplicity, our definition might
+type theory as a single substitution calculus (SSC). Single
+substitutions don't compose, so they do not form a category. Our
+definition is simpler than B-systems, but it is closely related to
+it. It is a minimalistic presentation where every operation and
+equation is well-motivated (see Section \ref{sec:ssc}). In addition to
+its pedagogical value due to its simplicity, our definition might
 contribute in the future to a possible coherent syntax for type
@@
 
-Our theory is in fact a bit too minimalistic to be workable: for
-example, consider a term $b$ which depends on a context $\Gamma$ and
-an extra variable of type $A$, that is, $b : \Tm\,(\Gamma\ext
-A)\,B$. We can weaken $b$ so that we add an extra $A$ variable
-\emph{before} the last one obtaining $b[\p^+] : Tm\,(\Gamma\ext A\ext
-A[\p])\,(B[\p])$ where $\p$ is the weakening substitution. Then, we
-instantiate the last $A[\p]$ variable in the context by the previous
-$A$ obtaining $b[\p^+][\langle\q\rangle]$ where $\q$ is De Bruijn
-index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an
-arbitrary model, but this equation is admissible in the \emph{syntax}.
+Our theory is in some sense too minimalistic: for example, consider a
+term $b$ which depends on a context $\Gamma$ and an extra variable of
+type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We can weaken $b$ so
+that we add an extra $A$ variable \emph{before} the last one obtaining
+$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])$ where $\p$ is
+weakening at the end of the context, and $\p^+$ is weakening just
+before the last variable in the context. Then, we instantiate the last
+$A[\p]$ variable in the context by the previous $A$ obtaining
+$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ is De
+Bruijn index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
+an arbitrary model, but this equation holds in the syntax.
 % p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
@@
 operations and equations of the parallel substitution calculus (CwF)
-are admissible in our syntax, moreover the set of single substitution
-based syntactic terms are isomorphic to the set of CwF-based
-terms. Every CwF-based model is also a model of our single
-substitution calculus, but not the other way: there are more
-SSC-models than CwF-models. The situation is analogous to that of
-extensional combinatory calculus and lambda calculus
-\cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic terms form
-isomorphic sets, but there are more combinatory models than lambda
-calculus models.
+are admissible in our syntax, moreover the set of SSC-based syntactic
+terms are isomorphic to the set of CwF-based terms. In contrast with
+B-systems which are equivalent to CwFs, our SSC-models do not form
+CwF-based models. Every CwF-based model is a model of our SSC, but not
+the other way: there are more SSC-models than CwF-models. The
+situation is analogous to the relationship of extensional combinatory
+calculus and lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}:
+the syntactic terms form isomorphic sets, but there are more
+combinatory models than lambda calculus models.
 
-TODO: rewrite from here. Add: discussion of type formers, SOGATs and say that we only do this
-for an example t.t. and not for all of them. We conjecture that for
-any SOGAT, the SSC and CwF-based syntaxes are equivalent.
+In this paper, we define the SSC and show how it can be extended with
+$\Pi$ types and a hierarchy of Coquand-universes
+\cite{coquandUniverse}. The rules for type formers are the same as for
+CwFs, the difference is only in the substitution calculus part. More
+generally, every second-order generalised algebraic theory (SOGAT) has
+a parallel and a single substition-based first order presentation
+\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
+the parallel and the single substitution calculi are isomorphic for
+the following SOGAT (we use implicit arguments and $\cong$ denotes
+isomorphism):
+\begin{alignat*}{10}
+  & \Ty && : \N\ra\Set \\
+  & \Tm && : \Ty\,i\ra\Set \\
+  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
+  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
+  & \U && : (i:\N)\ra\Ty\,(1+i) \\
+  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
+\end{alignat*}
+We conjecture that the methods of in this paper can be used to show
+that for any SOGAT, the two first-order presentations give isomorphic
+syntaxes.
 
-The summary of the above discussion: we define a new minimal syntax
-for type theory which is equivalent to the original syntax, but
+TODO: say something about alpha-normalisation.
+
+To summarise, our contributions are the following:
+\begin{itemize}
+\item A new generalised algebraic presentation of type theory in the
+  form of a minimalistic single substitution calculus.
+\item The $\alpha$-normalisation technique which is useful for working
+  with explicit substitution calculi.
+\item The proof of equivalence with the parallel, CwF-based syntax.
+\end{itemize}
+
+\subsection{Structure of the paper}
+
+In Section \ref{sec:ssc}, we introduce
 
@@
 
-Our main contribution is the novel definition of a minimalistic single
-substitution calculus, the $\alpha$-normalisation algorithm which can
-be useful for other explicit substitution calculi, and finally our
-proof of equivalence with the CwF-style syntax.
+We only require a working knowledge of an implementation of type
+theory such as Agda. Our metatheory in which this paper is written is
+thus type theory, and everything is formalised in Agda.
+
+\subsection{Related work}
+
+Algebraic formalisations of type theory: Brunerie
+\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
+normalisation, shallow embedding canonicity, gluing. Low level
+formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
+more abstract thing in a (weak) setoid model.
+
+Other SSC: 90s French, Thorsten thesis, B-systems, Luksa paper.
 
@@
 
-\subsection{Structure of the paper}
-
 \section{Single substitution syntax}
@@
 
-\section{The single substitution syntax is a category with families}
+Show examples of using the calculus, and where we need to derive more
+stuff.
+
+\section{Deriving parallel substituions}
 
@@
 from the single substitution syntax defined in \autoref{sec:ssc}.
+
+\section{Isomorphism}
 
@@
 models, but the syntaxes are isomorphic.
-
-SSC was used in the 90s for simple type theories, ask Hugo for references.
 
@@
 
-Maybe we can do without the extra type equations, by using assumptions.
-
-Luksa, Thorsten thesis.
+Separate sort for variables, without the extra type equations, by
+using assumptions, and proving everything by induction.
 

### 2024-09-09 types_2024_postproc/p.tex
stuff

@@
 \newcommand{\Tm}{\mathsf{Tm}}
+\newcommand{\Con}{\mathsf{Con}}
+\newcommand{\Sub}{\mathsf{Sub}}
 \newcommand{\p}{\mathsf{p}}
@@
 \newcommand{\cd}{\mathsf{c}}
+\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
 
@@
 
-In this paper, we define the SSC and show how it can be extended with
+We define the parallel substitution calculus operations by induction
+on the SSC-syntax. One of our constructors is the instantiation
+operation of single substitutions: $\blank[\blank] :
+\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
+terms. When doing induction on terms, we always have to provide a
+method for these constructors, but this can be difficult as
+substitutions don't compose. This is why we derive a new induction
+principle which works on $\alpha$-normal forms. Just as $\beta$-normal
+forms don't distinguish between $\beta$-equal terms, $\alpha$-normal
+forms don't include explicit substitutions, all the variables are
+instantiated. $\alpha$-normal forms are not usual normal forms in the
+sense that they are still quotiented by the computation/uniqueness
+rules like $\beta$/$\eta$ for function space.
+
+In this paper, we define the SSC and extend it with
 $\Pi$ types and a hierarchy of Coquand-universes
@@
 
-TODO: say something about alpha-normalisation.
-
 To summarise, our contributions are the following:
@@
 \item A new generalised algebraic presentation of type theory in the
-  form of a minimalistic single substitution calculus.
+  form of a minimalistic single substitution calculus (SSC).
 \item The $\alpha$-normalisation technique which is useful for working
   with explicit substitution calculi.
-\item The proof of equivalence with the parallel, CwF-based syntax.
+\item The proof of equivalence of the SSC-syntax with the CwF-based
+  syntax.
 \end{itemize}
@@
 
-In Section \ref{sec:ssc}, we introduce
-
-We first show how to get rid of explicit substitutions from the syntax
-using the technique of $\alpha$-normalisation (Section
-\ref{sec:alpha}). Then we show how to derive the rules of a CwF from
-our syntax (Section \ref{sec:cwf}). Finally, we show that our syntax
-is isomorphic to the CwF-style syntax (Section \ref{sec:iso}). We list
-related work and conclude in Section \ref{sec:conclusion}.
-
-We only require a working knowledge of an implementation of type
-theory such as Agda. Our metatheory in which this paper is written is
-thus type theory, and everything is formalised in Agda.
+In Section \ref{sec:ssc}, we introduce the single substitution
+calculus only relying on a working knowledge of an implementation of
+type theory such as Agda, no prior experience with the metatheory of
+type theory is required. We first show how to get rid of explicit
+substitutions from the syntax using the technique of
+$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
+prove several admissible equations. Then we show how to derive the
+rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
+show that our syntax is isomorphic to the CwF-style syntax (Section
+\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.
 
@@
 
+Most formalisation of type theory use low-level
+
 Algebraic formalisations of type theory: Brunerie
+
 \cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
@@
 \subsection{Formalisation}
+
+Our metatheory in which this paper is
+written is thus type theory, and everything is formalised in Agda.
 
@@
 
-\section{Deriving parallel substituions}
+\section{$\alpha$-normalisation}
+\label{sec:alpha}
+
+$\alpha$-normal forms could also be called substitution normal forms.
+
+\section{Isomorphism}
+\label{sec:cwf}
 
@@
 
+TODO: show that the SSC operations are derivable from CwF.
+
 \section{Isomorphism}
+\label{sec:iso}
 

### 2024-09-13 types_2024_postproc/p.tex
stuff

@@
 
+Type theory is a language which can be described as a second-order
+generalised algebraic theory (SOGAT
+\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
+\begin{enumerate}
+\item[(i)] It is an intrinsic presentation
+  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
+  opposed to extrinsic. That is, we only consider well-formed,
+  well-scoped and well-typed abstract syntax trees, there are no
+  meaningless terms.
+\item[(ii)] Terms are quotiented by the conversion relation.
+\item[(iii)] Every operation is stable under substitution.
+\end{enumerate}
+For example, type theory with $\Pi$ types and universes {\`a} la
+Coquand \cite{coquandUniverse} is specified by the following SOGAT.
+\begin{alignat*}{10}
+  & \Ty && : \N\ra\Set \\
+  & \Tm && : \Ty\,i\ra\Set \\
+  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
+  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
+  & \U && : (i:\N)\ra\Ty\,(1+i) \\
+  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
+\end{alignat*}
+We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
+and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
+denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
+$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
+and $\lam$ are second-order functions. We say that $\Pi$ binds a
+$\Tm$-variable in its second $\Ty$-argument. Every binder is
+represented by a second-order function, and as a result, we don't need
+to talk about contexts and substitutions when describing the theory:
+we re-use the binding features of the metatheory to describe those of
+our object theory (this technique is also called higher-order abstract
+syntax \cite{DBLP:conf/lics/Hofmann99} or logical framework
+\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
+is no good notion of homomorphism between second-order models, thus
+the semantics of a SOGAT is given by a first order generalised
+algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
+generalisations of single-sorted algebraic theories (such as monoids,
+groups) to multiple sorts where later sorts can be indexed over
+previous sorts, and the usual results from universal algebra (e.g.\ a
+category of models with finite limits and initial object, free and
+co-free models) carry over to GATs
+\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
+theory of categories where there is a sort for objects and a sort for
+morphisms double-indexed over objects.
+
+Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} present two different
+first order semantics for any SOGAT: one based on parallel
+substitutions, and one based on single substitutions.
+
+In the parallel semantics, the GAT always starts with a category with
+a terminal object, objects of the category are called contexts,
+morphisms are substitutions. Every sort and operation of the SOGAT
+gives rise to an operation in the GAT which is indexed by
+contexts. The sorts moreoever come with functorial instantiation
+(sometimes called substitution) operations, and all the operations are
+natural with respect to instantiation. For those sorts which are bound
+by an operation, we have context extension, meaning that contexts can
+carry variables of that sort (in other words, the sort is given by a
+locally representable presheaf). The parallel semantics of the GAT
+\begin{alignat*}{10}
+  & \Ty && : \Set \\
+  & \Tm && : \Ty \ra\Set 
+\end{alignat*}
+is the GAT of categories with families (CwFs
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}). To define the notion
+of model of a language (given by a SOGAT) is to take its parallel
+substitution semantics, which is a GAT, and that induces a notion of
+model. The syntax of the language is the initial model, which exists
+for every GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an
+intrinsic definition of syntax (only well-typed terms), which is
+quotiented by the equations in the SOGAT, which describe conversion
+for the language.
+
+The single substitution semantics is an alternative GAT xxx
+
+In a type-theoretic metatheory,
+the initial model of a GAT is a quotient inductive-inductive type
+(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
+family of inductive types where multiple types are defined mutually,
+later types can be indexed over previous ones, and equality
+constructors are also allowed, generalising quotients. The elimination
+principle of the QIIT corresponds to a structural induction principle
+and is equivalent to the statement that the syntax is the initial
+model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
+unique morphism from the syntax to any other model.
+
+conversion to GAT, algebra, models
+
 The traditional way of defining a programming language formally
@@
 we define ASTs as an inductive type instead of production rules of a
-grammar; (ii) terms are indexed by (object theoretic) types, thus only
-allowing ASTs to be built in well-typed ways; (iii) terms are
-quotiented by conversion rules, making convertible terms equal. The
-advantage of the intrinsic approach is that it provides shorter,
-easier to understand, less technical definitions, and there is no
-mismatch between syntax and semantics. Type theory is a generalised
-algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the
-notion of model is direct, and the syntax is simply the initial
-model. GATs are generalisations of single-sorted algebraic theories
-(such as monoids, groups) to multiple sorts where later sorts can be
-indexed over previous sorts. An example is categories with a sort of
-objects and a sort of morphisms double-indexed over objects. In a
-type-theoretic metatheory, the initial model of a GAT is a quotient
-inductive-inductive type (QIIT
-\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
-inductive types where multiple types are defined mutually, later types
-can be indexed over previous ones, and equality constructors are also
-allowed, generalising quotients. The elimination principle of the QIIT
-corresponds to a structural induction principle and is
-equivalent to the statement that the syntax is the initial model
-\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
-morphism from the syntax to any other model.
+grammar; (ii) types are part of the definition and terms are indexed
+by them, thus only allowing ASTs to be built in well-typed ways; (iii)
+terms are quotiented by conversion rules, making convertible terms
+equal. Compared to the traditional extrinsic approach, the intrinsic approach provides
+shorter, less technical definitions, and there
+is no mismatch between syntax and semantics. Type theory is a
+generalised algebraic theory (GAT
+\cite{DBLP:journals/apal/Cartmell86}), the notion of model is direct,
+and the syntax is simply the initial model.
 
@@
 natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
-basic substitution calculus part of the theory, and individual type
-formers (such as function space or product types) have to be specified
-separately. With the exception of B-systems, the listed definitions
-all define a parallel substitution calculus, where contexts and
-substitutions form a category. Like B-systems, in this paper we define
-type theory as a single substitution calculus (SSC). Single
-substitutions don't compose, so they do not form a category. Our
-definition is simpler than B-systems, but it is closely related to
-it. It is a minimalistic presentation where every operation and
-equation is well-motivated (see Section \ref{sec:ssc}). In addition to
-its pedagogical value due to its simplicity, our definition might
-contribute in the future to a possible coherent syntax for type
-theory, see Section \ref{sec:conclusion}.
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
+B-systems, the listed notions all define a parallel substitution
+calculus, where contexts and substitutions form a category. Like
+B-systems, in this paper we define a single substitution calculus
+(SSC). Single substitutions don't compose, so they do not form a
+category. Our definition is simpler than B-systems, but it is closely
+related to it. It is a minimalistic presentation where every operation
+and equation is well-motivated (see Section \ref{sec:ssc}). In
+addition to its pedagogical value due to its simplicity, in the future
+it might give rise to new models or contribute to a possible coherent
+syntax for type theory, see Section \ref{sec:conclusion}.
 
-Our theory is in some sense too minimalistic: for example, consider a
+CwFs, C-systems, B-systems and our SSC only provide the basic
+substitution calculus part of the theory, and individual type formers
+(such as function space or product types) have to be specified
+separately. There is a systematic way to do this: non-substructural
+programming languages can be specified as second-order generalised
+algebraic theories (SOGATs \cite{DBLP:conf/fscd/KaposiX24}), which are
+GATs allowing second-order operations. For example, type theory with
+$\Pi$ and a family of
+
+Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} showed that any SOGAT
+has two kinds of GAT semantics, one based on a parallel substitution
+calculus, and another based on a single substitution calculus. The
+parallel substitution GAT semantics of the SOGAT containing only the
+sorts of types and terms ($\Ty : \Set, \Tm : \Ty\ra\Set$) is simply
+the GAT of CwFs. The single substitution semantics of the same SOGAT
+is the SSC in this paper.
+
+SOGAT stuff
+
+Our SSC is in some sense too minimalistic: for example, consider a
 term $b$ which depends on a context $\Gamma$ and an extra variable of
-type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We can weaken $b$ so
+type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so
 that we add an extra $A$ variable \emph{before} the last one obtaining
-$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])$ where $\p$ is
+$
+b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])
+$
+where $\p$ is
 weakening at the end of the context, and $\p^+$ is weakening just
@@
 $A[\p]$ variable in the context by the previous $A$ obtaining
-$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ is De
-Bruijn index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
+$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ denotes the last
+variable in the context (a.k.a\ De Bruijn index $0$).
+We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
 an arbitrary model, but this equation holds in the syntax.
@@
 \cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
-arbitrary model, but they hold for syntactic terms. In fact, all the
+arbitrary model, but they hold in the syntax. In fact, we can show
+that the syntax xxx
+
+In fact, all the
 operations and equations of the parallel substitution calculus (CwF)
-are admissible in our syntax, moreover the set of SSC-based syntactic
-terms are isomorphic to the set of CwF-based terms. In contrast with
-B-systems which are equivalent to CwFs, our SSC-models do not form
-CwF-based models. Every CwF-based model is a model of our SSC, but not
-the other way: there are more SSC-models than CwF-models. The
-situation is analogous to the relationship of extensional combinatory
-calculus and lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}:
-the syntactic terms form isomorphic sets, but there are more
-combinatory models than lambda calculus models.
+are admissible in the syntax of SSC, moreover the set of syntactic SSC
+terms are isomorphic to the set of CwF terms. In contrast with
+B-systems which are equivalent to CwFs, our SSCs do not form
+CwFs. Every CwF-based model is a model of our SSC, but not the other
+way: there are more SSC-models than CwF-models. The situation is
+analogous to the relationship of extensional combinatory calculus and
+lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic
+terms form isomorphic sets, but there are more combinatory models than
+lambda calculus models.
 
@@
 the parallel and the single substitution calculi are isomorphic for
-the following SOGAT (we use implicit arguments and $\cong$ denotes
-isomorphism):
-\begin{alignat*}{10}
-  & \Ty && : \N\ra\Set \\
-  & \Tm && : \Ty\,i\ra\Set \\
-  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
-  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
-  & \U && : (i:\N)\ra\Ty\,(1+i) \\
-  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
-\end{alignat*}
+the following SOGAT
+
 We conjecture that the methods of in this paper can be used to show
@@
 
-Most formalisation of type theory use low-level
+In the 1990s single substitution calculi for type theory were popular,
+but usually in the extrinsic setting, see e.g.\ the thesis of
+Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
+\cite{voevodsky2014bsystems} as an algebraic way of describing type
+theories close to the notations using typing judgements. Using our
+notations B-systems can be described as follows. B-frame.
+
+Ehrhard
+
+we use Sub, but we could have used telescopes
+
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}
+
+contextual sCwF and contextual SSC with fct types
+
+Formalisations of type theory use low-level
 
 Algebraic formalisations of type theory: Brunerie
+
+Initiality conjecture.
 
@@
 
+\section{Minimisation}
+
+\section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
+
 \section{Conclusions}
 \label{sec:conclusion}
+
+Why are we still looking at alternative ways of defining type theory?
+Closer to the syntax/implementation, easier to understand/implement,
+fewer equations, hence easier to define models, coherent syntax?
 

### 2024-09-13 types_2024_postproc/p.tex
stuff

@@
 \cite{martinlof73predicative}, and not to the field studying type
-systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When answering this
-question, we aim for simplicity and ease of understanding.
+systems (e.g.\ \cite{DBLP:books/daglib/0005958}).
 
@@
   \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
-  opposed to extrinsic. That is, we only consider well-formed,
-  well-scoped and well-typed abstract syntax trees, there are no
-  meaningless terms.
+  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
+  only consider well-formed, well-scoped and well-typed abstract
+  syntax trees, there are no meaningless terms.
 \item[(ii)] Terms are quotiented by the conversion relation.
@@
 Coquand \cite{coquandUniverse} is specified by the following SOGAT.
-\begin{alignat*}{10}
+\begin{equation}\label{eq:tt}
+\begin{alignedat}{10}
   & \Ty && : \N\ra\Set \\
@@
   & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
-\end{alignat*}
+\end{alignedat}
+\end{equation}
 We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
@@
 and $\lam$ are second-order functions. We say that $\Pi$ binds a
-$\Tm$-variable in its second $\Ty$-argument. Every binder is
-represented by a second-order function, and as a result, we don't need
-to talk about contexts and substitutions when describing the theory:
-we re-use the binding features of the metatheory to describe those of
-our object theory (this technique is also called higher-order abstract
-syntax \cite{DBLP:conf/lics/Hofmann99} or logical framework
+$\Tm$-variable of type $A$ in its second $\Ty$-argument. Every binder
+is represented by a second-order function, and as a result, we don't
+need to talk about contexts and substitutions when describing the
+theory: we re-use the binding features of the metatheory to describe
+those of our object theory (this technique is also known as
+higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99} or
+logical framework
 \cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
@@
 groups) to multiple sorts where later sorts can be indexed over
-previous sorts, and the usual results from universal algebra (e.g.\ a
-category of models with finite limits and initial object, free and
-co-free models) carry over to GATs
+previous sorts. The usual results from universal algebra (e.g.\ models
+form a category with finite limits and initial object, existence of
+free and co-free models) carry over to GATs
 \cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
@@
 a terminal object, objects of the category are called contexts,
-morphisms are substitutions. Every sort and operation of the SOGAT
-gives rise to an operation in the GAT which is indexed by
-contexts. The sorts moreoever come with functorial instantiation
-(sometimes called substitution) operations, and all the operations are
-natural with respect to instantiation. For those sorts which are bound
-by an operation, we have context extension, meaning that contexts can
-carry variables of that sort (in other words, the sort is given by a
-locally representable presheaf). The parallel semantics of the GAT
-\begin{alignat*}{10}
+morphisms are substitutions. Every sort/operation of the SOGAT gives
+rise to a sort/operation in the GAT which is now indexed by
+contexts. The sorts moreoever come with instantiation (sometimes
+called substitution) operations which are functorial, and all the
+operations are natural with respect to instantiation. For those sorts
+which are bound by an operation in the SOGAT description, we have
+context extension, meaning that contexts can carry variables of that
+sort (in other words, the sort is given by a locally representable
+presheaf). The parallel semantics of the SOGAT
+\begin{equation}\label{eq:tytm}
+\begin{alignedat}{10}
   & \Ty && : \Set \\
   & \Tm && : \Ty \ra\Set 
-\end{alignat*}
+\end{alignedat}
+\end{equation}
 is the GAT of categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}). To define the notion
-of model of a language (given by a SOGAT) is to take its parallel
-substitution semantics, which is a GAT, and that induces a notion of
-model. The syntax of the language is the initial model, which exists
-for every GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an
-intrinsic definition of syntax (only well-typed terms), which is
-quotiented by the equations in the SOGAT, which describe conversion
-for the language.
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}). We define the notion
+of model of a language (a SOGAT) by its parallel substitution
+semantics, which is a GAT, and that induces a notion of model. The
+syntax of the language is the initial model, which exists for every
+GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
+syntax (only well-typed terms), quotiented by the equations in the
+SOGAT, which describe conversion for the language.
 
-The single substitution semantics is an alternative GAT xxx
+The single substitution semantics is a GAT which does not start with a
+category: there are contexts and maps between contexts, but these maps
+are single substitutions or single weakenings, they don't
+compose. There are fewer operations and fewer equations than in the
+parallel semantics. In this paper we investigate the single
+substitution semantics of the type theory specified by the SOGAT
+(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
+the operations are all easy to motivate, we illustrate this in Section
+\ref{sec:ssc}. Single substitution calculi are closer to the usual
+presentation of type theory using derivation rules, this was one of
+the motivations of Voevodsky for developing B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
+semantics of (\ref{eq:tytm}) is simpler and has more models than
+B-systems. GAT presentations of type theory are important for defining
+models, our presentation gives a new notion of model, which in the
+future might contribute to definitions of new models. Because our
+single substitution GAT does not involve a category, it might be
+easier to adapt it in a higher dimensional setting, see Section
+\ref{sec:conclusion}.
 
-In a type-theoretic metatheory,
-the initial model of a GAT is a quotient inductive-inductive type
-(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
-family of inductive types where multiple types are defined mutually,
-later types can be indexed over previous ones, and equality
-constructors are also allowed, generalising quotients. The elimination
-principle of the QIIT corresponds to a structural induction principle
-and is equivalent to the statement that the syntax is the initial
-model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
-unique morphism from the syntax to any other model.
-
-conversion to GAT, algebra, models
-
-The traditional way of defining a programming language formally
-\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
-steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
-as a formal grammar; (ii) specification of
-well-formedness/well-typedness relations saying which ASTs are
-well-behaved; (iii) specification of rewriting/conversion relations
-explaining how programs are executed. Instead of following this route,
-we use the intrinsic approach
-\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
-we define ASTs as an inductive type instead of production rules of a
-grammar; (ii) types are part of the definition and terms are indexed
-by them, thus only allowing ASTs to be built in well-typed ways; (iii)
-terms are quotiented by conversion rules, making convertible terms
-equal. Compared to the traditional extrinsic approach, the intrinsic approach provides
-shorter, less technical definitions, and there
-is no mismatch between syntax and semantics. Type theory is a
-generalised algebraic theory (GAT
-\cite{DBLP:journals/apal/Cartmell86}), the notion of model is direct,
-and the syntax is simply the initial model.
-
-There are several ways to define type theory as a GAT. For example,
-Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
-categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
-or C-systems
-\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
-natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
-B-systems, the listed notions all define a parallel substitution
-calculus, where contexts and substitutions form a category. Like
-B-systems, in this paper we define a single substitution calculus
-(SSC). Single substitutions don't compose, so they do not form a
-category. Our definition is simpler than B-systems, but it is closely
-related to it. It is a minimalistic presentation where every operation
-and equation is well-motivated (see Section \ref{sec:ssc}). In
-addition to its pedagogical value due to its simplicity, in the future
-it might give rise to new models or contribute to a possible coherent
-syntax for type theory, see Section \ref{sec:conclusion}.
-
-CwFs, C-systems, B-systems and our SSC only provide the basic
-substitution calculus part of the theory, and individual type formers
-(such as function space or product types) have to be specified
-separately. There is a systematic way to do this: non-substructural
-programming languages can be specified as second-order generalised
-algebraic theories (SOGATs \cite{DBLP:conf/fscd/KaposiX24}), which are
-GATs allowing second-order operations. For example, type theory with
-$\Pi$ and a family of
-
-Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} showed that any SOGAT
-has two kinds of GAT semantics, one based on a parallel substitution
-calculus, and another based on a single substitution calculus. The
-parallel substitution GAT semantics of the SOGAT containing only the
-sorts of types and terms ($\Ty : \Set, \Tm : \Ty\ra\Set$) is simply
-the GAT of CwFs. The single substitution semantics of the same SOGAT
-is the SSC in this paper.
-
-SOGAT stuff
-
-Our SSC is in some sense too minimalistic: for example, consider a
-term $b$ which depends on a context $\Gamma$ and an extra variable of
-type $A$, that is, $b : \Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so
-that we add an extra $A$ variable \emph{before} the last one obtaining
-$
-b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p])
-$
-where $\p$ is
-weakening at the end of the context, and $\p^+$ is weakening just
-before the last variable in the context. Then, we instantiate the last
-$A[\p]$ variable in the context by the previous $A$ obtaining
-$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$ denotes the last
-variable in the context (a.k.a\ De Bruijn index $0$).
-We cannot derive $b[\p^+][\langle\q\rangle] = b$ in
-an arbitrary model, but this equation holds in the syntax.
+The single substitution semantics of (\ref{eq:tytm}) is in some sense
+too minimalistic: for example, consider a term $b$ which depends on a
+context $\Gamma$ and an extra variable of type $A$, that is, $b :
+\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
+variable \emph{before} the last one obtaining $ b[\p^+] :
+\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
+the end of the context, and $\p^+$ is weakening just before the last
+variable in the context. Then, we instantiate the last $A[\p]$
+variable in the context by the previous $A$ obtaining
+$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
+denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
+We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary
+model, but this equation holds in the syntax.
 % p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
@@
 \cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
-arbitrary model, but they hold in the syntax. In fact, we can show
-that the syntax xxx
+arbitrary model, but they hold in the syntax. In fact, we show that
+the single substitution syntax of (\ref{eq:tt}) is equivalent to the
+parallel substitution syntax of (\ref{eq:tt}), and the techniques that
+we use here are not specific to the SOGAT (\ref{eq:tt}), we expect
+that they work for any SOGAT. Equivalence of the syntaxes means that
+the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
+general SOGAT, there are more models of the correponding single
+substitution GAT than the parallel substitution GAT. Every parallel
+model is also a single model, but not necessarily the other way. While
+B-systems are equivalent to CwFs (and C-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
+substitution models of (\ref{eq:tt}) than parallel substitution
+models. The situation is similar to that of extensional combinatory
+calculus and lambda calculus, where the syntaxes are equivalent, and
+every lambda model is a combinatory model, but not the other way round
+\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
+relationship of monoids over a set $X$ and nil/cons algebras over $X$,
+where every monoid is a nil/cons-algebra, but not the other way; it is
+well-known that free monoids over $X$ (syntax for monoid over $X$) are
+the same as lists of $X$ (the syntax for nil/cons algebras over $X$).
 
-In fact, all the
-operations and equations of the parallel substitution calculus (CwF)
-are admissible in the syntax of SSC, moreover the set of syntactic SSC
-terms are isomorphic to the set of CwF terms. In contrast with
-B-systems which are equivalent to CwFs, our SSCs do not form
-CwFs. Every CwF-based model is a model of our SSC, but not the other
-way: there are more SSC-models than CwF-models. The situation is
-analogous to the relationship of extensional combinatory calculus and
-lambda calculus \cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic
-terms form isomorphic sets, but there are more combinatory models than
-lambda calculus models.
+When showing that the parallel and single substitution syntaxes are
+equivalent, we define the parallel operations by induction on the
+single substitution syntax. One of the operations in the single
+substitution syntax is the instantiation operation: $\blank[\blank] :
+\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
+terms. When doing induction on types/terms, we always have to provide
+a method for these operations, but this can be difficult as
+substitutions don't compose. This is why we derive a new induction
+principle which works on $\alpha$-normal types and terms. Just as
+$\beta$-normal forms don't distinguish between $\beta$-equal terms,
+$\alpha$-normal forms don't include explicit substitutions, all the
+variables are instantiated. $\alpha$-normal forms are not usual normal
+forms in the sense that they are still quotiented by the
+computation/uniqueness rules like $\beta$/$\eta$ for function
+space. To derive the new induction principle for the syntax, we first
+prove $\alpha$-normalisation.
 
-We define the parallel substitution calculus operations by induction
-on the SSC-syntax. One of our constructors is the instantiation
-operation of single substitutions: $\blank[\blank] :
-\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
-terms. When doing induction on terms, we always have to provide a
-method for these constructors, but this can be difficult as
-substitutions don't compose. This is why we derive a new induction
-principle which works on $\alpha$-normal forms. Just as $\beta$-normal
-forms don't distinguish between $\beta$-equal terms, $\alpha$-normal
-forms don't include explicit substitutions, all the variables are
-instantiated. $\alpha$-normal forms are not usual normal forms in the
-sense that they are still quotiented by the computation/uniqueness
-rules like $\beta$/$\eta$ for function space.
+if we have U,Pi, we can minimise the syntax
 
-In this paper, we define the SSC and extend it with
-$\Pi$ types and a hierarchy of Coquand-universes
-\cite{coquandUniverse}. The rules for type formers are the same as for
-CwFs, the difference is only in the substitution calculus part. More
-generally, every second-order generalised algebraic theory (SOGAT) has
-a parallel and a single substition-based first order presentation
-\cite{DBLP:conf/fscd/KaposiX24}. Here we prove that the syntaxes of
-the parallel and the single substitution calculi are isomorphic for
-the following SOGAT
-
-We conjecture that the methods of in this paper can be used to show
-that for any SOGAT, the two first-order presentations give isomorphic
-syntaxes.
+if we have U,Pi,Sigma, then SSC semantics of (1)+Sigma -> PSC semantics of (1)+Sigma.
 
@@
 \item A new generalised algebraic presentation of type theory in the
-  form of a minimalistic single substitution calculus (SSC).
-\item The $\alpha$-normalisation technique which is useful for working
-  with explicit substitution calculi.
-\item The proof of equivalence of the SSC-syntax with the CwF-based
-  syntax.
+  form of a minimalistic single substitution calculus.
+\item The $\alpha$-normalisation technique and deriving parallel
+  substitutions from the single substitution syntax.
+\item Minimisation.
+\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
+  a parallel model from a single substitution model.
 \end{itemize}
@@
 
+There are several ways to define type theory as a GAT. For example,
+Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
+categories with families (CwFs
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
+or C-systems
+\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
+natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
+B-systems, the listed notions all define a parallel substitution
+calculus, where contexts and substitutions form a category.
+
 Ehrhard
@@
 we use Sub, but we could have used telescopes
-
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}
 
@@
 
-\subsection{Formalisation}
+In a type-theoretic metatheory,
+the initial model of a GAT is a quotient inductive-inductive type
+(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
+family of inductive types where multiple types are defined mutually,
+later types can be indexed over previous ones, and equality
+constructors are also allowed, generalising quotients. The elimination
+principle of the QIIT corresponds to a structural induction principle
+and is equivalent to the statement that the syntax is the initial
+model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
+unique morphism from the syntax to any other model.
+
+\subsection{Metatheory and formalisation}
 
@@
 
+Prop.
+
 \section{Single substitution syntax}
@@
 Contents of the TYPES abstract, but with more details.
+\begin{alignat*}{10}
+& \Con && : \Set \\
+& \Ty && : \Con\ra\N\ra\Set \\
+& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
+& \diamond && : \Con \\
+& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
+& \Sub && : \Con\ra\Con\ra\Set \\
+& \blank[\blank] && : \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta \\
+& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
+& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
+& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
+& \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
+& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
+& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
+& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
+& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
+& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
+& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
+& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
+& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
+& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
+& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
+& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
+& \q[^+] && : \q[\gamma^+] = \q \\
+& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
+& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
+& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
+& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
+& \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
+\end{alignat*}
 

### 2024-09-14 types_2024_postproc/p.tex
stuff

@@
 \newcommand{\ra}{\rightarrow}
+\newcommand{\Ra}{\Rightarrow}
 \newcommand{\Set}{\mathsf{Set}}
@@
 \newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
+\renewcommand{\tt}{\mathsf{tt}}
+\newcommand{\fst}{\mathsf{fst}}
+\newcommand{\snd}{\mathsf{snd}}
 
@@
 models, our presentation gives a new notion of model, which in the
-future might contribute to definitions of new models. Because our
-single substitution GAT does not involve a category, it might be
-easier to adapt it in a higher dimensional setting, see Section
+future might contribute to definitions of new models. It can also
+influence implementations of type theory. Because our single
+substitution GAT does not involve a category, it might be easier to
+adapt it to a higher dimensional setting, see Section
 \ref{sec:conclusion}.
@@
 
-if we have U,Pi, we can minimise the syntax
-
-if we have U,Pi,Sigma, then SSC semantics of (1)+Sigma -> PSC semantics of (1)+Sigma.
+In the particular theory (\ref{eq:tt}), every type is represented by a
+term. In the first order single substitution calculus, this can be
+used to adapt some equations and remove some others, obtaining a more
+minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
+with unit and $\Sigma$ types, more precisely with the following rules.
+\begin{equation}\label{eq:sigma}
+  \begin{alignedat}{10}
+    & \top && : \Ty\,i \\
+    & \tt && : \Tm\,\top \\
+    & \top\eta && : t = \tt \\
+    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
+    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
+\end{alignedat}
+\end{equation}
+Using the same insight about types and the universe, the fact that
+dependences can be represented by function spaces, and that
+collections of dependencies can be collected using $\Sigma$ types, we
+can actually derive a parallel model from a single substitution model.
+TODO: universe levels are a problem (maybe not if we use iterated
+function space instead of $\Sigma$); the resulting CwF is not the same as
+the free CwF over the SSC. Bootstrapping a CwF.
+CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
+SSC -> CwF -> SSC = we get a stronger thing than what we started with.
 
@@
 
-Contents of the TYPES abstract, but with more details.
+In this section, we introduce the syntax of type theory with function
+space and universes in a minimalistic way, only introducing operations
+that are unavoidable. We eschew boilerplate by only defining
+well-formed, well-scoped, well-typed terms that are quotiented by
+conversion. This means that the equalities that hold between terms are
+the ways we can convert terms to each other when running them as
+programs. We give a tutorial-style introduction to the syntax, we do
+not assume prior knowledge of the metatheory of type theory. The
+definition of type theory that we obtain is the same as the single
+substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
+SOGAT (\ref{eq:tt}).
+
+First of all, we need a sort of terms which have to be indexed by
+types because we only want well-typed terms. Both types and terms can
+include variables, and to keep track of the currently available
+variables, we also index them by contexts: a context is a list of
+types, the length is the number of available variables, and we add new
+variables at the end (that is, they are snoc-lists and not
+cons-lists). Types are also indexed by their universe level, this is a
+technical requirement for avoiding Russell's paradox. The index $i$ is
+an implicit argument of $\Tm$.
 \begin{alignat*}{10}
@@
 & \Ty && : \Con\ra\N\ra\Set \\
-& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
+& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
+\end{alignat*}
+Just as lists have two constructors, there are two ways to form
+contexts: the empty context $\diamond$ and context extension $\ext$
+which is like the snoc operation for lists. Context extension takes a
+type which can have variables in the context preceeding the type.
+\begin{alignat*}{10}
 & \diamond && : \Con \\
-& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\ra\Con \\
+& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
+\end{alignat*}
+We will refer to variables by their distance from the end of the
+context. We would like to describe the operation providing the last
+variable in a context, the zero De Bruijn index, we denote it by $\q :
+\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
+should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
+: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
+such an $A'$. For this, we will introduce an operation $\blank[\p] :
+\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
+of introducing just $\blank[\p]$, we generalise a bit and add a new
+sort $\Sub$ which for now only has the single element $\p$ and the
+operation $\blank[\blank]$ is called \emph{instantiation}. We
+instantiate the type with the weakening in
+$\Sub\,\Delta\,\Gamma$. Elements of $\Sub$ will be in general called
+substitutions, hence the name, but for now, we only have single end-of
+context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
+the codomain of a $\Sub\,\Delta\,\Gamma$.
+\begin{alignat*}{10}
 & \Sub && : \Con\ra\Con\ra\Set \\
-& \blank[\blank] && : \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta \\
-& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
 & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
-& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
-& \Pi && : (A:\Ty\,\Gamma)\ra\Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
-& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
-& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
-& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
-& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
-& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
-& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
-& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
-& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
-& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
+& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
+& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
+\end{alignat*}
+The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
+$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
+implicitly. Still, we only have the last variable $\q$ in the context,
+we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
+A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
+with its context. To obtain more variables, we also allow weakening of
+terms, more precisely, we introduce an instantiation operation for
+terms with an overloaded notation.
+\begin{alignat*}{10}
+& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
+\end{alignat*}
+Note that this is a dependent function as the type of the resulting
+term has to be weakened the same way as the term itself. Now we can
+define all variables counting from the end of the context (De Bruijn
+indices \cite{debruijn}): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
+$3 := \q[\p][\p][\p]$, and so on.
+
+Next we add dependent function space $\Pi$: the domain of a dependent
+function is a type in some context $\Gamma$, and the codomain can also
+refer to a variable in the domain, so we extend the context of the
+codomain type with the type of the domain. For simplicity, both types
+are at the same level.
+\begin{alignat*}{10}
+& \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i
+\end{alignat*}
+The nondependent function space $\blank\Ra\blank :
+\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i$ is a special case of $\Pi$
+and we define it as the abbreviation $A\Ra B := \Pi\,A\,(B[\p])$.
+
+Because we introduced weakening, we now have to explain what happens
+to a $\Pi$ type once we weaken it. Assume $A : \Ty\,\Gamma\,i$,
+$B:\Ty\,(\Gamma\ext A)\,i$ and we have $\p:\Sub\,(\Gamma\ext
+C)\,\Gamma$. Then we say $(\Pi\,A\,B)[\p] = \Pi\,(A[\p])\,(B[\p'])$,
+but $B$ cannot be weakened by $\p$ because there we need $\p' :
+\Sub\,(\Gamma\ext C\ext A[\p])\,(\Gamma\ext A)$. So $p'$ has to be a
+new kind of weakening which adds a new variable in the last but one
+position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
+introducing a new weakening which adds a new variable in the last but
+two position, and so on. We solve all of these issues by allowing the
+\emph{lifting} of a weakening: $\gamma^+$ will be the same as the
+weakening $\gamma : \Sub\,\Delta\,\Gamma$, except that it will add an
+extra variable both at the end of the domain and codomain
+context.
+\begin{alignat*}{10}
+& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A)
+\end{alignat*}
+Note that $A$ is an implicit argument of $\blank^+$, and that it has
+to be instantiated by the weakening $\gamma$ in the domain context in
+order to fit in. Now we can explain how instantiation acts on $\Pi$ by
+the following equation.
+\begin{alignat*}{10}
+& \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+])
+\end{alignat*}
+The above equation has 6 implicit arguments, namely $\Gamma$, $i$,
+$A$, $B$, $\Delta$ and $\gamma$.  This rule works for any weakening,
+no matter how many $\blank^+$s have been applied to it. As of now, all
+elements of $\Sub$ have the form
+\begin{alignat*}{10}
+  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
+  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
+\end{alignat*}
+where $+^n$ means the $n$-times iteration of $\blank^+$.
+
+Now that we have weakenings of the form $\gamma^+$, we have to say how
+they act on variables, that is, terms of the form
+$\q[\p]\dots[\p]$. We express this using two rules: we say what
+$\q[\gamma^+]$ computes to, and what $b[\p][\gamma^+]$ computes to
+where $b$ is an arbitrary term. $\q[\gamma^+]$ should be the same as
+$\q$ (with different implicit arguments as the $\q$ in
+$\q[\gamma^+]$), as the weakening happens somewhere in the middle of
+the context, so the index of the variable remains unchanged. However,
+assuming $\q:\Tm\,(\Gamma\ext A)\,(A[\p])$, we have $\q[\gamma^+] :
+\Tm\,(\Delta\ext A[\gamma])\,(A[\p][\gamma^+])$, but in the same
+context, we have $\q : \Tm\,(\Delta\ext A[\gamma])\,(A[\gamma][\p])$,
+hence the terms in the two sides of the equation $\q[\gamma^+] = \q$
+have different types. But these two types should be the same:
+weakening a type at the end of the context, and then applying another
+lifted weakening should be the same as first weakening somewhere and
+then weakening at the end. We first assume this equation for types,
+then the equation $\q[\gamma^+] = \q$ becomes well-typed (in the
+metatheory). The equation for $b[\p][\gamma^+]$ has the same shape as
+the newly assumed rule for types. Thus we add the following three
+equations.
+\begin{alignat*}{10}
 & [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
 & [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
-& \q[^+] && : \q[\gamma^+] = \q \\
+& \q[^+] && : \q[\gamma^+] = \q
+\end{alignat*}
+The second and third equations only make sense because of the first
+equation. The phenomenon that later equations (or operations) depend
+on previous equations is usual in the world of GATs. In fact, the
+later equations cannot even be stated without having some previous
+equations.
+
+Now that we have $[\p][^+]$ for types, we can derive the instantiation
+rule for nondependent function space as follows.
+\begin{alignat*}{10}
+  & (A\Ra B)[\gamma] && {=} \\
+  & (\Pi\,A\,(B[\p]))[\gamma] && {=}(\Pi[]) \\
+  & \Pi\,(A[\gamma])\,(B[\p][\gamma^+])\,\, && {=}([\p][^+]) \\
+  & \Pi\,(A[\gamma])\,(B[\gamma][\p]) && {=} \\
+  & (A[\gamma])\Ra(B[\gamma])
+\end{alignat*}
+
+So far our only terms are variables, but we would like to define
+functions via lambda abstraction. Abstraction takes a term in a
+context extended by the domain of the function. It comes with a rule
+for instantiation analogous to $\Pi[]$.
+\begin{alignat*}{10}
+& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
+& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+])
+\end{alignat*}
+Note that the equation $\lam[]$ only makes sense because of the
+previous equation $\Pi[]$: the left hand side is in
+$\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
+$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
+right hand side is in
+$\Tm\,\Delta\,(\Pi\,(A[\gamma])\,(B[\gamma^+]))$.
+
+The functions $\lambda x.x$ and $\lambda x\,y.x$ can be defined in our
+syntax as $\lam\,\q : \Tm\,\Gamma\,(A\Ra A)$ and
+$\lam\,(\lam\,(\q[\p])) :
+\Tm\,\Gamma\,\big(\Pi\,A\,(B\Ra(A[\p]))\big)$ which make sense for all
+$\Gamma$, $A$, $B$.
+
+The application operation for dependent function space is a bit
+tricky, because the return type depends on the input: the argument of
+the function will appear in the return type. We write application by
+an infix $\blank\cdot\blank :
+\Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,B'$,
+where $B' : \Ty\,\Gamma$ should be $B:\Ty\,(\Gamma\ext A)$ where the
+last variable is \emph{substituted} (instantiated) by $a$. For this,
+we introduce a new element of $\Sub$ called single substitution which
+goes the opposite way of $\p$. Now we can use instantiation
+$\blank[\blank]$ to substitute the last variable in $B$.
+\begin{alignat*}{10}
+& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
+& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle])
+\end{alignat*}
+We could have introduced new, separate sorts and $\blank[\blank]$
+operations for weakenings and substitutions, but we merge them for
+simplicity. There is no need for separation: just as weakenings can be
+lifted, single substitutions can also be lifted, and the
+weakening-rules $\Pi[]$, $\lam[]$ also work for single
+substitutions. We won't have more ways to introduce elements of
+$\Sub$, and there are no equations on $\Sub$. An element of $\Sub$ is
+either a single weakening $\p$ lifted a couple of times, or a single
+substitution lifted a couple of times. We call elements of $\Sub$
+substitutions for simplicity.
+
+With the introduction of $\blank\cdot\blank$, we need a new
+substitution rule, but again it only makes sense with an extra
+equation on types saying that first substituting the last variable and
+then an arbitrary substitution is the same as first the lifted version
+of the arbitrary substitution that does not touch the last variable,
+and then substituting the last variable.
+\begin{alignat*}{10}
+& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
+& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
+\end{alignat*}
+Following the introduction of the operation $\langle\blank\rangle$, we
+need to explain how it acts on variables. Given a variable in the
+middle of the context (a term $b[\p]$), substituting its last variable
+simply returns $b$. Substituting the last variable $\q$ reads out the
+term from $\langle\blank\rangle$. As usual, the rules only typecheck
+if we have an equation for types.
+\begin{alignat*}{10}
 & [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
 & [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
-& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
+& \q[\langle\rangle] && : \q[\langle a\rangle] = a
+\end{alignat*}
+We don't have a separate sort of variables, so the rule
+$[\p][\langle\rangle]$ holds not only for variables, but for arbitrary
+terms. This is not an issue, as weakening a term at the end of its
+context, and then substituting the newly introduced variable is the
+same as not doing anything.
+
+We add the computation and uniqueness rule for function space, where
+the uniqueness rule again needs an extra equation on the codomain type
+of the function to make sense. This is the typal version of the
+equation mentioned in the introduction.
+\begin{alignat*}{10}
+& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
 & [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
-& \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
+& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
+\end{alignat*}
+
+Finally, we add the rules for universes. A universe is a type
+containing codes for types, this is witnessed by $\El$ (elements) and
+$\cd$ (code), which make an isomorphism between terms of type $\U\,i$
+and types of level $i$. All three operations come with substitution
+rules ($\El[]$ and $\cd[]$ are interderivable, so it would be enough
+to assume one of the two, but we add both to completely align with the
+algorithm \cite{DBLP:conf/fscd/KaposiX24} generating the rules from
+the SOGAT (\ref{eq:tt})).
+\begin{alignat*}{10}
+  & \U && : (n:\N)\ra\Ty\,\Gamma\,(1+n) \\
+  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
+  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
+  & \U\beta && : \El\,(\cd\,A) = A \\
+  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
+  & \U[] && : (\U\,i)[\gamma] = \U\,i \\
+  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
+  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
 \end{alignat*}
@@
 TODO: show that the SSC operations are derivable from CwF.
-
-\section{Isomorphism}
-\label{sec:iso}
 

### 2024-09-14 types_2024_postproc/p.tex
sutff

@@
 \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows}
 
@@
 
-In Section \ref{sec:ssc}, we introduce the single substitution
+TODO. In Section \ref{sec:ssc}, we introduce the single substitution
 calculus only relying on a working knowledge of an implementation of
@@
 \cite{voevodsky2014bsystems} as an algebraic way of describing type
-theories close to the notations using typing judgements. Using our
-notations B-systems can be described as follows. B-frame.
+theories close to the notations using typing judgements. B-systems are
+a generalised algebraic (intrinsic) presentation of type theory using
+single substitutions. xxx
 
-There are several ways to define type theory as a GAT. For example,
-Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
-categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
-or C-systems
-\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
-natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.  With the exception of
-B-systems, the listed notions all define a parallel substitution
-calculus, where contexts and substitutions form a category.
+\begin{tikzpicture}
+  \node (b0) at (0,0) {$\top\cong B_0$};
+  \node (b1) at (2,0) {$B_1$};
+  \node (b2) at (4,0) {$B_2$};
+  \node (b3) at (6,0) {$\dots$};
+  \node (c1) at (3,1) {$\tilde{B}_1$};
+  \node (c2) at (5,1) {$\tilde{B}_1$};
+  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
+  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
+  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
+  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
+  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
+\end{tikzpicture}
 
-Ehrhard
+Using our
+notations B-systems can be described as follows. B-frame. TODO
 
@@
 
-contextual sCwF and contextual SSC with fct types
+The first generalised algebraic presentation of type theory was Thomas
+Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
+parallel substitution calculus with $\p$, $\blank^+$ and
+$\langle\blank\rangle$ operations, just like in our calculus. Our
+single substitution calculus is essentially Ehrhard's calculus with
+the categorical composition and identity operations
+removed. Categories with families (CwFs
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
+$\blank,\blank$ operations which makes it more apparent that
+substitutions are simple lists of terms. CwFs are equivalent to
+Ehrhard's calculus, and so are contextual categories/C-systems
+\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
+the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
+B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.
 
-Formalisations of type theory use low-level
+Kaposi and Luksa \cite{} defined a telescopic single substitution
+calculus for simple type theory, it can be seen as a simply typed
+version of B-systems. They show that the category of contextual models
+of their calculus is equivalent to the category of contextual simply
+typed CwFs with function space. Their equivalence result holds even
+without the presence of function space. In the presence of function
+space, the same equivalence holds for the simply typed version of our
+substitution calculus, see the formalisation accompanying this paper.
 
-Algebraic formalisations of type theory: Brunerie
+Most formalisations of type theory are extrinsic
+\cite{pujet,abel,metacoq}: they define the syntax as abstract syntax
+trees, and equip it with typing and conversion relations. A higher
+level representation is working in setoid hell
+\cite{chapman09eatitself}: terms are intrinsically well-typed, but
+conversion is still an explicit relation. For formalising the
+GAT-level syntax, one needs a stronger metatheory than ordinary Agda
+or Coq: the metatheory has to support quotient inductive-inductive
+types (QIITs \cite{DBLP:journals/pacmpl/KaposiKA19}), in other words,
+initial models of GATs. Altenkirch and Kaposi
+\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
+theory using postulated QIITs in Agda, together with parametricity and
+normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
+\cite{initiality-agda} constructed the initial contextual category in
+Agda using postulated quotients. QIITs are supported by the cubical
+set model \cite{} and the setoid model \cite{kaposi-qiit-setoid}, thus
+Cubical Agda \cite{} and setoid/observational type theory
+\cite{setoid,loicThesis} support QIITs. For example, in Cubical Agda,
+set-truncated and groupoid-truncated syntaxes of type theory have been
+formalised \cite{cohtt}. Without QIITs, tricks such as shallow
+embedding the syntax can be used to formalise metatheoretic results
+about type theory such as gluing
+\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
+canonicity and parametricity \cite{kaposi-shallow}.
 
-Initiality conjecture.
-
-\cite{initiality-agda} \cite{DBLP:conf/popl/AltenkirchK16},
-normalisation, shallow embedding canonicity, gluing. Low level
-formalisations: Abel, Loic, metacoq. They can be seen as unfolding the
-more abstract thing in a (weak) setoid model.
-
-Other SSC: 90s French, Thorsten thesis, B-systems, Luksa paper.
+At the level of abstraction of SOGATs, the difference between single
+and parallel substitution calculi are not visible. However, the
+metatheory of type theory can be still investigated in this abstract
+setting, via the methods of Synthethic Tait Computability
+\cite{DBLP:phd/us/Sterling22} or internal sconing
+\cite{DBLP:conf/fscd/BocquetKS23}.
 
@@
 % paper
-
-In a type-theoretic metatheory,
-the initial model of a GAT is a quotient inductive-inductive type
-(QIIT \cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general
-family of inductive types where multiple types are defined mutually,
-later types can be indexed over previous ones, and equality
-constructors are also allowed, generalising quotients. The elimination
-principle of the QIIT corresponds to a structural induction principle
-and is equivalent to the statement that the syntax is the initial
-model \cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a
-unique morphism from the syntax to any other model.
 
@@
 
+no need for parallel substitutions, empty substitution, parallel
+weakenings, telescopes, or combinations of these.
+
+say that we assume the QIIT version of this.
+
 \section{$\alpha$-normalisation}
@@
 \section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
+
+termification in the shallow paper is shown equationally complete
 

### 2024-09-14 types_2024_postproc/p.tex
citations

@@
 
-Kaposi and Luksa \cite{} defined a telescopic single substitution
+Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
 calculus for simple type theory, it can be seen as a simply typed
@@
 Most formalisations of type theory are extrinsic
-\cite{pujet,abel,metacoq}: they define the syntax as abstract syntax
-trees, and equip it with typing and conversion relations. A higher
-level representation is working in setoid hell
-\cite{chapman09eatitself}: terms are intrinsically well-typed, but
-conversion is still an explicit relation. For formalising the
-GAT-level syntax, one needs a stronger metatheory than ordinary Agda
-or Coq: the metatheory has to support quotient inductive-inductive
-types (QIITs \cite{DBLP:journals/pacmpl/KaposiKA19}), in other words,
-initial models of GATs. Altenkirch and Kaposi
+\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
+they define the syntax as abstract syntax trees, and equip it with
+typing and conversion relations. A higher level representation is
+working in setoid hell \cite{chapman09eatitself}: terms are
+intrinsically well-typed, but conversion is still an explicit
+relation. For formalising the GAT-level syntax, one needs a stronger
+metatheory than ordinary Agda or Coq: the metatheory has to support
+quotient inductive-inductive types (QIITs
+\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
+models of GATs. Altenkirch and Kaposi
 \cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
@@
 Agda using postulated quotients. QIITs are supported by the cubical
-set model \cite{} and the setoid model \cite{kaposi-qiit-setoid}, thus
-Cubical Agda \cite{} and setoid/observational type theory
-\cite{setoid,loicThesis} support QIITs. For example, in Cubical Agda,
-set-truncated and groupoid-truncated syntaxes of type theory have been
-formalised \cite{cohtt}. Without QIITs, tricks such as shallow
-embedding the syntax can be used to formalise metatheoretic results
-about type theory such as gluing
-\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
-canonicity and parametricity \cite{kaposi-shallow}.
+set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model \cite{kaposi-qiit-setoid}, thus
+Cubical Agda \cite{DBLP:journals/jfp/VezzosiMA21} and
+setoid/observational type theory \cite{setoid,DBLP:phd/hal/Pujet22} support
+QIITs. For example, in Cubical Agda, set-truncated and
+groupoid-truncated syntaxes of type theory have been formalised
+\cite{cohtt}. Without QIITs, tricks such as shallow embedding the
+syntax can be used to formalise metatheoretic results about type
+theory such as gluing \cite{kaposi_et_al:LIPIcs:2019:10532}, and its
+special cases canonicity and parametricity \cite{kaposi-shallow}.
 

### 2024-09-15 types_2024_postproc/p.tex
stuff

@@
 \usetikzlibrary{arrows}
+\usepackage{proof}
 
@@
 \newcommand{\snd}{\mathsf{snd}}
+\newcommand{\Lift}{\mathsf{Lift}}
+\newcommand{\mk}{\mathsf{mk}}
+\newcommand{\un}{\mathsf{un}}
 
@@
   & \U && : (i:\N)\ra\Ty\,(1+i) \\
-  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
+  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
+  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
+  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
 \end{alignedat}
 \end{equation}
-We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
-and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
-denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
-$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
-and $\lam$ are second-order functions. We say that $\Pi$ binds a
-$\Tm$-variable of type $A$ in its second $\Ty$-argument. Every binder
-is represented by a second-order function, and as a result, we don't
-need to talk about contexts and substitutions when describing the
-theory: we re-use the binding features of the metatheory to describe
-those of our object theory (this technique is also known as
-higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99} or
-logical framework
+Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
+we have to index types by their (universe) level, and types at lower
+level are included at all upper levels by $\Lift$. We use implicit
+arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
+$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
+isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
+and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
+second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
+type $A$ in its second $\Ty$-argument. Every binder is represented by
+a second-order function, and as a result, we don't need to talk about
+contexts and substitutions when describing the theory: we re-use the
+binding features of the metatheory to describe those of our object
+theory (this technique is also known as higher-order abstract syntax
+\cite{DBLP:conf/lics/Hofmann99} or logical framework
 \cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
@@
 \item A new generalised algebraic presentation of type theory in the
-  form of a minimalistic single substitution calculus.
+  form of a minimalistic single substitution calculus. We show that
+  there is no need for parallel substitutions, empty substitution,
+  parallel weakenings, telescopes, or combinations of these to
+  describe the syntax of type theory in an intrinsic way.
 \item The $\alpha$-normalisation technique and deriving parallel
@@
 but usually in the extrinsic setting, see e.g.\ the thesis of
-Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
-\cite{voevodsky2014bsystems} as an algebraic way of describing type
-theories close to the notations using typing judgements. B-systems are
-a generalised algebraic (intrinsic) presentation of type theory using
-single substitutions. xxx
+Altenkirch \cite{alti:phd93}.
 
+B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
+an algebraic way of describing type theories close to the notations
+using typing judgements. B-systems are an intrinsic, but essentially
+algebraic presentation of type theory using single
+substitutions. B-systems correspond to our single substitution
+calculus as essentially algebraic theories correpond to generalised
+algebraic theories, or sets with a map into $I$ correspond to indexed
+families over $I$ \cite[page
+  221]{DBLP:journals/apal/Cartmell86}. Another difference is that we
+have fewer and less general equations, resulting in the fact that we
+have more models than B-systems. However in the syntax of our theory,
+all the rules of B-systems are admissible. We describe the
+relationship in more detail. B-systems are B-frames together with a
+substitution, weakening and generic element operations. A B-frame is
+given by sets $B_i$, $\tilde{B}_i$ and functions between them as shown
+below \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. \\
 \begin{tikzpicture}
-  \node (b0) at (0,0) {$\top\cong B_0$};
-  \node (b1) at (2,0) {$B_1$};
+  \node (b0) at (0,0) {$\top$};
+  \node (b1) at (1.5,0) {$B_1$};
   \node (b2) at (4,0) {$B_2$};
-  \node (b3) at (6,0) {$\dots$};
-  \node (c1) at (3,1) {$\tilde{B}_1$};
+  \node (b3) at (7,0) {$\dots$};
+  \node (c1) at (2.5,1) {$\tilde{B}_1$};
   \node (c2) at (5,1) {$\tilde{B}_1$};
@@
   \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
-\end{tikzpicture}
-
-Using our
-notations B-systems can be described as follows. B-frame. TODO
-
-we use Sub, but we could have used telescopes
-
-The first generalised algebraic presentation of type theory was Thomas
-Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
-parallel substitution calculus with $\p$, $\blank^+$ and
-$\langle\blank\rangle$ operations, just like in our calculus. Our
-single substitution calculus is essentially Ehrhard's calculus with
-the categorical composition and identity operations
-removed. Categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
-$\blank,\blank$ operations which makes it more apparent that
-substitutions are simple lists of terms. CwFs are equivalent to
-Ehrhard's calculus, and so are contextual categories/C-systems
-\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
-the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
-B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.
+\end{tikzpicture} \\
+Using our notation, a B-frame contains the following data. \\
+\begin{tikzpicture}
+  \node (b0) at (0,0) {$\top$};
+  \node (b1) at (1.5,0) {$\Ty\,\diamond$};
+  \node (b2) at (4,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
+  \node (b3) at (7,0) {$\dots$};
+  \node (c1) at (2.5,1) [text width=3cm] {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
+  \node (c2) at (5.7,1.7) [text width=4cm,align=center] {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
+  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
+  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
+  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
+  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
+  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
+\end{tikzpicture} \\
+The substitution operation $\mathbb{S}$ for an element of
+$x:\tilde{B}_{n+1}$ is a homomorphism of the slice B-frames
+$\mathbb{B}/\partial(x) \ra \mathbb{B}/\mathsf{ft}(\partial(x))$. In
+our notation, $x = (\Gamma, A, a)$ where $a : \Tm\,\Gamma\,A$, and the
+homomorphism $\mathbb{S}$ corresponds to the following maps for the
+bottom row of the diagram:
+\begin{alignat*}{10}
+  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
+  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
+  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
+  & \dots,
+\end{alignat*}
+and similarly $\mathbb{S}$ also includes all the (lifted) substitution
+operations for terms corresponding to the top row of the
+diagram. Analogously, the weakening operation corresponds to
+$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
+our notation. The six groups of equations correspond to our
+$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
+$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
+order. However we don't have the lifted versions of our equations,
+equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
+stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
+for terms. The missing equations are admissible, see Section
+\ref{sec:admissible}. In the presence of universes and $\Pi$ types, we
+reduce the needed equations even more, see Section
+\ref{sec:minimisation}.
 
@@
 
-Most formalisations of type theory are extrinsic
+The first generalised algebraic presentation of type theory was Thomas
+Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
+parallel substitution calculus with $\p$, $\blank^+$ and
+$\langle\blank\rangle$ operations, just like in our calculus. Our
+single substitution calculus is essentially Ehrhard's calculus with
+the categorical composition and identity operations
+removed. Categories with families (CwFs
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
+$\blank,\blank$ operations which make it more apparent that
+substitutions are lists of terms. CwFs are equivalent to Ehrhard's
+calculus, and so are contextual categories/C-systems
+\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
+the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
+B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.
+
+Most computer formalisations of type theory are extrinsic
 \cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
@@
 Agda using postulated quotients. QIITs are supported by the cubical
-set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model \cite{kaposi-qiit-setoid}, thus
-Cubical Agda \cite{DBLP:journals/jfp/VezzosiMA21} and
-setoid/observational type theory \cite{setoid,DBLP:phd/hal/Pujet22} support
-QIITs. For example, in Cubical Agda, set-truncated and
-groupoid-truncated syntaxes of type theory have been formalised
-\cite{cohtt}. Without QIITs, tricks such as shallow embedding the
-syntax can be used to formalise metatheoretic results about type
-theory such as gluing \cite{kaposi_et_al:LIPIcs:2019:10532}, and its
-special cases canonicity and parametricity \cite{kaposi-shallow}.
+set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model
+\cite{kaposi-qiit-setoid}, thus Cubical Agda
+\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
+theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
+in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
+theory have been formalised \cite{cohtt}. Without QIITs, tricks such
+as shallow embedding the syntax can be used to formalise metatheoretic
+results about type theory such as gluing
+\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
+canonicity and parametricity \cite{kaposi-shallow}.
 
-At the level of abstraction of SOGATs, the difference between single
+At the level of abstraction of SOGATs
+\cite{uemura,DBLP:conf/fscd/KaposiX24}, the difference between single
 and parallel substitution calculi are not visible. However, the
@@
 
-Notation and formalisation techniques.
+Notation and formalisation techniques, including congruence rules.
 
@@
 codomain type with the type of the domain. For simplicity, both types
-are at the same level.
+are at the same level (this can be remedied using $\Lift$, see later).
 \begin{alignat*}{10}
@@
 \begin{alignat*}{10}
-& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
+& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
 & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
@@
 
-Finally, we add the rules for universes. A universe is a type
-containing codes for types, this is witnessed by $\El$ (elements) and
-$\cd$ (code), which make an isomorphism between terms of type $\U\,i$
-and types of level $i$. All three operations come with substitution
-rules ($\El[]$ and $\cd[]$ are interderivable, so it would be enough
-to assume one of the two, but we add both to completely align with the
-algorithm \cite{DBLP:conf/fscd/KaposiX24} generating the rules from
-the SOGAT (\ref{eq:tt})).
+We add the rules for universes. A universe is a type containing codes
+for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
+which make an isomorphism between terms of type $\U\,i$ and types of
+level $i$. All three operations come with substitution rules ($\El[]$
+and $\cd[]$ are interderivable, so it would be enough to assume one of
+the two, but we add both to completely align with the algorithm
+\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
+(\ref{eq:tt})).
 \begin{alignat*}{10}
-  & \U && : (n:\N)\ra\Ty\,\Gamma\,(1+n) \\
-  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
-  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
+  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
+  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
+  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
   & \U\beta && : \El\,(\cd\,A) = A \\
-  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
-  & \U[] && : (\U\,i)[\gamma] = \U\,i \\
-  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
-  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
+  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
 \end{alignat*}
+Finally, we add the rules for moving types one level up. Because we
+defined function space only between types at the same level, without
+lifting, we cannot even define the polymorphic identity function.
+\begin{alignat*}{10}
+  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
+  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
+  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
+  & \Lift\beta && : \un\,(\mk\,a) = a \\
+  & \Lift\eta && : \mk\,(\un\,a) = a
+\end{alignat*}
+This concludes the first-order single substitution semantics of the
+SOGAT \ref{eq:tt}. The theory can be extended with new type and term
+formers in the same fashion: each rule has to be indexed over
+arbitrary contexts and come equipped with a substitution rule. There
+is no need to add more structural (substitution calculus) rules.
 
-Show examples of using the calculus, and where we need to derive more
-stuff.
+The polymorphic identity function for types at the level 0 is defined
+as
+\[
+\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
+\]
+Note that readability of this term essentially relies on implicit
+arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
+$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
+argument. Inrinsically typed terms are the same as derivation trees,
+we illustrate this by deriving this term as follows.
+\[
+\infer[(*)]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
+\]
+Note that in step (*), $\lam$ made it sure that $\U\,0$ and
+$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
+this is why we had to lift $\El\,\q$. Another subtle thing is
+happening when deriving the third implicit argument of $\q$. We have
+to coerce $\q$ along the equality $\U[]$.
+\[
+\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
+\]
+When working informally (or in extensional type theory), we don't
+write the coercion, but Agda requires it.
 
-no need for parallel substitutions, empty substitution, parallel
-weakenings, telescopes, or combinations of these.
+\section{Admissible equations}
+\label{sec:admissible}
 
-say that we assume the QIIT version of this.
-
-\section{$\alpha$-normalisation}
-\label{sec:alpha}
+TODO: say that we assume the QIIT version of this.
 
 $\alpha$-normal forms could also be called substitution normal forms.
-
-\section{Isomorphism}
-\label{sec:cwf}
 
@@
 \section{Minimisation}
+\label{sec:minimisation}
 

### 2024-09-15 types_2024_postproc/p.tex
stuff

@@
 
+We conjecture that the same works for any SOGAT.
+
 \section{Minimisation}
 \label{sec:minimisation}
+
+In this section we show that if have we universes and $\Pi$ types
+(which is the case in the theory (\ref{eq:tt})), we can decrease the
+number of euqations.
+
+In our single substitution calculus, we have the equation $[\p][^+]$
+stated both for types and terms (we know that the left one is about
+types because of our convention that type metavariables are uppercase
+Latin letters and term metavariables are lowercase):
+\[
+B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
+\]
+We even need the first equation to typecheck the second one. We made
+the dependency explicit by adding a subscript of the equality in the
+equation for terms (in Agda, this dependency has to be made
+explicit). An alternative presentation of the second equation without
+requiring the first one is the following:
+\[
+[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
+\]
+That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
+we have that for any $b$ we have the equation (suitably over the input
+equation). Thus we add a conditional equation.
+
+It turns out that if we have Coquand-universes, the conditional
+$[\p][^+]'$ rule is enough: the input equation holds for $B := \U\,i$
+via $\U[]$, thus we get that for any $\hat{A} : \Tm\,\Gamma\,(\U\,i)$,
+$\hat{A}[\p][\gamma^+] = \hat{A}[\gamma][\p]$. But every type has a
+code in the universe, so for any $B : \Ty\,\Gamma\,i$,
+\begin{alignat*}{10}
+  & B[\p][\gamma^+] && {=}(\U\beta) \\
+  & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
+  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]') \\
+  & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
+  & \El\,(\cd\,B)[\gamma][\p] && {=}(\U\beta) \\
+  & B[\gamma][\p].
+\end{alignat*}
+Equation $[\q][^+] : \q[\gamma^+] = \q$ only makes sense if we have
+$[\p][^+]$ for types, so either we make $[\q][^+]$ conditional, or we
+coerce it along the derived equation.
+
+We play the exact same game with $[\p][\langle\rangle]$: we replace it
+with the conditional equation
+\[
+[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b,
+\]
+notice that we have the input for $B = \U\,i$, and derive the input
+equation for all types.
+
+Now we turn our attention to the equation
+\[
+  [\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle].
+\]
+We were forced to introduce it so that we can state the substitution
+rule for function application $\cdot[]$. What if we made $\cdot[]$
+itself conditional?
+
+xxx
 
@@
 
+\pagebreak
+
+\appendix
+
+\section{The single substitution GAT presentation of the SOGAT (\ref{eq:tt})}
+
+For reference, we list the rules of the theory introduced in Section
+\ref{sec:ssc}. The core substitution calculus:
+\begin{alignat*}{10}
+& \Con && : \Set \\
+& \Ty && : \Con\ra\N\ra\Set \\
+& \diamond && : \Con \\
+& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
+& \Sub && : \Con\ra\Con\ra\Set \\
+& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
+& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
+& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
+& \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
+& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
+& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
+& \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
+& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
+& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
+& \q[^+] && : \q[\gamma^+] = \q \\
+& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
+& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
+& \q[\langle\rangle] && : \q[\langle a\rangle] = a \\
+& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
+& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B
+\end{alignat*}
+Rules for individual type formers.
+\begin{alignat*}{10}
+  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
+  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
+  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
+  & \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
+  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
+  & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
+  & \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
+  & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
+  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
+  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
+  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
+  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
+  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
+  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
+  & \U\beta && : \El\,(\cd\,A) = A \\
+  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
+\end{alignat*}
+\begin{alignat*}{10}
+  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
+  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
+  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
+  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
+  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
+  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
+  & \Lift\beta && : \un\,(\mk\,a) = a \\
+  & \Lift\eta && : \mk\,(\un\,a) = a
+\end{alignat*}
+
 \end{document}

### 2024-09-15 types_2024_postproc/p.tex
minimisation

@@
 the operations are all easy to motivate, we illustrate this in Section
-\ref{sec:ssc}. Single substitution calculi are closer to the usual
+\ref{sec:tt}. Single substitution calculi are closer to the usual
 presentation of type theory using derivation rules, this was one of
@@
 
-TODO. In Section \ref{sec:ssc}, we introduce the single substitution
+TODO. In Section \ref{sec:tt}, we introduce the single substitution
 calculus only relying on a working knowledge of an implementation of
@@
 \section{Single substitution syntax}
-\label{sec:ssc}
+\label{sec:tt}
 
@@
 & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
-& \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+])
+& \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
 \end{alignat*}
@@
 \begin{alignat*}{10}
-& \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
+& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
 & [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
@@
 
+TODO: beta equation for id.
+
 \section{Admissible equations}
@@
 with a hierarchy of Coquand-universes and $\Pi$ types can be derived
-from the single substitution syntax defined in \autoref{sec:ssc}.
+from the single substitution syntax defined in \autoref{sec:tt}.
 
@@
 (which is the case in the theory (\ref{eq:tt})), we can decrease the
-number of euqations.
+number of equations.
 
-In our single substitution calculus, we have the equation $[\p][^+]$
+In our single substitution calculus, we have equation $[\p][^+]$
 stated both for types and terms (we know that the left one is about
@@
 $\hat{A}[\p][\gamma^+] = \hat{A}[\gamma][\p]$. But every type has a
-code in the universe, so for any $B : \Ty\,\Gamma\,i$,
+code in the universe, so for a $B : \Ty\,\Gamma\,i$,
 \begin{alignat*}{10}
@@
   & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
-  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]') \\
+  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]'\,\U[]) \\
   & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
@@
 \end{alignat*}
-Equation $[\q][^+] : \q[\gamma^+] = \q$ only makes sense if we have
-$[\p][^+]$ for types, so either we make $[\q][^+]$ conditional, or we
-coerce it along the derived equation.
+Equation $[\q][^+] : \q[\gamma^+] = \q$ also only makes sense if we
+have $[\p][^+]$ for types, so either we have to make $[\q][^+]$
+conditional, or we coerce it along the derived equation.
 
@@
 notice that we have the input for $B = \U\,i$, and derive the input
-equation for all types.
+equation for all types. We make $\q[\langle\rangle]$ conditional as
+well.
 
@@
 itself conditional?
+\begin{alignat*}{10}
+  & \cdot[]' :{} && (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
+  & && (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma])
+\end{alignat*}
+Again, for $B = \U$ we have the assumption $e$. But then $t$ is in
+$\Tm\,\Gamma\,(\Pi\,A\,\U)$, which is isomorphic to $\Ty\,(\Gamma\ext
+A)$. So we argue as follows for any $B : \Ty\,(\Gamma\ext A)$.
+\begin{alignat*}{10}
+  & B[\langle a\rangle][\gamma] && {=}(\U\beta) \\
+  & \El\,(\cd\,(B[\langle a\rangle][\gamma])) && {=}(\cd[]) \\
+  & \El\,((\cd\,B)[\langle a\rangle][\gamma]) && {=}(\Pi\beta) \\
+  & \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) && {=}(\cdot[]'\,\U[]) \\
+  & \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma]))\,\, && {=}(\lam[]) \\
+  & \El\,(\lam\,(\cd\,B[\gamma^+])\cdot (a[\gamma])) && {=}(\Pi\beta) \\
+  & \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) && {=}(\cd[]) \\
+  & \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] && {=}(\U\beta) \\
+  & B[\gamma^+][\langle a[\gamma]\rangle]
+\end{alignat*}
+Hence, the conditional $\cdot[]'$ equation implies the condition for
+all types.
 
-xxx
+Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
+$\Pi\eta$ conditional. This needs another change: we exchange
+$\Pi\beta$ for a more general variant which is also conditional on the
+same equation.
+\begin{alignat*}{10}
+  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
+  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
+\end{alignat*}
+But first we need to check that the usage of $\Pi\beta$ above when
+proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
+  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
+$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
+\begin{alignat*}{10}
+  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
+  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
+  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
+  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
+  & && \hat{B}[\langle a\rangle].
+\end{alignat*}
+Now we derive the assumption for any type $B : \Ty\,(\Gamma\ext A)$.
+\begin{alignat*}{10}
+  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
+  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
+  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
+  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
+  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
+  & \El\,(\cd\,B) && {=}(\U\beta) \\
+  & B
+\end{alignat*}
+With this we derived an even more minimalistic first-order theory for
+the SOGAT (\ref{eq:tt}), essentially relying on the universe and $\Pi$
+types. It is clear from the construction that the resulting theory is
+isomorphic to the one presented in Section \ref{sec:tt} (and Appendix
+\ref{app:tt}), all equations are interderivable. For reference, we
+list all the rules of the minimised calculus in Appendix
+\ref{app:minimisation}.
 
@@
 termification in the shallow paper is shown equationally complete
+
+intuition, similar to minimisation: the subst calc is just there to
+bootstrap U and Pi, those can do everything.
 
@@
 
-\section{The single substitution GAT presentation of the SOGAT (\ref{eq:tt})}
+\section{GAT presentation of (\ref{eq:tt})}
+\label{app:tt}
 
 For reference, we list the rules of the theory introduced in Section
-\ref{sec:ssc}. The core substitution calculus:
+\ref{sec:tt}. The core substitution calculus:
 \begin{alignat*}{10}
@@
   & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
-  & \lam[] && : (\lam\,t)[\gamma] = \lam\,(t[\gamma^+]) \\
+  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
   & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
   & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
-  & \Pi\beta && : \lam\,t\cdot a = t[\langle a\rangle] \\
+  & \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
   & \Pi\eta && : t = \lam\,(t[\p]\cdot\q) \\
+  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
+  & \U[] && : (\U\,i)[\gamma] = \U\,i \\  
+  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
+  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
+  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
+  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
+  & \U\beta && : \El\,(\cd\,A) = A \\
+  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
+  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
+  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])
+\end{alignat*}
+\begin{alignat*}{10}
+  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
+  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
+  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
+  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
+  & \Lift\beta && : \un\,(\mk\,a) = a \\
+  & \Lift\eta && : \mk\,(\un\,a) = a
+\end{alignat*}
+
+\section{Minimised GAT presentation of (\ref{eq:tt})}
+\label{app:minimisation}
+
+We list the rules of the theory derived in Section
+\ref{sec:minimisation}.
+\begin{alignat*}{10}
+  & \Con && : \Set \\
+  & \Ty && : \Con\ra\N\ra\Set \\
+  & \diamond && : \Con \\
+  & \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con \\
+  & \Sub && : \Con\ra\Con\ra\Set \\
+  & \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set \\
+  & \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
+  & \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
+  & \blank^+ && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta\ext A[\gamma])\,(\Gamma\ext A) \\
+  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
+  & \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma]) \\
+  & \q && : \Tm\,(\Gamma\ext A)\,(A[\p]) \\
+  & [\p][^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_e b[\gamma][\p] \\
+  & \q[^+]' && : (e:B[\p][\gamma^+] = B[\gamma][\p])\ra \q[\gamma^+] =_e \q \\
+  & [\p][\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra b[\p ][\langle a\rangle] =_e b \\
+  & \q[\langle\rangle]' && : (e:B[\p][\langle a\rangle] = B)\ra \q[\langle a\rangle] =_e a \\
+  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
+  & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
+  & \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
+  & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
+  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
+  & {\cdot}[]' && : (e:B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
+  & && \hphantom{{}:{}} (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]) \\
+  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
+  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b \\
   & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
@@
 \end{alignat*}
-\begin{alignat*}{10}
-  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
-  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
-  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
-  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
-  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
-  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
-  & \Lift\beta && : \un\,(\mk\,a) = a \\
-  & \Lift\eta && : \mk\,(\un\,a) = a
-\end{alignat*}
 

### 2024-09-15 types_2024_postproc/p.tex
stuff

@@
 This concludes the first-order single substitution semantics of the
-SOGAT \ref{eq:tt}. The theory can be extended with new type and term
-formers in the same fashion: each rule has to be indexed over
+SOGAT \ref{eq:tt}. For reference, we list the same rules again in
+Appendix \ref{app:tt}. The theory can be extended with new type and
+term formers in the same fashion: each rule has to be indexed over
 arbitrary contexts and come equipped with a substitution rule. There
@@
 Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
-$\Pi\eta$ conditional. This needs another change: we exchange
+$\Pi\eta$ conditional. This needs another change: we substitute
 $\Pi\beta$ for a more general variant which is also conditional on the
@@
 \end{alignat*}
-Now we derive the assumption for any type $B : \Ty\,(\Gamma\ext A)$.
+Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
 \begin{alignat*}{10}
@@
 \end{alignat*}
-With this we derived an even more minimalistic first-order theory for
-the SOGAT (\ref{eq:tt}), essentially relying on the universe and $\Pi$
-types. It is clear from the construction that the resulting theory is
-isomorphic to the one presented in Section \ref{sec:tt} (and Appendix
-\ref{app:tt}), all equations are interderivable. For reference, we
-list all the rules of the minimised calculus in Appendix
-\ref{app:minimisation}.
+This concludes the derivation of an even more minimalistic first-order
+theory for the SOGAT (\ref{eq:tt}), relying essentially on the
+universe and $\Pi$ types. It is clear from the construction that the
+resulting theory is isomorphic to the one presented in Section
+\ref{sec:tt} (and Appendix \ref{app:tt}), all equations are
+interderivable. For reference, we list all the rules of the minimised
+calculus in Appendix \ref{app:minimisation}.
 
@@
 
-We list the rules of the theory derived in Section
-\ref{sec:minimisation}.
+Rules of the theory derived in Section \ref{sec:minimisation}
+(omitting $\Lift$):
 \begin{alignat*}{10}

### 2024-09-16 types_2024_postproc/p.tex
stuff

@@
 \newcommand{\Sub}{\mathsf{Sub}}
+\newcommand{\Tel}{\mathsf{Tel}}
+\newcommand{\Tms}{\mathsf{Tms}}
 \newcommand{\p}{\mathsf{p}}
@@
 
+\section{stuff}
+
+There are equations in B-systems that cannot be derived from the equations of
+SSC. The following four equations are such, they are versions of equations in
+section \ref{sec:ssc} lifted under arbitrary amount of $\blank^+$s.
+\begin{equation}\label{eq:lifted}
+\begin{aligned}
+  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
+  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
+  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
+  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
+\end{aligned}
+\end{equation}
+In addition, we need the corresponding lifted equations for terms. Note that we
+do not have the non-lifted term versions of the last two equations in SSC.
+
+To formally state the equations in (\ref{eq:lifted}), we define telescopes as an
+inductive type together with a recursive operation to append telescopes to
+contexts.
+\begin{alignat*}{10}
+  & \Tel && : \Con\ra\Set \\
+  & \blank+\blank && : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con \\
+  & \diamond && : \Tel\,\Gamma \\
+  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Omega
+\end{alignat*}
+The appending operation is defined recursively on telescopes.
+\begin{alignat*}{10}
+  & \Gamma+\diamond && := \Gamma \\
+  & \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
+\end{alignat*}
+We then define a lifting operation over any telescope mutually with an
+instantiation operation on telescopes by recursion on telescopes.
+\begin{alignat*}{10}
+  & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
+  & \blank^{+^\Omega} && (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
+\end{alignat*}
+
+If we try to prove the equations in (\ref{eq:lifted}) by induction on the types
+and terms, we run into difficulties in the instantiation case, as we would have
+to commute the instantiations without the equations to do so. Therefore, we
+first $\alpha$-normalize the syntax to compute away all the instantiations, then
+prove the equations by induction on the $\alpha$-normal forms.
+
+We define $\alpha$-normal forms as inductive predicates in Prop
+
+separate weakening and single substitution
+
+Define instantiation operations by recursion on $\alpha$-normal forms
+
+Instead of doing induction on the syntax for each of the equations in
+(\ref{eq:lifted}), we define a general lemma which can lift any equation between
+instantiations over a telescope.
+
+We define $\Sub^*$ to be $\Sub$ with freely added identity and composition
+operations. All instantiation operations, lifting operations, and substitution
+rules can be redefined for $\Sub^*$.
+
+\begin{lemma}
+  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
+  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
+  $x$ in $A$, then
+  \begin{itemize}
+    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
+    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
+    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  asd
+\end{proof}
+
+We can simulate parallel substitutions using $\Sub^*$ as iterated single
+substitutions, however it does not satisfy the equations of parallel
+substitutions. Thus we define parallel substitutions as a list of terms with a
+map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
+into a sequence of instantiations of single substitutions.
+\begin{alignat*}{10}
+  & \Tms && : \Con\ra\Con\ra\Con \\
+  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
+  & \upvarepsilon && : \Tms\,\Gamma\,\diamond \\
+  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
+  & \llcorner\upvarepsilon\lrcorner && := id\circ\dots\circ\p \\
+  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
+\end{alignat*}
+
 \section{Minimisation}

### 2024-09-16 types_2024_postproc/p.tex
stuff

@@
   \begin{alignedat}{10}
-    & \top && : \Ty\,i \\
+    & \top && : \Ty\,0 \\
     & \tt && : \Tm\,\top \\
@@
 
+\subsection{Examples}
+
 The polymorphic identity function for types at the level 0 is defined
@@
 $1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
-argument. Inrinsically typed terms are the same as derivation trees,
-we illustrate this by deriving this term as follows.
+argument, and these arguments themselves are terms written using
+implicit arguments. Inrinsically typed terms are the same as
+derivation trees, we illustrate this by deriving this term as follows.
 \[
-\infer[(*)]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
+\infer[\text{(*)}]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
 \]
@@
 
-TODO: beta equation for id.
+TODO: beta equation for id, show what we can do with types in an open
+context.
+
+\subsection{Models}
+
+TODO: To make sure that our definitions of models are correct, we show
+that in Agda there is a model for both the parallel and single
+substitution calculi.
 
@@
 
+TODO: summarise this section with a theorem?
+
 \section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
+
+We can emulate the parallel substitution calculus using function
+space, U and Sigma types
 
@@
 fewer equations, hence easier to define models, coherent syntax?
+
+We showed a particular example for the SOGAT->single GAT translation,
+we showed that the syntax is equivalent to the other one, we
+conjecture that the method presented here generalises. For this
+particular SOGAT, the single GAT syntax can be simplified, and in
+fact, it is ``equivalent'' to the parallel theory.
 
@@
 model. For example, this was the goal of the initiality project for
-homotopy type theory \cite{brunerie}. Maybe list work on type theory
+homotopy type theory \cite{initiality-agda}. Maybe list work on type theory
 using QIITs, see TYPES mailing list.

### 2024-09-17 types_2024_postproc/p.tex
stuff

@@
 
+We mainly work in ETT.
+
 \section{Single substitution syntax}
@@
 
-There are equations in B-systems that cannot be derived from the equations of
-SSC. The following four equations are such, they are versions of equations in
-section \ref{sec:ssc} lifted under arbitrary amount of $\blank^+$s.
+The following four equations correspond to equations in B-systems which cannot
+be derived from the equations of SSC, these are versions of equations in section
+\ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
 \begin{equation}\label{eq:lifted}
@@
 \end{equation}
-In addition, we need the corresponding lifted equations for terms. Note that we
-do not have the non-lifted term versions of the last two equations in SSC.
+The correspoinding equations on terms are also needed. Note that we do not have
+the non-lifted term versions of the last two equations in SSC.
 
@@
 
-We define $\alpha$-normal forms as inductive predicates in Prop
-
-separate weakening and single substitution
-
-Define instantiation operations by recursion on $\alpha$-normal forms
+A type or term is $\alpha$-normal if it does not contain instantiation
+operations, except at the leaves of the syntax tree as variables. However,
+$\alpha$-normal can still contain $\beta$/$\eta$ redexes, so that we do not need
+to do full normalization. We define variables and $\alpha$-normal forms as
+inductive predicates in Prop.
+\begin{itemize}
+  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
+  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
+  is a variable.
+  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
+  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
+  $\alpha$-normal types.
+  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
+  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
+  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
+\end{itemize}
+The predicate is defined similarly for the other type and term formers. Notably,
+we do not state that instantiated types/terms are $\alpha$-normal. It is
+important to truncate the $\alpha$-normal predicates to be propositions to
+ensure that it preserves the $\beta$/$\eta$ conversion rules.
+
+We define instantiation operations by recursion on $\alpha$-normal forms. The
+instantiation operations need to be defined separately for weakenings and single
+substitutions for the recursion to be well-founded.
+
+$x[\p][\gamma^+]' = b[\gamma]'[\p]'$
+
+\begin{itemize}
+  \item $\p$ is a weakening.
+  \item $\gamma^+$ is a weakening if $\gamma$ is a weakening.
+  \item $\langle a \rangle$ is an $\alpha$-normal single substitution if $a$ is an
+  $\alpha$-normal term.
+  \item $\gamma^+$ is an $\alpha$-normal single substitution if $\gamma$ is an
+  $\alpha$-normal single substitution.
+\end{itemize}
+
+\begin{lemma}
+  \label{thm:alpha}
+  Every type and term is/has? an $\alpha$-normal form.
+\end{lemma}
+\begin{proof}
+  By induction on the syntax, normalizing the substitutions at the same time.
+\end{proof}
 
@@
 (\ref{eq:lifted}), we define a general lemma which can lift any equation between
-instantiations over a telescope.
-
-We define $\Sub^*$ to be $\Sub$ with freely added identity and composition
-operations. All instantiation operations, lifting operations, and substitution
-rules can be redefined for $\Sub^*$.
-
+instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
+freely added identity and composition operations, we do not require it to
+satisfy the laws of a category. All instantiation operations, lifting
+operations, and substitution rules can be redefined for $\Sub^*$.
 \begin{lemma}
+  \label{thm:lifting}
   Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
   A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
-  $x$ in $A$, then
+  $x:\Tm\,\Gamma\,A$, then
   \begin{itemize}
@@
 \begin{proof}
-  asd
+  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
+  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
+  the variable. Then the two latter equations can be proven by induction on
+  $\alpha$-normalized types and terms, still assuming the first equation.
+  Finally we prove the first equation by induction on the telescope.
 \end{proof}
@@
   & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
-  & \upvarepsilon && : \Tms\,\Gamma\,\diamond \\
+  & \epsilon && : \Tms\,\Gamma\,\diamond \\
   & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
-  & \llcorner\upvarepsilon\lrcorner && := id\circ\dots\circ\p \\
+  & \llcorner\epsilon\lrcorner && := id\circ\p\circ\dots\circ\p \\
   & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
 \end{alignat*}
+The computation of instantiating with $\Tms$ is illustrated bellow.
+\begin{gather*}
+  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = \\
+  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
+\end{gather*}
 
@@
 
+contextual isomorphism \cite{DBLP:journals/corr/abs-2211-07487}
+
 \section{Conclusions}
@@
 using assumptions, and proving everything by induction.
+
+If we have var, then we cannot do termification to get a CwF from a var-SSC.
 

### 2024-09-17 types_2024_postproc/p.tex
stuff

--- 2024-09-16 types_2024_postproc/p.tex
@@
 \newcommand{\un}{\mathsf{un}}
+\newcommand{\id}{\mathsf{id}}
 
 \begin{document}
+
+\newtheorem{problem}[theorem]{Problem}
+\theoremstyle{remark}
+\newtheorem{construction}[theorem]{Construction}
 
@@
 \end{alignat*}
+\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
 This concludes the first-order single substitution semantics of the
-SOGAT \ref{eq:tt}. For reference, we list the same rules again in
-Appendix \ref{app:tt}. The theory can be extended with new type and
-term formers in the same fashion: each rule has to be indexed over
-arbitrary contexts and come equipped with a substitution rule. There
-is no need to add more structural (substitution calculus) rules.
+SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
+Appendix \ref{app:tt}.
+\end{definition}
+The theory can be extended with new type and term formers in the same
+fashion: each rule has to be indexed over arbitrary contexts and come
+equipped with a substitution rule. There is no need to add more
+structural (substitution calculus) rules. For example, we show how to
+extend Definition \ref{def:tt} with the SOGAT-rules (\ref{eq:sigma}).
+\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
+  We extend Definition \ref{def:tt} with the following rules.
+  \begin{alignat*}{10}
+    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
+    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
+    & \top\eta && : t = \tt \\
+    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
+    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
+    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
+    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
+    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
+  \end{alignat*}
+  To save space, we compressed the introduction, elimination, $\beta$
+  and $\eta$ rules for $\Sigma$ into one isomorphism, and did not list
+  substitution laws for $\fst$ and $\snd$ as they are derivable.
+\end{definition}
 
@@
 \end{alignat*}
-This concludes the derivation of an even more minimalistic first-order
-theory for the SOGAT (\ref{eq:tt}), relying essentially on the
-universe and $\Pi$ types. It is clear from the construction that the
-resulting theory is isomorphic to the one presented in Section
-\ref{sec:tt} (and Appendix \ref{app:tt}), all equations are
-interderivable. For reference, we list all the rules of the minimised
+We summarise this section by formally stating what we defined.
+
+\begin{definition}[Minimised single substitution calculus with $\Pi$ and $\U$]\label{def:min}
+  This section defined the minimised first-order version of the SOGAT
+  (\ref{eq:tt}), relying essentially on the presernce of universes and
+  $\Pi$ types. For reference, list all the rules of the minimised
   calculus in Appendix \ref{app:minimisation}.
-
-TODO: summarise this section with a theorem?
+\end{definition}
+\begin{theorem}
+  The GATs of Definition \ref{def:min} and Definition \ref{def:tt} are
+  isomorphic, in particular, all equations are interderivable.
+\end{theorem}
+\begin{proof}
+  Clear from the construction in this section.
+\end{proof}
 
@@
 
-We can emulate the parallel substitution calculus using function
-space, U and Sigma types
-
-termification in the shallow paper is shown equationally complete
-
-intuition, similar to minimisation: the subst calc is just there to
-bootstrap U and Pi, those can do everything.
+In this section we show that if the single substitution calculus has
+certain type formers, then it is actually weakly equivalent to CwFs
+with the same type formers. The idea is that
+\begin{itemize}
+\item using $\Sigma$ types we emulate contexts,
+\item using functions between these $\Sigma$ types we emulate parallel
+  substitutions,
+\item using the functions into the universe we emulate dependent types,
+\end{itemize}
+and then single substitutions are just there to set up $\Sigma$, $\Pi$
+and $\U$, and these types are enough to bootstrap the parallel
+substitution calculus.
+
+\begin{definition}[CwF]
+  A category with $\N$-many families is defined as a category (objects
+  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
+  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
+  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
+  action on morphisms $\blank[\blank]$); for each $i$ a locally
+  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
+  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
+  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
+  with an isomorphism
+  $
+  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
+  $
+  natural in $\Delta$.
+\end{definition}
+\begin{problem}
+  Every CwF (with type formers $\Pi, \dots, \Sigma$) is an SSC (with
+  the same type formers, see Definition \ref{def:sigma}).
+\end{problem}
+\begin{proof}[Construction]
+  Most operations are the same, we set $\gamma^+ :=
+  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$.
+\end{proof}
+The following construction is also known as the standard model,
+metacircular interpretation \cite{DBLP:conf/popl/AltenkirchK16},
+contextualisation \cite{DBLP:conf/fscd/BocquetKS23}. Following
+\cite{kaposi-shallow} we call it termification, as most sorts in the
+new model are terms in the old model.
+\begin{problem}[Termification]
+  From a model of SSC (with $\Pi, \dots, \Sigma$, see Definition
+  \ref{def:sigma}), we define a CwF (with the same type formers).
+\end{problem}
+\begin{proof}[Construction]
+  We define iterated lifting $\Lift^k :
+  \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(k+i)$ by induction on $k$, together
+  with $\un^k : \Tm\,(\Lift^k\,A)\cong\Tm\,A : \mk^k$. The category
+  part of the CwF is given by types in the empty context and functions
+  between them, suitably lifted (on the left hand side of $:=$ there
+  is the component in the new CwF-model, on the right hand side the
+  components refer to the old SSC-model):
+  \begin{alignat*}{10}
+    & \Con && := (i:\N)\times\Ty\,\diamond\,i \\
+    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Lift^{\Gamma-\Delta}\,\Delta\Ra\Lift^{\Delta-\Gamma}\,\Gamma)
+  \end{alignat*}
+  In the definition of $\Sub$, we did not write the projections for
+  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$
+  depending on the context, and we used the truncating subtraction of
+  natural numbers. Composition of substitutions is quite involved, but
+  it is just function composition written with explicit weakenings and
+  appropriate (un)lifting. The category laws hold. The empty context is
+  modelled by $\top$, its $\eta$ law holds 
+  \begin{alignat*}{10}
+    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\
+    & && \hspace{10em}(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
+    & \rlap{$\id := \lam\,\q \hspace{4.35em} \diamond := (0,\top) \hspace{4.35em} \epsilon := \lam\,(\mk^\Gamma\,\tt)$}
+  \end{alignat*}
+  Types are given by functions into $\U$, terms are dependent
+  functions into the type, with lots of lifting adjustments.
+  \begin{alignat*}{10}
+    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)\,(\U\,i)}) \\
+    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\
+    & && \hspace{4em} \big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
+  \end{alignat*}
+  To make the notation readable, we will not write the lifting
+  decorations from now on, and we repeat the previous definitions
+  without mentioning universe levels, but all of what follows is
+  definable properly with lifting.
+  \begin{alignat*}{10}
+    & \Con && := \Ty\,\diamond \\
+    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma) \\
+    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \\
+    & \id && := \lam\,\q \\
+    & \diamond && := \top \\
+    & \epsilon && := \lam\,\tt \\
+    & \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U) \\
+    & A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) \\
+    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \\
+    & a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\
+    & \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\
+    & (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q) \\
+    & \p && := \lam\,(\fst\,\q) \\
+    & \q && := \lam\,(\snd\,\q)
+  \end{alignat*}
+  Context extension is given by $\Sigma$ types and pairing/projections
+  by pairing/projections of $\Sigma$. All the CwF equations hold, for
+  example we derive the functor law for type substitution as follows.
+  \begin{alignat*}{10}
+    & A[\gamma\circ\delta] && {=} \\
+    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q))[\p]\cdot\q)) && {=}(\lam[]) \\
+    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta) \\
+    & \lam\,(A[\p]\cdot(\gamma[\p]\cdot(\delta[\p]\cdot\q))) && {=}(\Pi\beta) \\
+    & \lam\,(\lam\,(A[\p][\p^+]\cdot(\gamma[\p][\p^+]\cdot\q))\cdot(\delta[\p]\cdot\q))\,\, && {=}(\lam[]) \\
+    & \lam\,(\lam\,(A[\p]\cdot(\gamma[\p]\cdot\q))[\p]\cdot(\delta[\p]\cdot\q)) && {=} \\
+    & A[\gamma][\delta]
+  \end{alignat*}
+  Type formers are added by adjusting them to handle contexts built up
+  by $\Sigma$ types, for example,
+  \[
+  \Pi\,A\,B := \lam\,(\cd\,(\Pi\,(\El\,(A[\p]\cdot\q))\,(\El\,(B[\p][\p]\cdot(\q[\p],\q)))))
+  \]
+  and $\lam\,b := \lam\,(\lam\,(b[\p][\p]\cdot(\q[\p],\q)))$. All
+  substitution laws and $\beta$/$\eta$-laws hold.
+\end{proof}
+The notion of equivalence between CwFs that we use is contextual
+isomorphism \cite{DBLP:journals/corr/abs-2211-07487}: a weak
+CwF-morphism (pseudomorphism) which is bijective on types and
+terms. Note that a conextual isomorphism preserves all type formers
+specified by universal properties.
+\begin{problem}
+  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
+  obtained by first seeing it as an SSC and then termifying it. There
+  is a contextual isomorphism between $M'$ and $M$.
+\end{problem}
+\begin{proof}[Construction]
+  We denote the components of the contextual isomorphism $F$ as follows.
+  \begin{alignat*}{10}
+    & F : \Con_{M'}\ra\Con_M \\
+    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma) \\
+    & F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
+    & F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
+  \end{alignat*}
+  We define them in the same order as follows, omitting the $M$ subscripts.
+  \begin{alignat*}{10}
+    & F : \Ty\,\diamond\ra\Con \\
+    & F\,\Gamma := \diamond\ext\Gamma \\
+    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) \\
+    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \\
+    & F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
+    & F\,A := \El\,(A[\p]\cdot\q) \\
+    & F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
+    & F\,a := a[\p]\cdot\q
+  \end{alignat*}
+  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
+  is functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
+  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
+  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
+  A))\,(F\,\Gamma\ext F\,A)$ is an isomorphism.
+\end{proof}
+Starting with an SSC-model, termifying it and comparing the result
+with the original SSC does not provide an isomorphism because we can't
+define the map $F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
+\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on substitutions is
+not definable using single substitutions. We leave formulating the
+right notion of contextual isomorphism for SSCs as future work.
 
@@
 
+If we have var, then we cannot do termification to get a CwF from a var-SSC.
+
+Prove full equivalence of CwF+ with SSC+.
+
 \bibliography{b}{}
@@
 
-For reference, we list the rules of the theory introduced in Section
-\ref{sec:tt}. The core substitution calculus:
+Definition \ref{def:tt} in full detail: \\
+The core substitution calculus:
 \begin{alignat*}{10}
@@
 
-Rules of the theory derived in Section \ref{sec:minimisation}
-(omitting $\Lift$):
+Definition \ref{def:min} in full detail (omitting $\Lift$):
 \begin{alignat*}{10}

### 2024-09-17 types_2024_postproc/p.tex
stuff
structure

@@
 formal system based on Martin-Löf's original definition
-\cite{martinlof73predicative}, and not to the field studying type
-systems (e.g.\ \cite{DBLP:books/daglib/0005958}).
+\cite{martinlof73predicative}, and not to the study of type systems
+(e.g.\ \cite{DBLP:books/daglib/0005958}).
 
@@
 we have to index types by their (universe) level, and types at lower
-level are included at all upper levels by $\Lift$. We use implicit
+levels are included at upper levels by $\Lift$. We use implicit
 arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
@@
 \cite{DBLP:conf/lics/Hofmann99} or logical framework
-\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
-is no good notion of homomorphism between second-order models, thus
-the semantics of a SOGAT is given by a first order generalised
-algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
-generalisations of single-sorted algebraic theories (such as monoids,
-groups) to multiple sorts where later sorts can be indexed over
-previous sorts. The usual results from universal algebra (e.g.\ models
-form a category with finite limits and initial object, existence of
-free and co-free models) carry over to GATs
-\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
-theory of categories where there is a sort for objects and a sort for
-morphisms double-indexed over objects.
+\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} and
+our notation comes from two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
+second-order models, thus the semantics of a SOGAT is given by a first
+order generalised algebraic theory (GAT
+\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
+single-sorted algebraic theories (such as monoids, groups) to multiple
+sorts where later sorts can be indexed over previous sorts. The usual
+results from universal algebra (e.g.\ models form a category with
+finite limits and initial object, existence of free and co-free
+models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
+example of a GAT is the theory of categories where there is a sort of
+objects and a sort of morphisms double-indexed over objects.
 
@@
 is the GAT of categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}). We define the notion
-of model of a language (a SOGAT) by its parallel substitution
-semantics, which is a GAT, and that induces a notion of model. The
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
+We define the notion ``model of the SOGAT'' by its parallel substitution
+semantics, which is a GAT, and that has an obvious notion of model. The
 syntax of the language is the initial model, which exists for every
@@
 B-systems. GAT presentations of type theory are important for defining
-models, our presentation gives a new notion of model, which in the
+models. Our presentation gives a new notion of model, which in the
 future might contribute to definitions of new models. It can also
@@
 
-The single substitution semantics of (\ref{eq:tytm}) is in some sense
+The single substitution semantics is in some sense
 too minimalistic: for example, consider a term $b$ which depends on a
@@
 denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
-We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary
-model, but this equation holds in the syntax.
+We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary single substitution
+model, but this equation holds in the single substitution syntax.
 % p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
@@
 parallel substitution syntax of (\ref{eq:tt}), and the techniques that
-we use here are not specific to the SOGAT (\ref{eq:tt}), we expect
+we use are not specific to the SOGAT (\ref{eq:tt}), we expect
 that they work for any SOGAT. Equivalence of the syntaxes means that
@@
 \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
-substitution models of (\ref{eq:tt}) than parallel substitution
+substitution models of (\ref{eq:tytm}) than parallel substitution
 models. The situation is similar to that of extensional combinatory
@@
 \Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
-terms. When doing induction on types/terms, we always have to provide
+terms. When doing induction on types/terms, we have to provide
 a method for these operations, but this can be difficult as
@@
 $\beta$-normal forms don't distinguish between $\beta$-equal terms,
-$\alpha$-normal forms don't include explicit substitutions, all the
-variables are instantiated. $\alpha$-normal forms are not usual normal
+$\alpha$-normal forms don't include explicit substitutions.
+$\alpha$-normal forms are not usual normal
 forms in the sense that they are still quotiented by the
@@
 In the particular theory (\ref{eq:tt}), every type is represented by a
-term. In the first order single substitution calculus, this can be
+term. In its single substitution calculus, this can be
 used to adapt some equations and remove some others, obtaining a more
-minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
+minimalistic, but equivalent GAT.
+
+Theory (\ref{eq:tt}) can be extended
 with unit and $\Sigma$ types, more precisely with the following rules.
@@
 collections of dependencies can be collected using $\Sigma$ types, we
-can actually derive a parallel model from a single substitution model.
-TODO: universe levels are a problem (maybe not if we use iterated
-function space instead of $\Sigma$); the resulting CwF is not the same as
-the free CwF over the SSC. Bootstrapping a CwF.
-CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
-SSC -> CwF -> SSC = we get a stronger thing than what we started with.
+can actually derive a parallel model from any single substitution
+model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
+single and parallel substitution models are actually equivalent in a
+weak sense.
 
@@
   substitutions from the single substitution syntax.
-\item Minimisation.
-\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
-  a parallel model from a single substitution model.
+\item For type theory with $\Pi$, $\U$, a minimised presentation of
+  the equations which results in an isomorphic theory.
+\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
+  single and parallel models are equivalent in a weak sense.
 \end{itemize}
@@
 
-TODO. In Section \ref{sec:tt}, we introduce the single substitution
-calculus only relying on a working knowledge of an implementation of
-type theory such as Agda, no prior experience with the metatheory of
-type theory is required. We first show how to get rid of explicit
-substitutions from the syntax using the technique of
-$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
-prove several admissible equations. Then we show how to derive the
-rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
-show that our syntax is isomorphic to the CwF-style syntax (Section
-\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.
+In Section \ref{sec:tt}, we introduce the single substitution calculus
+(SSC) only relying on a working knowledge of an implementation of type
+theory such as Agda, no prior experience with the metatheory of type
+theory is required. Section \ref{sec:admissible} shows that in the
+syntax of SSC, several new equations are admissible, and that the
+syntax of SSC is actually isomorphic to the CwF-based syntax. In
+Section \ref{sec:minimisation} we minimise our SSC obtaining a theory
+with fewer equations, relying on the presence of $\Pi$ and $\U$ in our
+theory. In Section \ref{sec:cwf}, we show that in the presence of
+$\Pi$, $\U$ and $\Sigma$, SSC-models are actually equivalent to
+CwF-models in a weak sense. We conclude in Section
+\ref{sec:conclusion}.
 
@@
 Our metatheory in which this paper is
-written is thus type theory, and everything is formalised in Agda.
+written is thus type theory, some sections are formalised in Agda.
 
@@
 \section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
+\label{sec:cwf}
 
@@
 
-\begin{definition}[CwF]
+\begin{definition}[CwF]\label{def:cwf}
   A category with $\N$-many families is defined as a category (objects

### 2024-09-17 types_2024_postproc/p.tex
meta

@@
 
-Our metatheory in which this paper is
-written is thus type theory, some sections are formalised in Agda.
-
-As Agda does not support them directly, we postulate their
-constructors and induction principle and add the computation rules
-using rewrite rules \cite{DBLP:conf/types/Cockx19}.
-
-Notation and formalisation techniques, including congruence rules.
-
-Prop.
-
-We mainly work in ETT.
+Our metatheory is observational type theory
+\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
+(QIITs). On paper we usually omit writing coercions, so our notation
+is close to extensional type theory. Sections \ref{sec:tt} and
+\ref{sec:admissible} of this paper are formalised in the proof
+assistant Agda, we use a strict Prop-valued
+\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
+coercion rule (transport rule) and postulated QIITs with computation
+rules added using rewrite rules
+\cite{DBLP:journals/pacmpl/CockxTW21}. Our notation on paper is close
+to Agda's, we write $\Pi$ types as $(x:A)\ra B$, $\Sigma$ types as
+$(x:A)\times B$, and use implicit arguments and overloading
+extensively.
 

### 2024-09-17 types_2024_postproc/p.tex
stuff

@@
 \newcommand{\Set}{\mathsf{Set}}
+\newcommand{\Prop}{\mathsf{Prop}}
 \newcommand{\Ty}{\mathsf{Ty}}
@@
   & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
-  & \blank^{+^\Omega} && (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
+  & \blank^{+^\Omega} && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
 \end{alignat*}
@@
 operations, except at the leaves of the syntax tree as variables. However,
-$\alpha$-normal can still contain $\beta$/$\eta$ redexes, so that we do not need
-to do full normalization. We define variables and $\alpha$-normal forms as
-inductive predicates in Prop.
+$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
+not need to do full normalization. We define variables and $\alpha$-normal forms
+as inductive predicates in $\Prop$.
 \begin{itemize}
@@
 
-We define instantiation operations by recursion on $\alpha$-normal forms. The
-instantiation operations need to be defined separately for weakenings and single
-substitutions for the recursion to be well-founded.
-
-$x[\p][\gamma^+]' = b[\gamma]'[\p]'$
-
+We prove that the $\alpha$-normal predicate holds for $\alpha$-normal types and
+terms instantiated with $\alpha$-normal substitutions. However, this needs to be
+proved separately for weakenings and $\alpha$-normal single substitutions for
+the induction to be well-founded. We define predicates for weakening and
+$\alpha$-normal single substitutions as follows.
 \begin{itemize}
@@
   \label{thm:alpha}
-  Every type and term is/has? an $\alpha$-normal form.
+  The $\alpha$-normal predicate holds for any type and term.
 \end{lemma}
 \begin{proof}
-  By induction on the syntax, normalizing the substitutions at the same time.
+  By induction on the syntax, $\alpha$-normalizing the substitutions at the same
+  time.
 \end{proof}
 
-Instead of doing induction on the syntax for each of the equations in
+Instead of doing induction on $\alpha$-normal forms for each of the equations in
 (\ref{eq:lifted}), we define a general lemma which can lift any equation between
@@
 freely added identity and composition operations, we do not require it to
-satisfy the laws of a category. All instantiation operations, lifting
-operations, and substitution rules can be redefined for $\Sub^*$.
+satisfy the laws of a category as we will not compare $\Sub^*$s directly for
+equality, only between types and terms instantiated with $\Sub^*$s. All
+instantiation operations, lifting operations, and substitution rules can be
+redefined for $\Sub^*$.
 \begin{lemma}
@@
   \end{itemize}
+  Note that the later equations depend the the earlier ones.
 \end{lemma}
@@
   x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
-  the variable. Then the two latter equations can be proven by induction on
-  $\alpha$-normalized types and terms, still assuming the first equation.
-  Finally we prove the first equation by induction on the telescope.
+  the variable. Then the two latter equations can be proven by mutual induction
+  on $\alpha$-normalized types and terms, still assuming the first equation.
+  Finally we prove the first equation by induction on the telescope, using the
+  previously proven equation for types, discharging its assumption.
 \end{proof}
@@
   & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
-  & \llcorner\epsilon\lrcorner && := id\circ\p\circ\dots\circ\p \\
+  & \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
   & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
 \end{alignat*}
-The computation of instantiating with $\Tms$ is illustrated bellow.
+The computation of instantiating with $\Tms$ is illustrated below.
 \begin{gather*}
@@
 \end{gather*}
+All CwF operations and equations can be defined with $\Tms$ by induction.
+
+\begin{theorem}
+  Contexts, types, and terms in the SSC syntax are isomorphic to the
+  corresponding sorts in the CwF syntax.
+\end{theorem}
+\begin{proof}\leavevmode
+  \begin{description}
+    \item[$\Rightarrow$] By recursion on the syntax, SSC operations can be
+    trivially interpreted by CwF operations.
+    \item[$\Leftarrow$] By recursion on the syntax, using $\Tms$ to interpret
+    parallel substitutions.
+  \end{description}
+  The roundtrips are proven by induction on the syntax.
+\end{proof}
 

### 2024-09-17 types_2024_postproc/p.tex
stuff

--- 2024-09-17 types_2024_postproc/p.tex
@@
 
-\subsection{Models}
-
-TODO: To make sure that our definitions of models are correct, we show
-that in Agda there is a model for both the parallel and single
-substitution calculi.
+As a sanity check for our notion of SSC-model, we defined the standard
+(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
+SSC-based calculus calculus in Agda. We used the techniques of Kovács
+\cite{DBLP:conf/csl/Kovacs22} to define inductive-recursive universes
+to model the infinite hierarchy of types, and all the SSC-equations
+hold by reflexivity in this model. We defined the analogous CwF-based
+standard model.
 
@@
   appropriate (un)lifting. The category laws hold. The empty context is
-  modelled by $\top$, its $\eta$ law holds 
+  modelled by $\top$, its $\eta$ law holds via $\eta$ for $\top$.
   \begin{alignat*}{10}
@@
 
-\section{Conclusions}
+\section{Conclusions and further work}
 \label{sec:conclusion}

### 2024-09-17 types_2024_postproc/p.tex
stuff

@@
 
-TODO: beta equation for id, show what we can do with types in an open
-context.
+Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
+\begin{alignat*}{10}
+  & \lam\,(\lam\,\q)\cdot\cd\,A\cdot a && {=}(\Pi\beta) \\
+  & (\lam\,\q[\langle\cd\,A\rangle])\cdot a && {=}(\lam[]) \\
+  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a && {=}([\q][^+]) \\
+  & (\lam\,\q)\cdot a && {=}(\Pi\beta) \\
+  & \q[\langle a\rangle] && {=}(\q[\langle\rangle]) \\
+  & a
+\end{alignat*}
+If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
+$a' : \Tm\,(\diamond\ext C)\,A'$, we have to weaken our function to
+accept them:
+\begin{alignat*}{10}
+  & \lam\,(\lam\,\q)[\p] : \Tm\,&& (\diamond\ext C)\,\\
+  & && \Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big)
+\end{alignat*}
+But we have
+\[
+\lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
+\]
+(with different implicit arguments), so we are able apply $\cd\,A'$
+and $a'$ just as before. What we cannot do is to directly weaken a
+term in context $\diamond$ to an arbitrary metavariable context
+$\Gamma$, we can only do this step by step by applying weakening
+$\blank[\p]$ for each type in the context.
 

### 2024-09-17 types_2024_postproc/p.tex
stuff

@@
 relates to CwF as extensional combinatory calculus relates to lambda
-calculus. All the results in this paper were formalised in Agda.
+calculus. In case of additional type formers, we show that SSC models
+are actually equivalent to CwF-based models in a weak sense. Most
+results in this paper were formalised in Agda.
 \end{abstract}
@@
 
-TODO: say that we assume the QIIT version of this.
-
-$\alpha$-normal forms could also be called substitution normal forms.
-
-In this section we show that a category with families (CwF) equipped
-with a hierarchy of Coquand-universes and $\Pi$ types can be derived
-from the single substitution syntax defined in \autoref{sec:tt}.
-
-TODO: show that the SSC operations are derivable from CwF.
-
-We conjecture that the same works for any SOGAT.
-
-\section{stuff}
+In this section we show that our SSC-based syntax is isomorphic to the
+CwF-based syntax for the same theory (namely the SOGAT
+(\ref{eq:tt})). We conjecture that the method presented here works
+generically for any SOGAT.
+
+\begin{definition}[CwF]\label{def:cwf}
+  A category with $\N$-many families is defined as a category (objects
+  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
+  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
+  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
+  action on morphisms $\blank[\blank]$); for each $i$ a locally
+  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
+  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
+  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
+  with an isomorphism
+  $
+  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
+  $
+  natural in $\Delta$).
+\end{definition}
 
 The following four equations correspond to equations in B-systems which cannot
-be derived from the equations of SSC, these are versions of equations in section
+be derived from the equations of SSC, these are versions of equations in Section
 \ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
@@
 \end{equation}
-The correspoinding equations on terms are also needed. Note that we do not have
+The corresponding equations on terms are also needed. Note that we do not have
 the non-lifted term versions of the last two equations in SSC.
@@
 not need to do full normalization. We define variables and $\alpha$-normal forms
-as inductive predicates in $\Prop$.
+as inductive predicates in $\Prop$ as follows.
 \begin{itemize}
@@
 The predicate is defined similarly for the other type and term formers. Notably,
-we do not state that instantiated types/terms are $\alpha$-normal. It is
+we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
 important to truncate the $\alpha$-normal predicates to be propositions to
@@
 \begin{proof}
-  By induction on the syntax, $\alpha$-normalizing the substitutions at the same
-  time.
+  By induction on the syntax, $\alpha$-normalizing the substitutions
+  at the same time. Note that induction on the syntax refers to the
+  elimination principle of the corresponding QIIT.
 \end{proof}
@@
   \label{thm:lifting}
-  Given and $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
+  Given $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
   A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
@@
 substitutions. Thus we define parallel substitutions as a list of terms with a
-map into $\Sub^*$, and reuse instantiation operation of $\Sub^*$ to convert it
+map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to convert it
 into a sequence of instantiations of single substitutions.
@@
   & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
-  & \El\,((\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
-  & \El\,((\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
+  & \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
+  & \El\,(\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
   & \El\,(\cd\,B) && {=}(\U\beta) \\
@@
 
-\begin{definition}[CwF]\label{def:cwf}
-  A category with $\N$-many families is defined as a category (objects
-  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
-  ($\epsilon$ denotes the unique morphism into it); for each $i$ a
-  presheaf of types (action on objects denoted $\Ty\,\blank\,i$,
-  action on morphisms $\blank[\blank]$); for each $i$ a locally
-  representable dependent presheaf $\Tm$ over $\Ty\,\blank\,i$
-  (actions denoted $\Tm$, $\blank[\blank]$, local representability is
-  denoted $\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con$
-  with an isomorphism
-  $
-  (\p\circ\blank,\q[\blank]) : \Sub\,\Delta\,(\Gamma\ext A) \cong (\gamma:\Sub\,\Delta\,\Gamma)\times\Tm\,\Delta\,(A[\gamma]) : (\blank,\blank)
-  $
-  natural in $\Delta$.
-\end{definition}
 \begin{problem}
@@
   example we derive the functor law for type substitution as follows.
-  \begin{alignat*}{10}
+  \begin{equation}
+  \begin{alignedat}{10}\label{eq:functor}
     & A[\gamma\circ\delta] && {=} \\
     & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q))[\p]\cdot\q)) && {=}(\lam[]) \\
-    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta) \\
+    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta, {\cdot}[]) \\
+    & \lam\,(A[\p]\cdot(\gamma[\p][\p^+][\langle\q\rangle]\cdot(\delta[\p][\p^+][\langle\q\rangle]\cdot\q))) && {=}([\p][^+]) \\
+    & \lam\,(A[\p]\cdot(\gamma[\p][\p][\langle\q\rangle]\cdot(\delta[\p][\p][\langle\q\rangle]\cdot\q))) && {=}([\p][\langle\rangle]) \\
     & \lam\,(A[\p]\cdot(\gamma[\p]\cdot(\delta[\p]\cdot\q))) && {=}(\Pi\beta) \\
@@
     & A[\gamma][\delta]
-  \end{alignat*}
+  \end{alignedat}
+  \end{equation}
   Type formers are added by adjusting them to handle contexts built up
@@
 
-In this paper we introduced a new single substitution calculus which
-can be seen as a minimalistic variant of B-systems. We illustrate its
-usage through the definition of a type theory with dependent function
-space and a hierarchy of universes. We aim for accessibility: this
-paper can be seen as a tutorial introduction to the syntax of type
-theory inside type theory. All the operations and equations are
-well-motivated and provide a minimalistic description of the syntax of
-type theory. It is minimalistic in the sense that the syntax is
-isomorphic to the CwF-syntax, but there are more models.
-
-The relationship of our single substitution calculus and the CwF-based
-theory is similar to that of combinatory logic and lambda calculus
-\cite{DBLP:conf/fscd/AltenkirchKSV23}. Combinatory logic has more
-models than lambda calculus, but the syntaxes are
-isomorphic. Similarly, the single substitution calculus has more
-models, but the syntaxes are isomorphic.
-
-A more abstract, higher level description of a language is simply
-stating that a language is an algebraic theory: the terms of the
-language are given by a sort of the theory, the different term formers
-are operations, and the conversion relation is specified by the
-equations. Such a description contains less information than the
-traditional one, for example, conversion is not oriented, so it is not
-clear how to turn it into a rewriting relation; we cannot choose how
-much annotations we add to terms, they are always fully
-annotated. this is similar to how asts contain less information than
-strings, for example, the bracket-removal strategy is not encoded in
-asts. the algebraic description automatically induces a notion of
-model and the existence of an initial model
-\cite{DBLP:journals/pacmpl/KaposiKA19} which we call the syntax. The
-algebraic approach is especially natural for languages with dependent
-types where typing and conversion have to be specified mutually
-\cite{DBLP:conf/popl/AltenkirchK16}.
-
-QIIT is nice because syntax and semantics are given together. The
-notion of model/semantics is traditionally given separately, and work
-has to be done to show that the above specified syntax generates a
-model. For example, this was the goal of the initiality project for
-homotopy type theory \cite{initiality-agda}. Maybe list work on type theory
-using QIITs, see TYPES mailing list.
-
-We can abstract over these strategies and define our language as a
-second-order generalised algebraic theory
-\cite{DBLP:journals/corr/abs-1904-04097}, but to state initiality of
-our syntax, at some point we need to turn it into a first-order
-algebraic theory \cite{DBLP:conf/fscd/BocquetKS23}.
-
-We conjecture that any SOGAT can be turned into an SSC.
+Can we remove even more structural rules? the termification doesn't
+work if we don't have e.g. $[\p][^+]$ and $[\p][\langle\rangle]$ for all terms, and
 
 Separate sort for variables, without the extra type equations, by
-using assumptions, and proving everything by induction.
-
-If we have var, then we cannot do termification to get a CwF from a var-SSC.
+using assumptions, and proving everything by induction. If we have
+var, then we cannot do termification to get a CwF from a var-SSC, see
+the proof of equation (\ref{eq:functor}).
 
@@
 \bibliographystyle{plain}
+
+\pagebreak
+
+\phantom{A}
 

### 2024-09-18 types_2024_postproc/p.tex
finished

@@
 \newcommand{\id}{\mathsf{id}}
+\newcommand{\Type}{\mathsf{Type}}
 
@@
 rules added using rewrite rules
-\cite{DBLP:journals/pacmpl/CockxTW21}. Our notation on paper is close
-to Agda's, we write $\Pi$ types as $(x:A)\ra B$, $\Sigma$ types as
-$(x:A)\times B$, and use implicit arguments and overloading
-extensively.
+\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
+uniqueness of identity proofs (UIP), and our metatheory is
+incompatible with homotopy type theory \cite{HoTTbook}. Our notation
+on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
+$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
+overloading extensively.
 
@@
 but $B$ cannot be weakened by $\p$ because there we need $\p' :
-\Sub\,(\Gamma\ext C\ext A[\p])\,(\Gamma\ext A)$. So $p'$ has to be a
+\Sub\,(\Gamma\ext C\ext A[\p])$ $(\Gamma\ext A)$. So $p'$ has to be a % hack
 new kind of weakening which adds a new variable in the last but one
@@
 $\blank[\p]$ for each type in the context.
+
+\subsection{Standard model}
+\label{sec:standard}
 
@@
   \begin{alignat*}{10}
-    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\
+    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\[-1.5em]
     & && \hspace{10em}(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
@@
     & \Ty\,\Gamma\,i && := \Tm\,\diamond\,(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)\,(\U\,i)}) \\
-    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\
+    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\[-1em]
     & && \hspace{4em} \big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
@@
   F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
-  A))\,(F\,\Gamma\ext F\,A)$ is an isomorphism.
+  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack
 \end{proof}
-Starting with an SSC-model, termifying it and comparing the result
-with the original SSC does not provide an isomorphism because we can't
-define the map $F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
-\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on substitutions is
-not definable using single substitutions. We leave formulating the
-right notion of contextual isomorphism for SSCs as future work.
+The other roundtrip, that is, starting with an SSC-model, termifying
+it and comparing the result with the original SSC does not provide an
+isomorphism because we can't define the map $F :
+\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
+\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
+substitutions. Such a map is not definable only using single
+substitutions. However, the set of terms and types are still
+isomorphic after the roundtrip. We leave formulating the right notion
+of contextual isomorphism for SSCs as future work.
 
@@
 Why are we still looking at alternative ways of defining type theory?
-Closer to the syntax/implementation, easier to understand/implement,
-fewer equations, hence easier to define models, coherent syntax?
-
-We showed a particular example for the SOGAT->single GAT translation,
-we showed that the syntax is equivalent to the other one, we
-conjecture that the method presented here generalises. For this
-particular SOGAT, the single GAT syntax can be simplified, and in
-fact, it is ``equivalent'' to the parallel theory.
-
-Can we remove even more structural rules? the termification doesn't
-work if we don't have e.g. $[\p][^+]$ and $[\p][\langle\rangle]$ for all terms, and
-
-Separate sort for variables, without the extra type equations, by
-using assumptions, and proving everything by induction. If we have
-var, then we cannot do termification to get a CwF from a var-SSC, see
-the proof of equation (\ref{eq:functor}).
-
-Prove full equivalence of CwF+ with SSC+.
+New definitions might be closer to actual implementations, suggest new
+implementation techniques, they might be better pedagogically, have
+more models, or contribute to a coherent definition. We elaborate on
+the latter.
+
+Defining a useable syntax of type theory internally to homotopy type
+theory (HoTT \cite{HoTTbook}) is an open problem: the (h)set-truncated
+syntax can be formalised as a QIIT, but then we cannot even define the
+$\Type$-interpretation, that is, a function from the syntax which
+interprets closed types by the metatheoretic universe $\Type$ (rather
+than an inductive recursive universe as in Subsection
+\ref{sec:standard} or \cite[Section
+  4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
+syntax of type theory can be defined using infinitely many coherene
+rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
+$\Sub$, we have the associativity of composition of substitutions
+equation. If we truncate our syntax to be a set, then this is
+enough. If we don't truncate, there are two different ways of proving
+$((\gamma\circ\delta)\circ\theta)\circ\xi =
+\gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
+be the same, so in such a (wild) syntax substitutions don't form a
+set, hence they don't have decidable equality \cite{hedberg}. The
+solution is to add the pentagon law, but then this introduces new
+higher equalities, which require more coherence laws, so in the end we
+have an infinite sequence of higher coherence laws that need to be
+part of the syntax. Our single substitution calculus does not have
+composition for $\Sub$, so there is no need for higher coherence laws
+like associativity. However as it currently stands, our syntax is not
+coherent, it needs set-truncation when working in HoTT. For example,
+there are two different ways to prove the following equation:
+\begin{alignat*}{10}
+  & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
+  & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
+\end{alignat*}
+In a coherent version of our syntax we either need to state that these
+two proofs are equal, or we have to remove one of them. We can remove
+the first one by limiting $[\p][\langle\rangle]$ to only work on
+variables (this needs the introduction of a new sort for
+variables). If we do this, the minimisation of Section
+\ref{sec:minimisation} does not work anymore, so we need the extra
+four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
+$[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
+types make the syntax non-coherent again. A possible way out of this
+would be to make the equations that depend on one of these for
+equations conditional. But then the methods of Section
+\ref{sec:admissible} don't work anymore for re-proving the needed
+admissible equations. Also, if only have the rule $[\p][^+]$ only for
+variables, then we cannot redo the termification of Section
+\ref{sec:cwf}, in particular it is not clear how to prove equation
+(\ref{eq:functor}).
+
+In this paper we unfolded a particular example of the SOGAT $\ra$
+single substitution GAT translation
+\cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
+syntax is isomorphic to the parallel syntax, and we conjecture that
+our method works for any SOGAT. For our particular SOGAT, the single
+substitution syntax can be simplified, and we even have that any
+single substitution model is equivalent to a parallel model in a weak
+sense. This shows that if we have enough type formers, the
+substitution calculus can be very weak, and still enough to bootstrap
+the usual theory. It needs further investigation whether we can remove
+even more structural rules, and still obtain isomorphic syntaxes. We
+would also like to understand what a good notion of weak equivalence
+is for single substitution models.
 

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 relates to CwF as extensional combinatory calculus relates to lambda
-calculus. In case of additional type formers, we show that SSC models
+calculus. If we have some additional type formers, we show that SSC models
 are actually equivalent to CwF-based models in a weak sense. Most
@@
 \end{equation}
-Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
+Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
 we have to index types by their (universe) level, and types at lower
@@
 and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
-second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
-type $A$ in its second $\Ty$-argument. Every binder is represented by
+second-order functions. We say that $\Pi$ binds a $\Tm$-variable
+in its second $\Ty$-argument. Every binder is represented by
 a second-order function, and as a result, we don't need to talk about
@@
 theory (this technique is also known as higher-order abstract syntax
-\cite{DBLP:conf/lics/Hofmann99} or logical framework
-\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} and
-our notation comes from two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
+\cite{DBLP:conf/lics/Hofmann99}, logical framework
+\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
+two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
 second-order models, thus the semantics of a SOGAT is given by a first
@@
 rise to a sort/operation in the GAT which is now indexed by
-contexts. The sorts moreoever come with instantiation (sometimes
+contexts. The sorts moreover come with instantiation (sometimes
 called substitution) operations which are functorial, and all the
@@
 semantics of (\ref{eq:tytm}) is simpler and has more models than
-B-systems. GAT presentations of type theory are important for defining
-models. Our presentation gives a new notion of model, which in the
-future might contribute to definitions of new models. It can also
-influence implementations of type theory. Because our single
-substitution GAT does not involve a category, it might be easier to
-adapt it to a higher dimensional setting, see Section
-\ref{sec:conclusion}.
+B-systems.
 

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 
-The single substitution semantics is in some sense
-too minimalistic: for example, consider a term $b$ which depends on a
-context $\Gamma$ and an extra variable of type $A$, that is, $b :
-\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
-variable \emph{before} the last one obtaining $ b[\p^+] :
-\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
-the end of the context, and $\p^+$ is weakening just before the last
-variable in the context. Then, we instantiate the last $A[\p]$
-variable in the context by the previous $A$ obtaining
-$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
-denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
-We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary single substitution
-model, but this equation holds in the single substitution syntax.
-% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
-% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
-This is analogous to e.g.\ parametricity results
-\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
-arbitrary model, but they hold in the syntax. In fact, we show that
-the single substitution syntax of (\ref{eq:tt}) is equivalent to the
-parallel substitution syntax of (\ref{eq:tt}), and the techniques that
-we use are not specific to the SOGAT (\ref{eq:tt}), we expect
-that they work for any SOGAT. Equivalence of the syntaxes means that
+The single substitution semantics is in some sense too minimalistic:
+for example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
+\emph{derivable} in any model, but it is \emph{admissible} in the
+syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ meaning that it is a term
+which depends on a context $\Gamma$ and an extra variable of type $A$;
+$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version
+where we weakened \emph{before} the last variable; in
+$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
+the last variable for the previous one.)  This is analogous to
+e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
+which do not hold in an arbitrary model, but they hold in the
+syntax. In this paper, we show that for the particular SOGAT
+(\ref{eq:tt}), the \emph{syntaxes} of the single and parallel
+semantics are isomorphic. The techniques that we use for this proof
+are not specific to the SOGAT (\ref{eq:tt}), we expect
+that they work for any SOGAT. Isomorphism of the syntaxes means that
 the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
@@
 substitution models of (\ref{eq:tytm}) than parallel substitution
-models. The situation is similar to that of extensional combinatory
+models (CwFs). The situation is similar to that of extensional combinatory
 calculus and lambda calculus, where the syntaxes are equivalent, and
@@
 where every monoid is a nil/cons-algebra, but not the other way; it is
-well-known that free monoids over $X$ (syntax for monoid over $X$) are
-the same as lists of $X$ (the syntax for nil/cons algebras over $X$).
+well-known that free monoids over $X$ (syntax of monoid over $X$) are
+the same as lists of $X$ (the syntax of nil/cons algebras over $X$).
 
@@
 terms. When doing induction on types/terms, we have to provide
-a method for these operations, but this can be difficult as
+a method for these operations, but this is difficult as
 substitutions don't compose. This is why we derive a new induction
@@
 In the particular theory (\ref{eq:tt}), every type is represented by a
-term. In its single substitution calculus, this can be
-used to adapt some equations and remove some others, obtaining a more
-minimalistic, but equivalent GAT.
+term. In its single substitution calculus, we use this to adapt some
+equations and remove some others, obtaining a more minimalistic, but
+isomorphic GAT.
 

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 an algebraic way of describing type theories close to the notations
-using typing judgements. B-systems are an intrinsic, but essentially
+using typing judgements. B-systems are an intrinsic, essentially
 algebraic presentation of type theory using single
@@
 all the rules of B-systems are admissible. We describe the
-relationship in more detail. B-systems are B-frames together with a
+relationship in more detail. B-systems are B-frames together with
 substitution, weakening and generic element operations. A B-frame is
@@
 our notation, $x = (\Gamma, A, a)$ where $a : \Tm\,\Gamma\,A$, and the
-homomorphism $\mathbb{S}$ corresponds to the following maps for the
-bottom row of the diagram:
+homomorphism $\mathbb{S}$ corresponds to the following maps for $\Ty$
+(mapping between the bottom rows of the diagrams):
 \begin{alignat*}{10}
@@
 and similarly $\mathbb{S}$ also includes all the (lifted) substitution
-operations for terms corresponding to the top row of the
-diagram. Analogously, the weakening operation corresponds to
+operations for terms (top rows of the
+diagrams). Analogously, the weakening operation corresponds to
 $\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
   parallel weakenings, telescopes, or combinations of these to
-  describe the syntax of type theory in an intrinsic way.
+  describe the syntax of type theory in an intrinsic way. We present
+  our calculus in an easy-to-understand way.
 \item The $\alpha$-normalisation technique and deriving parallel
@@
 $\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
-overloading extensively.
+overloading extensively, for example $\blank[\blank]$ for types and
+terms overloaded.
 
@@
 conversion. This means that the equalities that hold between terms are
-the ways we can convert terms to each other when running them as
+the ways we can convert terms into each other when running them as
 programs. We give a tutorial-style introduction to the syntax, we do
@@
 types, the length is the number of available variables, and we add new
-variables at the end (that is, they are snoc-lists and not
+variables at the end (that is, they are snoc-lists rather than
 cons-lists). Types are also indexed by their universe level, this is a
@@
 context. We would like to describe the operation providing the last
-variable in a context, the zero De Bruijn index, we denote it by $\q :
+variable in a context, the zero De Bruijn index \cite{debruijn}, we denote it by $\q :
 \Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
@@
 operation $\blank[\blank]$ is called \emph{instantiation}. We
-instantiate the type with the weakening in
-$\Sub\,\Delta\,\Gamma$. Elements of $\Sub$ will be in general called
+instantiate the type with the $\Sub\,\Delta\,\Gamma$ weakening.
+Elements of $\Sub$ will be later called
 substitutions, hence the name, but for now, we only have single end-of
@@
 define all variables counting from the end of the context (De Bruijn
-indices \cite{debruijn}): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
+indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
 $3 := \q[\p][\p][\p]$, and so on.
@@
 new kind of weakening which adds a new variable in the last but one
-position. But then we need an equation also for $(\Pi\,A\,B)[\p']$,
+position. But after introducing $\p'$, we would need another equation computing $(\Pi\,A\,B)[\p']$,
 introducing a new weakening which adds a new variable in the last but
@@
 $\Tm\,\Delta\,(\Pi\,A\,B[\gamma])$, in the right hand side,
-$t[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
+$b[\gamma^+] : \Tm\,(\Delta\ext A[\gamma])\,(B[\gamma^+])$, hence the
 right hand side is in
@@
 $\lam\,(\lam\,(\q[\p])) :
-\Tm\,\Gamma\,\big(\Pi\,A\,(B\Ra(A[\p]))\big)$ which make sense for all
+\Tm\,\Gamma\,\big(\Pi\,A$ $(B\Ra(A[\p]))\big)$ which make sense for all
 $\Gamma$, $A$, $B$.
@@
 of the function to make sense. This is the typal version of the
-equation mentioned in the introduction.
+equation mentioned in the introduction (we don't have its term version).
 \begin{alignat*}{10}
@@
 The theory can be extended with new type and term formers in the same
-fashion: each rule has to be indexed over arbitrary contexts and come
+fashion: each operation has to be indexed over arbitrary contexts and come
 equipped with a substitution rule. There is no need to add more

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 be derived from the equations of SSC, these are versions of equations in Section
-\ref{sec:tt} lifted under arbitrary amount of $\blank^+$s.
+\ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
 \begin{equation}\label{eq:lifted}
@@
 \end{itemize}
-The predicate is defined similarly for the other type and term formers. Notably,
+The predicate is defined similarly for the other type and term formers such as $\lam$. Notably,
 we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
@@
 equality, only between types and terms instantiated with $\Sub^*$s. All
-instantiation operations, lifting operations, and substitution rules can be
+instantiation operations, lifting operations, and substitution rules are
 redefined for $\Sub^*$.

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 important to truncate the $\alpha$-normal predicates to be propositions to
-ensure that it preserves the $\beta$/$\eta$ conversion rules.
+ensure that they preserve the $\beta$/$\eta$ conversion rules.
 
@@
   \label{thm:lifting}
-  Given $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
-  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
-  $x:\Tm\,\Gamma\,A$, then
+  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if $A[\gamma_0] =
+  A[\gamma_1]$ and $x[\gamma_0] = x[\gamma_1]$ for any $A:\Ty\,\Gamma$ and
+  variable $x:\Tm\,\Gamma\,A$, then
   \begin{itemize}
-    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
-    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
-    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
+    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
+    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
+    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
   \end{itemize}
@@
 \begin{proof}
-  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
-  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
+  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
+  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
   the variable. Then the two latter equations can be proven by mutual induction

### 2024-09-18 types_2024_postproc/p.tex
stuff

@@
 \end{itemize}
-The predicate is defined similarly for the other type and term formers such as $\lam$. Notably,
-we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
-important to truncate the $\alpha$-normal predicates to be propositions to
-ensure that they preserve the $\beta$/$\eta$ conversion rules.
+The predicate is defined similarly for the other type and term formers such as
+$\lam$. Notably, we do not state that instantiated types/terms are
+$\alpha$-normal (except variables). It is important to truncate the
+$\alpha$-normal predicate to be propositions to ensure that it preserves the
+$\beta$/$\eta$ conversion rules.
 
@@
 \end{itemize}
+We define $\alpha$-normal substitutions to be the disjoint union of weakenings
+and $\alpha$-normal single substitutions.
 
@@
 \end{gather*}
-All CwF operations and equations can be defined with $\Tms$ by induction.
+All CwF operations and equations can be defined with $\Tms$ by induction, using
+Lemma \ref{thm:lifting} to avoid further induction on the syntax.
 
@@
   Contexts, types, and terms in the SSC syntax are isomorphic to the
-  corresponding sorts in the CwF syntax.
+  corresponding sorts in the CwF syntax. In addition $\Tms$ is isomorphic to CwF
+  substitutions.
 \end{theorem}

### 2024-09-18 types_2024_postproc/p.tex
tuff

--- 2024-09-18 types_2024_postproc/p.tex
@@
 we have that for any $b$ we have the equation (suitably over the input
-equation). Thus we add a conditional equation.
+equation). Thus this is a conditional equation.
 
@@
   This section defined the minimised first-order version of the SOGAT
-  (\ref{eq:tt}), relying essentially on the presernce of universes and
-  $\Pi$ types. For reference, list all the rules of the minimised
+  (\ref{eq:tt}), relying essentially on the presence of universes and
+  $\Pi$ types. For reference, we list all the rules of the minimised
   calculus in Appendix \ref{app:minimisation}.
@@
   Most operations are the same, we set $\gamma^+ :=
-  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$.
+  (\gamma\circ\p,\q)$ and $\langle a\rangle := (\id,a)$. All equations are derivable.
 \end{proof}
@@
   \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(k+i)$ by induction on $k$, together
-  with $\un^k : \Tm\,(\Lift^k\,A)\cong\Tm\,A : \mk^k$. The category
+  with $\un^k : \Tm\,(\Lift^k\,A)$ $\cong\Tm\,A : \mk^k$. The category % hack
   part of the CwF is given by types in the empty context and functions
@@
   In the definition of $\Sub$, we did not write the projections for
-  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$
-  depending on the context, and we used the truncating subtraction of
+  $\Con$, so $\Delta$ can mean $\Delta_{.1}$ or $\Delta_{.2}$, and we used the truncating subtraction of
   natural numbers. Composition of substitutions is quite involved, but
   it is just function composition written with explicit weakenings and
-  appropriate (un)lifting. The category laws hold. The empty context is
+  appropriate (un)liftings. The category laws hold. The empty context is
   modelled by $\top$, its $\eta$ law holds via $\eta$ for $\top$.
@@
   \begin{alignat*}{10}
-    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)\,(\U\,i)}) \\
+    & \Ty\,\Gamma\,i && := \Tm\,\diamond\,\big(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)}\,(\U\,i)\big) \\
     & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\[-1em]
@@
   \end{alignat*}
-  To make the notation readable, we will not write the lifting
-  decorations from now on, and we repeat the previous definitions
-  without mentioning universe levels, but all of what follows is
-  definable properly with lifting.
+  To make the notation readable, from now on, we will not write the
+  lifting decorations or universe from now on. We repeat the previous
+  definitions again without writing decorations, but all of what
+  follows can be defined properly.
   \begin{alignat*}{10}
@@
 The other roundtrip, that is, starting with an SSC-model, termifying
-it and comparing the result with the original SSC does not provide an
+it and comparing the result with the original SSC-model does not provide an
 isomorphism because we can't define the map $F :
@@
 Defining a useable syntax of type theory internally to homotopy type
-theory (HoTT \cite{HoTTbook}) is an open problem: the (h)set-truncated
+theory (HoTT \cite{HoTTbook}) is an open problem \cite{mike,nicolai}: the (h)set-truncated
 syntax can be formalised as a QIIT, but then we cannot even define the
 $\Type$-interpretation, that is, a function from the syntax which
-interprets closed types by the metatheoretic universe $\Type$ (rather
+interprets closed types in the metatheoretic universe $\Type$ (rather
 than an inductive recursive universe as in Subsection
@@
 rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
-$\Sub$, we have the associativity of composition of substitutions
-equation. If we truncate our syntax to be a set, then this is
+$\Sub$, we have an equation in the syntax on the associativity of composition of substitutions.
+If we truncate our syntax to be a set, then this is
 enough. If we don't truncate, there are two different ways of proving
@@
 two proofs are equal, or we have to remove one of them. We can remove
-the first one by limiting $[\p][\langle\rangle]$ to only work on
+the first one by limiting $[\p][\langle\rangle]$ to only apply to
 variables (this needs the introduction of a new sort for
@@
 types make the syntax non-coherent again. A possible way out of this
-would be to make the equations that depend on one of these for
-equations conditional. But then the methods of Section
+would be to make the equations that depend on one of these 
+ conditional. But then the methods of Section
 \ref{sec:admissible} don't work anymore for re-proving the needed
-admissible equations. Also, if only have the rule $[\p][^+]$ only for
-variables, then we cannot redo the termification of Section
-\ref{sec:cwf}, in particular it is not clear how to prove equation
+admissible equations. Also, if we only have the rule $[\p][^+]$ for
+variables, we cannot redo the termification of Section
+\ref{sec:cwf}, e.g.\ it is not clear how to prove equation
 (\ref{eq:functor}).
@@
 syntax is isomorphic to the parallel syntax, and we conjecture that
-our method works for any SOGAT. For our particular SOGAT, the single
-substitution syntax can be simplified, and we even have that any
+our method works for any SOGAT. For our particular SOGAT, we minimised the single
+substitution syntax, and we even have that any
 single substitution model is equivalent to a parallel model in a weak

### 2025-01-24 cpp/p.tex
stuff

@@
 variables, and there are four equations on types which are needed to
-typecheck other equations. All the other equations are substitution
-(naturality) rules or computation rules for different type
-formers. SSC provides a simple, minimalistic alternative to parallel
-substitution calculi or B-systems for defining type theory. SSC
-relates to CwF as extensional combinatory calculus relates to lambda
-calculus. If we have some additional type formers, we show that SSC models
-are actually equivalent to CwF-based models in a weak sense. Most
-results in this paper were formalised in Agda.
+typecheck the other equations. SSC provides a simple, minimalistic
+alternative to parallel substitution calculi or B-systems for defining
+type theory. SSC relates to CwF as extensional combinatory calculus
+relates to lambda calculus: there are more models of the former, but
+the syntaxes are equivalent. If we have some additional type formers,
+we show that SSC models are actually equivalent to CwF-based models in
+a weak sense. Most results in this paper were formalised in Agda.
 \end{abstract}
@@
 
+In this paper, we will answer this question by defining type theory as
+a generalised algebraic theory (GAT
+\cite{DBLP:journals/apal/Cartmell86}). This has the following
+properties.
+\begin{enumerate}
+\item[(i)] It is an intrinsic presentation
+  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
+  opposed to extrinsic
+  \cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}. That
+  is, we only consider well-formed, well-scoped and well-typed
+  abstract syntax trees, there are no meaningless terms.
+\item[(ii)] The equations of the algebraic theory describe the
+  (symmetric) conversion relation.
+\end{enumerate}
+For type theory, such an intrinsic quotiented presentation is very
+natural as typing depends on conversion. The algebraic approach avoids
+lots of boilerplate coming from using an explicit conversion relation
+(as in e.g.\ \cite{chapman09eatitself}). In the algebraic approach,
+syntax and semantics are concordant: the syntax is simply the initial
+model, which always exists for any GAT
+\cite{DBLP:journals/pacmpl/KaposiKA19}. Such a syntax is quotiented
+which means that any function defined on the syntax needs to respect
+conversion. For example, it is not possible to define a printing
+function which distinguishes the terms $(\lambda x.x)\,1$ and $1$, as
+these are convertible. However, it is still possible to define
+normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
+this level of abstraction.
+
+The algebraic/intrinsic approach is also more abstract than extrinsic
+approaches, which means that we don't have to make ad-hoc choices. For
+example, we do not have to decide whether we want function space {\`a}
+la Curry or Church, whether we work with paranoid or economic rules
+\cite[Section 5]{winterhalter:hal-01626651}, or whether we have
+explicit or recursively defined substitution (instantiation of
+variables by terms). These choices are forced by the notion of GAT:
+the domain of a lambda appears in its type, se we are always {\`a} la
+Church; algebraic theories don't allow recursively defined operators,
+so instantiation is an explicit operation, and so on. However, there
+are still some options for the exact definition of the substitution
+calculus: most notions of type theory use parallel substitutions,
+which means that the instantiation operation replaces all variables in
+a context by terms in another context \emph{at the same time}. Some
+examples of algebraic definitions of type theory using parallel
+substitutions ranging from the more syntactic to the more semantic:
+Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories with
+families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's
+natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
+categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
+\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
+categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
+\cite{10.1145/3204492}. In traditional extrinsic presentations
+intantiation replaces one variable by a single term. To bridge the gap
+between the intrinsic parallel substitution world and the extrinsic
+single substitution world, Voevodsky introduced B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
+desription of a single substitution calculus. B-systems involve
+complex rules describing telescopes, weakenings and substitutions
+under telescopes, with several equations.
+
+In this paper, we introduce a new single substitution calculus which
+is simpler and more minimalistic than the previous algebraic
+approaches. It shows that to explain type theory in an algebraic way,
+there is no need for categories, parallel substitutions or weakenings,
+empty substitution, telescopes, or combinations of these. All our
+operations are easy to motivate, we showcase this in Section
+\ref{sec:tt} which is a tutorial introduction to our theory. Our
+single substitution calculus is in a sweet spot: our notations read
+like traditional extrinsic syntactic presentations, but they are
+algebraic, thus come with a well behaved model theory, and can be
+easily related to more semantic descriptions.
+
+Our single substitution theory is in some sense too minimalistic: for
+example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
+\emph{derivable} in any model, but it is \emph{admissible} in the
+syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ is a term which depends on
+a context $\Gamma$ and an extra variable of type $A$; $b[\p^+] :
+\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version where we
+weakened \emph{before} the last variable; in
+$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
+the last variable for the previous one.)  This is analogous to
+e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
+which do not hold in an arbitrary model, but they hold in the
+syntax. In this paper, we show that the \emph{syntax} of our single
+substitution theory and the syntax of the corresponding parallel
+substitution theory (CwF-based theory) are isomorphic. All CwF-based
+models are models of our single substitution theory, but not
+necessarily the other way. While B-systems are equivalent to CwFs (and
+C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
+single substitution models than CwF-based models. This relationship is
+similar to that of extensional combinatory calculus and lambda
+calculus, where the syntaxes are equivalent, and every lambda model is
+a combinatory model, but not the other way round
+\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
+relationship between monoids over a set $X$ and nil/cons algebras over
+$X$, where every monoid is a nil/cons-algebra, but not the other way;
+but it is well-known that free monoids over $X$ (syntax of monoid over
+$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
+over $X$).
+
+Technology to show that syntaxes are equivalent: alpha normalisation
+-- terms which don't include instantiation, but are still quotiented.
+
+Minimisation: (i) if we have a universe representing every type as a
+term, we can remove some equation, obtaining a more minimalistic, but
+isomorphic GAT. (ii) if we even have Sigma tyes, we can actually
+derive a parallel model from an SSC model.
+
+SOGAT stuff
+
+The techniques that we use for this proof are not specific to the
+SOGAT (\ref{eq:tt}), we expect that they work for any
+SOGAT. Isomorphism of the syntaxes means that the corresponding sorts
+($\Con$, $\Ty$, $\Tm$) are isomorphic. For a general SOGAT, there are
+more models of the corresponding single substitution GAT than the
+parallel substitution GAT.
+
+xxx
+
+For formalisation on a computer, the algebraic approach also has its
+downsides: as mutually defined recursive operations are not allowed,
+substitution (instantiation of variables by terms) is explicit, which
+means that it is a constructor, and does not compute by
+definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
+  u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
+is not definitional. This makes formalisation difficult, and results
+in the situation called ``transport hell''.
+
+Related work:
+- SOGAT->GAT conversion
+- B systems
+- others
+
 Type theory is a language which can be described as a second-order
@@
 \cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
-\begin{enumerate}
-\item[(i)] It is an intrinsic presentation
-  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
-  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
-  only consider well-formed, well-scoped and well-typed abstract
-  syntax trees, there are no meaningless terms.
-\item[(ii)] Terms are quotiented by the conversion relation.
-\item[(iii)] Every operation is stable under substitution.
-\end{enumerate}
+
 For example, type theory with $\Pi$ types and universes {\`a} la
@@
 
-The single substitution semantics is in some sense too minimalistic:
-for example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
-\emph{derivable} in any model, but it is \emph{admissible} in the
-syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ meaning that it is a term
-which depends on a context $\Gamma$ and an extra variable of type $A$;
-$b[\p^+] : \Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version
-where we weakened \emph{before} the last variable; in
-$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
-the last variable for the previous one.)  This is analogous to
-e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
-which do not hold in an arbitrary model, but they hold in the
-syntax. In this paper, we show that for the particular SOGAT
-(\ref{eq:tt}), the \emph{syntaxes} of the single and parallel
-semantics are isomorphic. The techniques that we use for this proof
-are not specific to the SOGAT (\ref{eq:tt}), we expect
-that they work for any SOGAT. Isomorphism of the syntaxes means that
-the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
-general SOGAT, there are more models of the correponding single
-substitution GAT than the parallel substitution GAT. Every parallel
-model is also a single model, but not necessarily the other way. While
-B-systems are equivalent to CwFs (and C-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
-substitution models of (\ref{eq:tytm}) than parallel substitution
-models (CwFs). The situation is similar to that of extensional combinatory
-calculus and lambda calculus, where the syntaxes are equivalent, and
-every lambda model is a combinatory model, but not the other way round
-\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
-relationship of monoids over a set $X$ and nil/cons algebras over $X$,
-where every monoid is a nil/cons-algebra, but not the other way; it is
-well-known that free monoids over $X$ (syntax of monoid over $X$) are
-the same as lists of $X$ (the syntax of nil/cons algebras over $X$).
-
 When showing that the parallel and single substitution syntaxes are
@@
 Using the same insight about types and the universe, the fact that
-dependences can be represented by function spaces, and that
+dependencies can be represented by function spaces, and that
 collections of dependencies can be collected using $\Sigma$ types, we
@@
 incompatible with homotopy type theory \cite{HoTTbook}. Our notation
-on paper is close to Agda's, we write $\Pi$ types as $(x:A)\ra B$,
-$\Sigma$ types as $(x:A)\times B$, and use implicit arguments and
-overloading extensively, for example $\blank[\blank]$ for types and
-terms overloaded.
+on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
+$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
+overloading extensively, for example instantiation $\blank[\blank]$ is
+overloaded for types and terms.
 
@@
 
+TODO: named CwFs would be even closer to usual syntax.
+
 \bibliography{b}{}

### 2025-01-26 cpp/p.tex
stuff

@@
 a generalised algebraic theory (GAT
-\cite{DBLP:journals/apal/Cartmell86}). This has the following
-properties.
-\begin{enumerate}
-\item[(i)] It is an intrinsic presentation
-  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
-  opposed to extrinsic
-  \cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}. That
-  is, we only consider well-formed, well-scoped and well-typed
-  abstract syntax trees, there are no meaningless terms.
-\item[(ii)] The equations of the algebraic theory describe the
-  (symmetric) conversion relation.
-\end{enumerate}
-For type theory, such an intrinsic quotiented presentation is very
-natural as typing depends on conversion. The algebraic approach avoids
-lots of boilerplate coming from using an explicit conversion relation
-(as in e.g.\ \cite{chapman09eatitself}). In the algebraic approach,
-syntax and semantics are concordant: the syntax is simply the initial
-model, which always exists for any GAT
-\cite{DBLP:journals/pacmpl/KaposiKA19}. Such a syntax is quotiented
-which means that any function defined on the syntax needs to respect
-conversion. For example, it is not possible to define a printing
-function which distinguishes the terms $(\lambda x.x)\,1$ and $1$, as
-these are convertible. However, it is still possible to define
+\cite{DBLP:journals/apal/Cartmell86}). GATs are multi-sorted algebraic
+theories where later sorts can be indexed by previous sorts. An
+example is categories where the sort of morphisms is double-indexed
+over objects.
+
+The GAT presentation of type theory is intrinsic
+\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} rather
+than extrinsic
+\cite{alti:phd93,abel2013normalization,DBLP:conf/cpp/AdjedjLMPP24}:
+this means that we only consider well-formed, well-scoped and
+well-typed abstract syntax trees, there are no meaningless terms; in
+the GAT presentation, the conversion relation is expressed by
+equations of the algebraic theory.
+
+Some advantages of using a GAT presentation:
+\begin{itemize}
+\item  In type theory, typing depends on conversion:
+  in an extrinsic presentation, this is expressed by a rule such as
+  the following.
+  \[
+  \infer{\Gamma\vdash t : B}{A \sim B && \Gamma\vdash t : A}
+  \]
+  When conversion is equality, this rule holds by the general
+  properties of equality; and we also avoid lots of boilerplate
+  stating that e.g.\ conversion is a congruence with respect to all
+  operations (for an intrinsic presentation with explicit conversion
+  relation see \cite{chapman09eatitself}).
+\item Syntax and semantics are concordant: the syntax is simply the
+  initial model, which always exists for any GAT and is called a
+  quotient inductive-inductive type
+  \cite{DBLP:journals/pacmpl/KaposiKA19}. The fact that the syntax is
+  quotiented means that any function defined on it needs to respect
+  the equations. For example, it is not possible to define a printing
+  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
+  as these are convertible. However, it is still possible to define
   normalisation \cite{lmcs:4005} and typechecking \cite{typecheck} at
   this level of abstraction.
-
-The algebraic/intrinsic approach is also more abstract than extrinsic
-approaches, which means that we don't have to make ad-hoc choices. For
-example, we do not have to decide whether we want function space {\`a}
-la Curry or Church, whether we work with paranoid or economic rules
-\cite[Section 5]{winterhalter:hal-01626651}, or whether we have
-explicit or recursively defined substitution (instantiation of
-variables by terms). These choices are forced by the notion of GAT:
-the domain of a lambda appears in its type, se we are always {\`a} la
-Church; algebraic theories don't allow recursively defined operators,
-so instantiation is an explicit operation, and so on. However, there
-are still some options for the exact definition of the substitution
-calculus: most notions of type theory use parallel substitutions,
-which means that the instantiation operation replaces all variables in
-a context by terms in another context \emph{at the same time}. Some
-examples of algebraic definitions of type theory using parallel
-substitutions ranging from the more syntactic to the more semantic:
-Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories with
-families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's
-natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
-categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
+\item The GAT approach is also more abstract than extrinsic
+  approaches, which means that we don't have to make ad-hoc
+  choices. For example, we do not have to decide whether we want
+  function space {\`a} la Curry or {\`a} la Church, whether we work
+  with paranoid or economic typing rules \cite[Section
+    5]{winterhalter:hal-01626651}, or whether we have explicit or
+  recursively defined substitution (instantiation of variables by
+  terms, written $t[x\mapsto u]$ or $t[u/x]$).  These choices are
+  forced by the notion of GAT: the domain of a lambda appears in its
+  type, se we are always {\`a} la Church; algebraic theories don't
+  allow recursively defined operators, so instantiation is an explicit
+  operation, and so on.
+\end{itemize}
+
+% For formalisation on a computer, the algebraic approach also has its
+% downsides: as mutually defined recursive operations are not allowed,
+% substitution (instantiation of variables by terms) is explicit, which
+% means that it is a constructor, and does not compute by
+% definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
+%   u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
+% is not definitional. This makes formalisation difficult, and results
+% in the situation called ``transport hell''.
+
+However, even at the GAT level of abstraction, there are some choices
+to be made about the instantiation operation. It is usually convenient
+to use parallel substitutions, which means that the instantiation
+operation replaces all variables in a context by terms in another
+context \emph{at the same time}, that is, $t[x_0\mapsto
+  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are all the free
+variables in $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is called a
+parallel substitution, and the terms $u_0,\dots,u_n$ are all in the
+same context. Parallel substitutions are very natural in the algebraic
+setting because they form the morphisms in a category where objects
+are contexts (lists of types), and then it is clear what equations
+they need to satisfy. Some examples of algebraic definitions of type
+theory using parallel substitutions ranging from the more syntactic to
+the more semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard},
+categories with families (CwF
+\cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's natural
+models \cite{DBLP:journals/mscs/Awodey18}, contextual categories
+\cite{DBLP:journals/apal/Cartmell86}, C-systems
 \cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
 categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
-\cite{10.1145/3204492}. In traditional extrinsic presentations
-intantiation replaces one variable by a single term. To bridge the gap
-between the intrinsic parallel substitution world and the extrinsic
-single substitution world, Voevodsky introduced B-systems
+\cite{10.1145/3204492}.
+
+In traditional extrinsic presentations
+(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
+simply replaces all free occurrences of a variable $x$ with a term $u$
+as in the notation $t[x\mapsto u]$. A substitutions consists of a
+variable name (a pointer into a context) and a term. Substitutions
+cannot be composed, thus they do not form a category. To bridge the
+gap between the intrinsic parallel substitution world and the
+extrinsic single substitution world, Voevodsky introduced B-systems
 \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
@@
 
-In this paper, we introduce a new single substitution calculus which
-is simpler and more minimalistic than the previous algebraic
-approaches. It shows that to explain type theory in an algebraic way,
-there is no need for categories, parallel substitutions or weakenings,
-empty substitution, telescopes, or combinations of these. All our
-operations are easy to motivate, we showcase this in Section
-\ref{sec:tt} which is a tutorial introduction to our theory. Our
-single substitution calculus is in a sweet spot: our notations read
-like traditional extrinsic syntactic presentations, but they are
+In this paper, we introduce a new single substitution calculus (SSC)
+which is simpler and more minimalistic than the existing algebraic
+approaches. The instantiation operation $\blank[\blank]$ takes either a
+single substitution or a single weakening as an argument, and there
+are eight equations explaining how these behave: four of these say how
+to instantiate variables and the other four equations are needed to
+typecheck the first four. The SSC shows that to explain type theory in
+an algebraic way, there is no need for categories, parallel
+substitutions or weakenings, empty substitution, telescopes, or
+combinations of these. All our operations are easy to motivate, we
+showcase this in Section \ref{sec:tt} which is a tutorial introduction
+to our theory. We believe that our SSC is in a sweet spot: our
+notations are close to traditional extrinsic notations, but they are
 algebraic, thus come with a well behaved model theory, and can be
@@
 
-Our single substitution theory is in some sense too minimalistic: for
-example, the equation $b[\p^+][\langle\q\rangle] = b$ is not
-\emph{derivable} in any model, but it is \emph{admissible} in the
-syntax. (Here $b : \Tm\,(\Gamma\ext A)\,B$ is a term which depends on
-a context $\Gamma$ and an extra variable of type $A$; $b[\p^+] :
-\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p^+])$ is a version where we
-weakened \emph{before} the last variable; in
-$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ we substituted
-the last variable for the previous one.)  This is analogous to
-e.g.\ parametricity results \cite{DBLP:journals/jfp/BernardyJP12},
-which do not hold in an arbitrary model, but they hold in the
-syntax. In this paper, we show that the \emph{syntax} of our single
-substitution theory and the syntax of the corresponding parallel
-substitution theory (CwF-based theory) are isomorphic. All CwF-based
-models are models of our single substitution theory, but not
-necessarily the other way. While B-systems are equivalent to CwFs (and
-C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
-single substitution models than CwF-based models. This relationship is
-similar to that of extensional combinatory calculus and lambda
-calculus, where the syntaxes are equivalent, and every lambda model is
-a combinatory model, but not the other way round
-\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
+Our SSC is in some sense too minimalistic: for example, the equation
+$b[\p^+][\langle\q\rangle] = b$ is not \emph{derivable} in any model,
+but it is \emph{admissible} in the syntax. (Here $b : \Tm\,(\Gamma\ext
+A)\,B$ is a term which depends on a context $\Gamma$ and an extra
+variable of type $A$; $b[\p^+] : \Tm\,(\Gamma\ext A\ext
+A[\p])\,(B[\p^+])$ is a version where we weakened \emph{before} the
+last variable; in $b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$
+we substituted the last variable for the previous one.)  This is
+analogous to e.g.\ parametricity results
+\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
+arbitrary model, but they hold in the syntax. In this paper, we show
+that the \emph{syntax} of our single substitution theory and the
+syntax of the corresponding parallel substitution theory (CwF-based
+theory) are isomorphic. All CwF-based models are models of our single
+substitution theory, but not necessarily the other way. While
+B-systems are equivalent to CwFs (and C-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more SSC models
+than CwF-based models. This relationship is similar to that of
+extensional combinatory calculus and lambda calculus, where the
+syntaxes are equivalent, and every lambda model is a combinatory
+model, but not the other way round
+\cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
 relationship between monoids over a set $X$ and nil/cons algebras over
@@
 
-Technology to show that syntaxes are equivalent: alpha normalisation
--- terms which don't include instantiation, but are still quotiented.
-
-Minimisation: (i) if we have a universe representing every type as a
-term, we can remove some equation, obtaining a more minimalistic, but
-isomorphic GAT. (ii) if we even have Sigma tyes, we can actually
-derive a parallel model from an SSC model.
-
-SOGAT stuff
-
-The techniques that we use for this proof are not specific to the
-SOGAT (\ref{eq:tt}), we expect that they work for any
-SOGAT. Isomorphism of the syntaxes means that the corresponding sorts
-($\Con$, $\Ty$, $\Tm$) are isomorphic. For a general SOGAT, there are
-more models of the corresponding single substitution GAT than the
-parallel substitution GAT.
-
-xxx
-
-For formalisation on a computer, the algebraic approach also has its
-downsides: as mutually defined recursive operations are not allowed,
-substitution (instantiation of variables by terms) is explicit, which
-means that it is a constructor, and does not compute by
-definition. For example, $(t\,a)[x\mapsto u] = (t[x\mapsto
-  u])\,(a[x\mapsto u])$ is a derivable equation in the syntax, but it
-is not definitional. This makes formalisation difficult, and results
-in the situation called ``transport hell''.
-
-Related work:
-- SOGAT->GAT conversion
-- B systems
-- others
-
-Type theory is a language which can be described as a second-order
-generalised algebraic theory (SOGAT
-\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
-
-For example, type theory with $\Pi$ types and universes {\`a} la
-Coquand \cite{coquandUniverse} is specified by the following SOGAT.
-\begin{equation}\label{eq:tt}
-\begin{alignedat}{10}
-  & \Ty && : \N\ra\Set \\
-  & \Tm && : \Ty\,i\ra\Set \\
-  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
-  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
-  & \U && : (i:\N)\ra\Ty\,(1+i) \\
-  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
-  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
-  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
-\end{alignedat}
-\end{equation}
-Because types are isomorphic to certain terms (witnessed by $\El$ and $\cd$),
-we have to index types by their (universe) level, and types at lower
-levels are included at upper levels by $\Lift$. We use implicit
-arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
-$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
-isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
-and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
-second-order functions. We say that $\Pi$ binds a $\Tm$-variable
-in its second $\Ty$-argument. Every binder is represented by
-a second-order function, and as a result, we don't need to talk about
-contexts and substitutions when describing the theory: we re-use the
-binding features of the metatheory to describe those of our object
-theory (this technique is also known as higher-order abstract syntax
-\cite{DBLP:conf/lics/Hofmann99}, logical framework
-\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} or
-two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
-second-order models, thus the semantics of a SOGAT is given by a first
-order generalised algebraic theory (GAT
-\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
-single-sorted algebraic theories (such as monoids, groups) to multiple
-sorts where later sorts can be indexed over previous sorts. The usual
-results from universal algebra (e.g.\ models form a category with
-finite limits and initial object, existence of free and co-free
-models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
-example of a GAT is the theory of categories where there is a sort of
-objects and a sort of morphisms double-indexed over objects.
-
-Kaposi and Xie \cite{DBLP:conf/fscd/KaposiX24} present two different
-first order semantics for any SOGAT: one based on parallel
-substitutions, and one based on single substitutions.
-
-In the parallel semantics, the GAT always starts with a category with
-a terminal object, objects of the category are called contexts,
-morphisms are substitutions. Every sort/operation of the SOGAT gives
-rise to a sort/operation in the GAT which is now indexed by
-contexts. The sorts moreover come with instantiation (sometimes
-called substitution) operations which are functorial, and all the
-operations are natural with respect to instantiation. For those sorts
-which are bound by an operation in the SOGAT description, we have
-context extension, meaning that contexts can carry variables of that
-sort (in other words, the sort is given by a locally representable
-presheaf). The parallel semantics of the SOGAT
-\begin{equation}\label{eq:tytm}
-\begin{alignedat}{10}
-  & \Ty && : \Set \\
-  & \Tm && : \Ty \ra\Set 
-\end{alignedat}
-\end{equation}
-is the GAT of categories with families (CwFs
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
-We define the notion ``model of the SOGAT'' by its parallel substitution
-semantics, which is a GAT, and that has an obvious notion of model. The
-syntax of the language is the initial model, which exists for every
-GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
-syntax (only well-typed terms), quotiented by the equations in the
-SOGAT, which describe conversion for the language.
-
-The single substitution semantics is a GAT which does not start with a
-category: there are contexts and maps between contexts, but these maps
-are single substitutions or single weakenings, they don't
-compose. There are fewer operations and fewer equations than in the
-parallel semantics. In this paper we investigate the single
-substitution semantics of the type theory specified by the SOGAT
-(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
-the operations are all easy to motivate, we illustrate this in Section
-\ref{sec:tt}. Single substitution calculi are closer to the usual
-presentation of type theory using derivation rules, this was one of
-the motivations of Voevodsky for developing B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
-semantics of (\ref{eq:tytm}) is simpler and has more models than
-B-systems.
-
-When showing that the parallel and single substitution syntaxes are
-equivalent, we define the parallel operations by induction on the
-single substitution syntax. One of the operations in the single
-substitution syntax is the instantiation operation: $\blank[\blank] :
-\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
-terms. When doing induction on types/terms, we have to provide
-a method for these operations, but this is difficult as
-substitutions don't compose. This is why we derive a new induction
-principle which works on $\alpha$-normal types and terms. Just as
-$\beta$-normal forms don't distinguish between $\beta$-equal terms,
-$\alpha$-normal forms don't include explicit substitutions.
-$\alpha$-normal forms are not usual normal
-forms in the sense that they are still quotiented by the
-computation/uniqueness rules like $\beta$/$\eta$ for function
-space. To derive the new induction principle for the syntax, we first
-prove $\alpha$-normalisation.
-
-In the particular theory (\ref{eq:tt}), every type is represented by a
-term. In its single substitution calculus, we use this to adapt some
-equations and remove some others, obtaining a more minimalistic, but
-isomorphic GAT.
-
-Theory (\ref{eq:tt}) can be extended
-with unit and $\Sigma$ types, more precisely with the following rules.
-\begin{equation}\label{eq:sigma}
-  \begin{alignedat}{10}
-    & \top && : \Ty\,0 \\
-    & \tt && : \Tm\,\top \\
-    & \top\eta && : t = \tt \\
-    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
-    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
-\end{alignedat}
-\end{equation}
-Using the same insight about types and the universe, the fact that
-dependencies can be represented by function spaces, and that
-collections of dependencies can be collected using $\Sigma$ types, we
-can actually derive a parallel model from any single substitution
-model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
-single and parallel substitution models are actually equivalent in a
-weak sense.
-
-To summarise, our contributions are the following:
+We introduce a technology for proving admissible rules in our
+theory. We define $\alpha$-normal forms and show
+$\alpha$-normalisation: every term admits an $\alpha$-normal form.
+Thus, when one wants to prove something about all terms, it is enough
+to prove it for $\alpha$-normal forms. Just as $\beta$-normal forms
+don't distinguish between $\beta$-convertible terms, $\alpha$-normal
+forms don't include explicit substitutions. $\alpha$-normal forms are
+not like ordinary normal forms because they are still quotiented by
+the computation/uniqueness rules like $\beta$/$\eta$ for function
+space. We use induction on $\alpha$-normal forms to define parallel
+substitutions and prove all the rules of CwFs in the syntax of SSC.
+
+Our SSC has 8 equations, and these are enough to describe type
+theories with arbitrary choice of type formers (see the description of
+SOGATs in Subsection \ref{sec:related}). In the particular case of
+type theory with a Coquand-universe and $\Pi$ types, we can do better:
+4 (conditional) equations on variables are enough, because using the
+fact that every type is represented by a terma, we can derive the other
+4 equations.
+
+If in addition to having a universe and $\Pi$ types, our type theory
+also supports unit and $\Sigma$ types, we can actually derive a
+CwF-based model from an SSC-model. The idea is that $\Sigma$ types can
+represent contexts, functions between $\Sigma$ types represent
+parallel substitutions, functions into the universe represent depenent
+types. Then the SSC is just there to formulate $\Sigma$, $\Pi$, $\U$,
+and then we use these to define a parallel substitution calculus. So
+although SSC models and CwFs without any type formers are not
+equivalent, if we equip them with enough type formers, they become
+(weakly) equivalent.
+
+All of our constructions can be understood as happening internal to
+extenional type theory, and most results in this paper are formalised
+in Agda. Most results in this paper were presented at the TYPES 2024
+conference \cite{singleTypes}.
+
 \begin{itemize}
@@
 
-In Section \ref{sec:tt}, we introduce the single substitution calculus
-(SSC) only relying on a working knowledge of an implementation of type
-theory such as Agda, no prior experience with the metatheory of type
-theory is required. Section \ref{sec:admissible} shows that in the
-syntax of SSC, several new equations are admissible, and that the
-syntax of SSC is actually isomorphic to the CwF-based syntax. In
-Section \ref{sec:minimisation} we minimise our SSC obtaining a theory
-with fewer equations, relying on the presence of $\Pi$ and $\U$ in our
-theory. In Section \ref{sec:cwf}, we show that in the presence of
-$\Pi$, $\U$ and $\Sigma$, SSC-models are actually equivalent to
-CwF-models in a weak sense. We conclude in Section
-\ref{sec:conclusion}.
+After discussing related work, in Section \ref{sec:tt}, we introduce
+our single substitution calculus (SSC) only relying on a working
+knowledge of an implementation of type theory such as Agda, no prior
+experience with the metatheory of type theory is required. Section
+\ref{sec:admissible} shows that in the syntax of SSC, several new
+equations are admissible, and that the syntax of SSC is actually
+isomorphic to the CwF-based syntax. In Section \ref{sec:minimisation}
+we minimise our SSC obtaining a theory with fewer equations, relying
+on the presence of $\Pi$ and $\U$ in our theory. In Section
+\ref{sec:cwf}, we show that in the presence of $\Pi$, $\U$ and
+$\Sigma$, SSC-models are actually equivalent to CwF-models in a weak
+sense. We conclude in Section \ref{sec:conclusion}.
 
@@
 
-In the 1990s single substitution calculi for type theory were popular,
-but usually in the extrinsic setting, see e.g.\ the thesis of
-Altenkirch \cite{alti:phd93}.
-
-B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
+\paragraph{B-systems} B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
 an algebraic way of describing type theories close to the notations
@@
 algebraic presentation of type theory using single
-substitutions. B-systems correspond to our single substitution
-calculus as essentially algebraic theories correpond to generalised
-algebraic theories, or sets with a map into $I$ correspond to indexed
-families over $I$ \cite[page
-  221]{DBLP:journals/apal/Cartmell86}. Another difference is that we
-have fewer and less general equations, resulting in the fact that we
-have more models than B-systems. However in the syntax of our theory,
-all the rules of B-systems are admissible. We describe the
+substitutions. B-systems correspond to our SSC as essentially
+algebraic theories correspond to generalised algebraic theories, or
+sets with a map into $I$ correspond to indexed families over $I$
+\cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
+that we have fewer and less general equations, resulting in the fact
+that we have more models than B-systems. However in the syntax of our
+theory, all the rules of B-systems are admissible. We describe the
 relationship in more detail. B-systems are B-frames together with
@@
 
+\paragraph{Ehrhard's calculus}
 The first generalised algebraic presentation of type theory was Thomas
@@
 
+\paragraph{Formalisations of type theory}
 Most computer formalisations of type theory are extrinsic
@@
 \cite{initiality-agda} constructed the initial contextual category in
-Agda using postulated quotients. QIITs are supported by the cubical
-set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model
-\cite{kaposi-qiit-setoid}, thus Cubical Agda
+Agda using postulated quotients. Although there is no general proof
+published about this, we have experimental evidence that QIITs are
+supported by the cubical set model \cite{DBLP:conf/lics/CoquandHM18}
+and the setoid model \cite{kaposi-qiit-setoid}, thus Cubical Agda
 \cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
@@
 
-At the level of abstraction of SOGATs
-\cite{uemura,DBLP:conf/fscd/KaposiX24}, the difference between single
-and parallel substitution calculi are not visible. However, the
-metatheory of type theory can be still investigated in this abstract
-setting, via the methods of Synthethic Tait Computability
+\paragraph{SOGATs}
+We can define languages with binders as second-order theories
+following higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99}
+and logical frameworks
+\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
+generalised algebraic theories (SOGATs
+\cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
+theory even more abstractly than GATs: they even abstract over the way
+of defining the substitution calculus: contexts, substitutions,
+instantiations are simply modelled by the metatheoretic function
+space. For example, a type theory with $\Pi$, Coquand-universes and a
+universe level lifting operation is described by the following SOGAT
+(where $\cong$ means isomorphism):
+\begin{equation}\label{eq:tt}
+\begin{alignedat}{10}
+  & \Ty && : \N\ra\Set \\
+  & \Tm && : \Ty\,i\ra\Set \\
+  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
+  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
+  & \U && : (i:\N)\ra\Ty\,(1+i) \\
+  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
+  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
+  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
+\end{alignedat}
+\end{equation}
+% Adding $\top$ and $\Sigma$ types (Definition \ref{def:sigma}) is
+% described as the following extension of the above SOGAT:
+% \begin{equation}\label{eq:sigma}
+%   \begin{alignedat}{10}
+%     & \top && : \Ty\,0 \\
+%     & \tt && : \Tm\,\top \\
+%     & \top\eta && : t = \tt \\
+%     & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
+%     & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
+% \end{alignedat}
+% \end{equation}
+Second-order models of SOGATs are not well-behaved as there is no good
+notion of homomorphism between them. This is why Kaposi and Xie
+\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: each GAT
+comes with a category of models with initial model (syntax), (co)free
+models, and so on. Actually, Kaposi and Xie define two different
+translations: one based on parallel substitutions and one based on
+single substitutions. If we apply their parallel translation to the
+SOGAT
+\begin{equation}\label{eq:tytm}
+\begin{alignedat}{10}
+  & \Ty && : \Set \\
+  & \Tm && : \Ty \ra\Set
+\end{alignedat}
+\end{equation}
+with only two sorts and no operations or equations, we obtain the GAT
+of category with families (CwF). If we use their single substitution
+translation on the SOGAT (\ref{eq:tt}), we obtain Definition
+\ref{def:tt} (also listed in Appendix \ref{app:tt}). This paper can be
+seen as taking a big enough type theory (described as a SOGAT), and
+investigating how the result of its single substitution translation
+relates to the result of its parallel translation (the latter
+first-order theory is quite well-known).
+
+There are techniques to prove properties of type theories without
+choosing between parallel or single substitutions, and staying at the
+SOGAT level of abstraction: Synthethic Tait Computability
 \cite{DBLP:phd/us/Sterling22} or internal sconing
 \cite{DBLP:conf/fscd/BocquetKS23}.
-
-% TODO: say that this is the paper version of the \cite{singleTypes}
-% paper
 
@@
   To make the notation readable, from now on, we will not write the
-  lifting decorations or universe from now on. We repeat the previous
-  definitions again without writing decorations, but all of what
-  follows can be defined properly.
+  lifting decorations or universe levels. We repeat the previous
+  definitions again without writing decorations.
   \begin{alignat*}{10}

### 2025-01-30 cpp/p.tex
stuff

@@
   \[
-  \infer{\Gamma\vdash t : B}{A \sim B && \Gamma\vdash t : A}
+  \infer{\Gamma\vdash t : B}{\Gamma\vdash A \sim B && \Gamma\vdash t : A}
   \]
-  When conversion is equality, this rule holds by the general
+  When conversion $\sim$ is equality, this rule holds by the general
   properties of equality; and we also avoid lots of boilerplate
@@
   choices. For example, we do not have to decide whether we want
-  function space {\`a} la Curry or {\`a} la Church, whether we work
+  function space {\`a} la Curry or {\`a} la Church; whether we work
   with paranoid or economic typing rules \cite[Section
-    5]{winterhalter:hal-01626651}, or whether we have explicit or
+    5]{winterhalter:hal-01626651}; whether we have explicit or
   recursively defined substitution (instantiation of variables by
@@
   forced by the notion of GAT: the domain of a lambda appears in its
-  type, se we are always {\`a} la Church; algebraic theories don't
+  type, so we are always {\`a} la Church; algebraic theories don't
   allow recursively defined operators, so instantiation is an explicit
-  operation, and so on.
+  operation; and so on.
 \end{itemize}
@@
 context \emph{at the same time}, that is, $t[x_0\mapsto
-  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are all the free
-variables in $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is called a
-parallel substitution, and the terms $u_0,\dots,u_n$ are all in the
-same context. Parallel substitutions are very natural in the algebraic
-setting because they form the morphisms in a category where objects
-are contexts (lists of types), and then it is clear what equations
-they need to satisfy. Some examples of algebraic definitions of type
-theory using parallel substitutions ranging from the more syntactic to
-the more semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard},
-categories with families (CwF
-\cite{DBLP:conf/types/Dybjer95,Castellan2021}), Awodey's natural
-models \cite{DBLP:journals/mscs/Awodey18}, contextual categories
-\cite{DBLP:journals/apal/Cartmell86}, C-systems
+  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
+in the context of $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is
+called a parallel substitution, where the terms $u_0,\dots,u_n$ are
+all in the same context. Parallel substitutions are very natural in
+the algebraic setting because they form the morphisms in a category
+where objects are contexts (lists of types), then types/terms are
+preheaves, context extension is a representability condition, so the
+equations of the algebraic theory are forced by the categorical
+structure. Some examples of algebraic definitions of type theory using
+parallel substitutions ranging from the more syntactic to the more
+semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
+with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
+Awodey's natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
+categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
 \cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
@@
 (e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
-simply replaces all free occurrences of a variable $x$ with a term $u$
-as in the notation $t[x\mapsto u]$. A substitutions consists of a
+replaces free occurrences of a single variable $x$ with a term $u$ as
+in the notation $t[x\mapsto u]$. A substitution consists of a
 variable name (a pointer into a context) and a term. Substitutions
@@
 In this paper, we introduce a new single substitution calculus (SSC)
-which is simpler and more minimalistic than the existing algebraic
+which is simpler and more minimalistic than existing algebraic
 approaches. The instantiation operation $\blank[\blank]$ takes either a
@@
 arbitrary model, but they hold in the syntax. In this paper, we show
-that the \emph{syntax} of our single substitution theory and the
-syntax of the corresponding parallel substitution theory (CwF-based
-theory) are isomorphic. All CwF-based models are models of our single
-substitution theory, but not necessarily the other way. While
-B-systems are equivalent to CwFs (and C-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more SSC models
-than CwF-based models. This relationship is similar to that of
-extensional combinatory calculus and lambda calculus, where the
+that the \emph{syntaxes} of our SSC and the corresponding parallel
+substitution calculus (CwF-based calculus) are isomorphic. All
+CwF-based models are models of our SSC, but not
+necessarily the other way. While B-systems are equivalent to CwFs (and
+C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
+SSC models than CwF-based models. This relationship is similar to that
+of extensional combinatory calculus and lambda calculus, where the
 syntaxes are equivalent, and every lambda model is a combinatory
@@
 Our SSC has 8 equations, and these are enough to describe type
-theories with arbitrary choice of type formers (see the description of
+theories with arbitrary choice of type formers (see the paragraph on
 SOGATs in Subsection \ref{sec:related}). In the particular case of
 type theory with a Coquand-universe and $\Pi$ types, we can do better:
-4 (conditional) equations on variables are enough, because using the
-fact that every type is represented by a terma, we can derive the other
-4 equations.
+the 4 (conditional) equations are enough, because we can derive the
+other 4 equations using the fact that every type is represented by a
+term.
 
@@
 represent contexts, functions between $\Sigma$ types represent
-parallel substitutions, functions into the universe represent depenent
-types. Then the SSC is just there to formulate $\Sigma$, $\Pi$, $\U$,
-and then we use these to define a parallel substitution calculus. So
+parallel substitutions, functions into the universe represent
+dependent types. Then the SSC is just there for bootstrapping
+purposes: we use the SSC to formulate $\Sigma$, $\Pi$, $\U$, and then
+we use these to define a parallel substitution calculus. In summary,
 although SSC models and CwFs without any type formers are not
@@
 
-All of our constructions can be understood as happening internal to
-extenional type theory, and most results in this paper are formalised
-in Agda. Most results in this paper were presented at the TYPES 2024
+All of the constructions in this paper can be understood as happening
+internal to extensional type theory, and most results were formalised
+in Agda. Some results in this paper were presented at the TYPES 2024
 conference \cite{singleTypes}.
 
+In summary, our contributions are the following:
 \begin{itemize}
 \item A new generalised algebraic presentation of type theory in the
-  form of a minimalistic single substitution calculus. We show that
-  there is no need for parallel substitutions, empty substitution,
-  parallel weakenings, telescopes, or combinations of these to
-  describe the syntax of type theory in an intrinsic way. We present
-  our calculus in an easy-to-understand way.
-\item The $\alpha$-normalisation technique and deriving parallel
-  substitutions from the single substitution syntax.
+  form of a minimalistic single substitution calculus. Our calculus
+  does not feature parallel substitutions/weakenings, empty
+  substitution, telescopes. We present our calculus in an
+  easy-to-understand way where all operations are well-motivated.
+\item The $\alpha$-normalisation technique which shows that the syntax
+  of our single substitution calculus is isomorphic to the syntax of
+  the CwF-based theory.
 \item For type theory with $\Pi$, $\U$, a minimised presentation of
@@
 
-After discussing related work, in Section \ref{sec:tt}, we introduce
-our single substitution calculus (SSC) only relying on a working
-knowledge of an implementation of type theory such as Agda, no prior
-experience with the metatheory of type theory is required. Section
-\ref{sec:admissible} shows that in the syntax of SSC, several new
-equations are admissible, and that the syntax of SSC is actually
-isomorphic to the CwF-based syntax. In Section \ref{sec:minimisation}
-we minimise our SSC obtaining a theory with fewer equations, relying
-on the presence of $\Pi$ and $\U$ in our theory. In Section
-\ref{sec:cwf}, we show that in the presence of $\Pi$, $\U$ and
-$\Sigma$, SSC-models are actually equivalent to CwF-models in a weak
-sense. We conclude in Section \ref{sec:conclusion}.
-
-\subsection{Related work}
+After discussing related work and our metatheory, in Section
+\ref{sec:tt}, we introduce our single substitution calculus (SSC) only
+relying on a working knowledge of an implementation of type theory
+such as Agda, no prior experience with the metatheory of type theory
+is required. Section \ref{sec:admissible} shows that in the syntax of
+SSC, several new equations are admissible, and that the syntax of SSC
+is actually isomorphic to the CwF-based syntax. In Section
+\ref{sec:minimisation}, we minimise our SSC obtaining a theory with
+fewer equations, relying on the presence of $\Pi$ and $\U$ in our
+theory. In Section \ref{sec:cwf}, we show that in the presence of
+$\Pi$, $\U$, $\Sigma$, SSC-models are actually equivalent to
+CwF-models in a weak sense. We conclude in Section
+\ref{sec:conclusion}.
+
+\subsection{Related work}\label{sec:related}
 
@@
 algebraic presentation of type theory using single
-substitutions. B-systems correspond to our SSC as essentially
-algebraic theories correspond to generalised algebraic theories, or
-sets with a map into $I$ correspond to indexed families over $I$
+substitutions. B-systems relate to our SSC as essentially
+algebraic theories relate to generalised algebraic theories, or
+sets with a map into $I$ relate to indexed families over $I$
 \cite[page 221]{DBLP:journals/apal/Cartmell86}. Another difference is
@@
 $\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
-order. However we don't have the lifted versions of our equations,
+order. However we don't include the lifted versions of these equations,
 equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
@@
 parallel substitution calculus with $\p$, $\blank^+$ and
-$\langle\blank\rangle$ operations, just like in our calculus. Our
+$\langle\blank\rangle$ operations, just like our calculus. Our
 single substitution calculus is essentially Ehrhard's calculus with
@@
 relation. For formalising the GAT-level syntax, one needs a stronger
-metatheory than ordinary Agda or Coq: the metatheory has to support
+metatheory than plain Agda or Coq: the metatheory has to support
 quotient inductive-inductive types (QIITs
@@
 \cite{initiality-agda} constructed the initial contextual category in
-Agda using postulated quotients. Although there is no general proof
-published about this, we have experimental evidence that QIITs are
-supported by the cubical set model \cite{DBLP:conf/lics/CoquandHM18}
-and the setoid model \cite{kaposi-qiit-setoid}, thus Cubical Agda
+Agda using postulated quotients. Although we don't have a general
+proof, we have experimental evidence that QIITs are supported by the
+cubical set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid
+model \cite{kaposi-qiit-setoid}, thus Cubical Agda
 \cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
@@
 \cite{uemura,DBLP:conf/fscd/KaposiX24}) can be used to present type
-theory even more abstractly than GATs: they even abstract over the way
-of defining the substitution calculus: contexts, substitutions,
-instantiations are simply modelled by the metatheoretic function
-space. For example, a type theory with $\Pi$, Coquand-universes and a
-universe level lifting operation is described by the following SOGAT
-(where $\cong$ means isomorphism):
+theory more abstractly than GATs: SOGATs abstract over the exact
+definition of the substitution calculus (whether substitutions are
+single or parallel; whether we have CwF-style or Ehrhard-style
+operations): contexts, substitutions, instantiations are simply
+modelled by the metatheoretic function space. For example, a type
+theory with $\Pi$, Coquand-universes and a universe level lifting
+operation is described by the following SOGAT (where $\cong$ means
+isomorphism):
 \begin{equation}\label{eq:tt}
@@
 % \end{equation}
-Second-order models of SOGATs are not well-behaved as there is no good
-notion of homomorphism between them. This is why Kaposi and Xie
-\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: each GAT
-comes with a category of models with initial model (syntax), (co)free
-models, and so on. Actually, Kaposi and Xie define two different
-translations: one based on parallel substitutions and one based on
-single substitutions. If we apply their parallel translation to the
-SOGAT
+Second-order models of SOGATs are not well-behaved, as there is no
+good notion of homomorphism between them. This is why Kaposi and Xie
+\cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
+SOGAT then is a model of the GAT the we obtain by their
+translation. GATs come with well-behaved metatheory: a category of
+models witha initial model (syntax), (co)free models
+\cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
+translation is not unique: Kaposi and Xie
+\cite{DBLP:conf/fscd/KaposiX24} define two different translations, one
+based on parallel substitutions and one based on single
+substitutions. If we apply their parallel translation to the SOGAT
 \begin{equation}\label{eq:tytm}
@@
 of category with families (CwF). If we use their single substitution
-translation on the SOGAT (\ref{eq:tt}), we obtain Definition
-\ref{def:tt} (also listed in Appendix \ref{app:tt}). This paper can be
-seen as taking a big enough type theory (described as a SOGAT), and
-investigating how the result of its single substitution translation
-relates to the result of its parallel translation (the latter
-first-order theory is quite well-known).
+translation to the SOGAT (\ref{eq:tt}), we obtain Definition
+\ref{def:tt} (also listed in Appendix \ref{app:tt}). The current paper
+can be seen as taking a big enough type theory (described as a SOGAT),
+and investigating how the result of its single substitution
+translation relates to the result of its parallel translation (the
+latter first-order theory is quite well-known). There is no general
+result about the relationship of the two translations.
 
@@
 SOGAT level of abstraction: Synthethic Tait Computability
-\cite{DBLP:phd/us/Sterling22} or internal sconing
+\cite{DBLP:phd/us/Sterling22} and internal sconing
 \cite{DBLP:conf/fscd/BocquetKS23}.

### 2025-01-30 cpp/p.tex
stuff

@@
 forms don't include explicit substitutions. $\alpha$-normal forms are
-not like ordinary normal forms because they are still quotiented by
-the computation/uniqueness rules like $\beta$/$\eta$ for function
-space. We use induction on $\alpha$-normal forms to define parallel
-substitutions and prove all the rules of CwFs in the syntax of SSC.
+not like ordinary $\beta/\eta$-normal forms because they are still
+quotiented by the computation/uniqueness rules like $\beta$/$\eta$ for
+function space. We use induction on $\alpha$-normal forms to define
+parallel substitutions and prove all the rules of CwFs in the syntax
+of SSC.
 
@@
 
-\subsection{Metatheory and formalisation}
+\subsection{Metatheory and formalisation}\label{sec:metatheory}
 
@@
 \ref{sec:admissible} of this paper are formalised in the proof
-assistant Agda, we use a strict Prop-valued
-\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
-coercion rule (transport rule) and postulated QIITs with computation
-rules added using rewrite rules
+assistant Agda. The formalisation is availble at TODO. We use a strict
+Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
+with postulated coercion rule (transport rule) and postulated QIITs
+with computation rules added using rewrite rules
 \cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
@@
 overloading extensively, for example instantiation $\blank[\blank]$ is
-overloaded for types and terms.
+overloaded for types and terms. For isomorphisms, we use the notation
+$f:X\cong Y:g$ meaning $f:X\ra Y$ and $g :Y\ra X$ together with a
+$\beta$ equality $f\,(g\,y) = y$ for all $y$, and an $\eta$ equality
+saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
+and $g$ as a constructor).
 
@@
 the uniqueness rule again needs an extra equation on the codomain type
-of the function to make sense. This is the typal version of the
+of the function to make sense. This is the type version of the
 equation mentioned in the introduction (we don't have its term version).
@@
 \end{alignat*}
+Note that we don't have definitional commutation of type formers and
+$\Lift$ such as $\Lift\,(A\Ra B) = (\Lift\,A\Ra\Lift\,B)$. It however
+holds as a definitional isomorphism (see Subsection
+\ref{sec:examples}). This adds extra administration when using our
+theory, but makes stating our theory simpler, and includes more
+models. The situation is similar to the correspondance between
+$\Ty\,\Gamma\,i$ and $\Tm\,\Gamma\,(\U\,i)$ which could have been
+stated as an equality (making universes {\`a} la Russell).
+
 \begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
@@
 structural (substitution calculus) rules. For example, we show how to
-extend Definition \ref{def:tt} with the SOGAT-rules (\ref{eq:sigma}).
+extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
 \begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$, $\top$, and $\Sigma$ types]\label{def:sigma}
@@
   To save space, we compressed the introduction, elimination, $\beta$
-  and $\eta$ rules for $\Sigma$ into one isomorphism, and did not list
-  substitution laws for $\fst$ and $\snd$ as they are derivable.
+  and $\eta$ rules of $\Sigma$ into one isomorphism, and did not list
+  substitution laws for $\fst$ and $\snd$ as they are derivable (see
+  the examples below).
 \end{definition}
 
-\subsection{Examples}
+\subsection{Examples}\label{sec:examples}
 
@@
 $\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
-this is why we had to lift $\El\,\q$. Another subtle thing is
+this is why we had to lift $\El\,\q$. Another subtlety is
 happening when deriving the third implicit argument of $\q$. We have
@@
 $\blank[\p]$ for each type in the context.
+
+The operations for $\U$ and $\Lift$ can be written more consicely as
+the following isomorphisms:
+\[
+\El:\Tm\,\Gamma\,(\U\,i) \cong \Ty\,\Gamma\,i:\cd \hspace{5em} \un:\Tm\,\Gamma\,(\Lift\,A)\cong\Tm\,\Gamma\,A:\mk
+\]
+The $\U\beta$, $\U\eta$ and $\Lift\beta$, $\Lift\eta$ rules express
+that the roundtrips are identities. These isomorphisms are also
+\emph{natural}, or compatible with substitutions: these are expressed
+by the equations $\El[]$, $\un[]$, $\cd[]$, $\mk[]$.
+
+$\Sigma$ types were introduced by a similar isomorphism, and we give
+names to the roundtrips according to our convention introduced in
+Section \ref{sec:metatheory}. xxx
+
+but
+naturality was only stated for the right to left direction (rule
+${,}[]$). Naturalities in the direction are derivable:
+\begin{alignat*}{10}
+  & (\fst\,w)[\gamma] && {=}(xxx) \\
+  & \fst\,((\fst\,w)[\gamma],(\snd\,w)[\gamma]) = \\
+  & \fst\,((\fst\,w,\snd\,w)[\gamma]) = \\
+  & \fst\,(w[\gamma])
+\end{alignat*}
+
+We show the commutation of $\Lift$ and $\Pi$ types as a sequence of
+isomorphisms:
+\begin{alignat*}{10}
+  & \Tm\,\Gamma\,(\Lift\,(\Pi\,A\,B)) && {\cong}\text{(1)} \\
+  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{(2)} \\
+  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{(3)} \\
+  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(4)} \\
+  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{(5)} \\
+  & \Tm\,\Gamma\,(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]))
+\end{alignat*}
+Steps (1) and (3) are the isomorphism comprised of $\mk$ and $\un$.
 

### 2025-01-30 cpp/p.tex
stuff

@@
 translation. GATs come with well-behaved metatheory: a category of
-models witha initial model (syntax), (co)free models
+models with initial model (syntax), (co)free models
 \cite{andras,DBLP:phd/hal/Moeneclaey22}, and so on. However, the
@@
 with only two sorts and no operations or equations, we obtain the GAT
-of category with families (CwF). If we use their single substitution
+of category with families (CwF). If we apply their single substitution
 translation to the SOGAT (\ref{eq:tt}), we obtain Definition
@@
 programs. We give a tutorial-style introduction to the syntax, we do
-not assume prior knowledge of the metatheory of type theory. The
-definition of type theory that we obtain is the same as the single
-substitution GAT semantics \cite{DBLP:conf/fscd/KaposiX24} of the
-SOGAT (\ref{eq:tt}).
+not assume prior knowledge of the metatheory of type theory.
 
@@
 \subsection{Examples}\label{sec:examples}
+
+In this subsection, we show how to use the above defined calculus, and
+motivate the next section on admissible equations.
 
@@
 and $a'$ just as before. What we cannot do is to directly weaken a
-term in context $\diamond$ to an arbitrary metavariable context
-$\Gamma$, we can only do this step by step by applying weakening
-$\blank[\p]$ for each type in the context.
+term in context $\diamond$ to an arbitrary context $\Gamma$, where
+$\Gamma$ is a metavariable. If we know that $\Gamma$ has lengh $n$,
+then we can do the weakening by
+$\blank\underbrace{[\p][\p]\dots[\p]}_{\text{$n$ times}}$. If we work
+in the syntax, then we can derive such weakenings (and more) using the
+methods of Section \ref{sec:admissible}.
 
@@
 $\Sigma$ types were introduced by a similar isomorphism, and we give
-names to the roundtrips according to our convention introduced in
-Section \ref{sec:metatheory}. xxx
-
-but
-naturality was only stated for the right to left direction (rule
-${,}[]$). Naturalities in the direction are derivable:
-\begin{alignat*}{10}
-  & (\fst\,w)[\gamma] && {=}(xxx) \\
-  & \fst\,((\fst\,w)[\gamma],(\snd\,w)[\gamma]) = \\
-  & \fst\,((\fst\,w,\snd\,w)[\gamma]) = \\
-  & \fst\,(w[\gamma])
-\end{alignat*}
-
-We show the commutation of $\Lift$ and $\Pi$ types as a sequence of
-isomorphisms:
-\begin{alignat*}{10}
-  & \Tm\,\Gamma\,(\Lift\,(\Pi\,A\,B)) && {\cong}\text{(1)} \\
-  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{(2)} \\
-  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{(3)} \\
-  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(4)} \\
-  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{(5)} \\
-  & \Tm\,\Gamma\,(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]))
-\end{alignat*}
-Steps (1) and (3) are the isomorphism comprised of $\mk$ and $\un$.
+names to the roundtrips eqautions according to our convention introduced in
+Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
+$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
+(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
+direction (${,}[]$), because naturalities in the other direction are
+derivable:
+\begin{alignat*}{10}
+  & (\fst\,w)[\gamma] && {=}(\Sigma\beta_1)                             && (\snd\,w)[\gamma] && {=}(\Sigma\beta_2)                             \\
+  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{4em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
+  & \fst\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        && \snd\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        \\
+  & \fst\,(w[\gamma])                                                && && \snd\,(w[\gamma])                                                   
+\end{alignat*}
+For $\Pi$ types, we construct a similar isomorphism:
+\[
+(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam
+\]
+The roundtrips are proven as follows:
+\[
+\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t
+\]
+The last step is the admissible equation (\ref{eq:lifted4}) proven in
+Section \ref{sec:admissible}. As above with weakening into an
+arbitrary context, this equation holds in the syntax, bot not in an
+arbitrary model.
+%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
+The $\eta$ rountrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
+
+We show another isomorphism between terms depending on a variable and
+terms depending on the lifted version of the same variable:
+\[
+(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
+\]
+To derive the roundtrips, we again need admissible equations. Here is
+the proof of the first roundtrip equality, the other is analogous:
+\begin{alignat*}{10}
+  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[]) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+]) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[]) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
+  & t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
+  & t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
+  & t
+\end{alignat*}
+Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
+\begin{alignat*}{10}
+  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) && {\cong}\text{($\Lift$ iso)} \\
+  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{($\Pi$ iso)} \\
+  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{($\Lift$ iso)} \\
+  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(lifted variable iso)} \\
+  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{($\Pi$ iso)} \\
+  & \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
+\end{alignat*}
+We hope that the above examples demonstrated the for equations
+(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
+section.
 
@@
 (metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
-SSC-based calculus calculus in Agda. We used the techniques of Kovács
+SSC-based calculus in Agda. We used the techniques of Kovács
 \cite{DBLP:conf/csl/Kovacs22} to define inductive-recursive universes
@@
 In this section we show that our SSC-based syntax is isomorphic to the
-CwF-based syntax for the same theory (namely the SOGAT
-(\ref{eq:tt})). We conjecture that the method presented here works
-generically for any SOGAT.
+CwF-based syntax with the same type formers.
+\footnote{Using the two translations from SOGATs to GATs
+\cite{DBLP:conf/fscd/KaposiX24}, we can rephrase the contents of this
+section as follows: the syntaxes of the single and parallel
+substitution first order versions of the SOGAT \ref{eq:tt} are
+isomorphic.}
 
@@
 \ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
-\begin{equation}\label{eq:lifted}
-\begin{aligned}
-  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \\
-  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \\
-  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \\
-  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B
-\end{aligned}
-\end{equation}
+\begin{alignat}{10}
+  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \label{eq:lifted1}\\
+  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2}\\
+  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3}\\
+  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B\label{eq:lifted4}
+\end{alignat}
 The corresponding equations on terms are also needed. Note that we do not have
@@
 
-To formally state the equations in (\ref{eq:lifted}), we define telescopes as an
+To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we define telescopes as an
 inductive type together with a recursive operation to append telescopes to
@@
 
-If we try to prove the equations in (\ref{eq:lifted}) by induction on the types
+If we try to prove the equations in (\ref{eq:lifted1})--(\ref{eq:lifted4}) by induction on the types
 and terms, we run into difficulties in the instantiation case, as we would have
@@
 
-Instead of doing induction on $\alpha$-normal forms for each of the equations in
-(\ref{eq:lifted}), we define a general lemma which can lift any equation between
+Instead of doing induction on $\alpha$-normal forms for each of equations 
+(\ref{eq:lifted1})--(\ref{eq:lifted4}), we define a general lemma which can lift any equation between
 instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with

### 2025-01-30 cpp/p.tex
stuff

@@
 we show that SSC models are actually equivalent to CwF-based models in
-a weak sense. Most results in this paper were formalised in Agda.
+a weak sense. Sections 2 and 3 of this paper were formalised in Agda.
 \end{abstract}
@@
 
-Some advantages of using a GAT presentation:
+The GAT presentation has the following advantages:
 \begin{itemize}
@@
   \node (c1) at (2.5,1) {$\tilde{B}_1$};
-  \node (c2) at (5,1) {$\tilde{B}_1$};
+  \node (c2) at (5,1) {$\tilde{B}_2$};
   \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
@@
 \end{tikzpicture} \\
-Using our notation, a B-frame contains the following data. \\
+Using our notation, a B-frame contains the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
 \begin{tikzpicture}
@@
 typed CwFs with function space. Their equivalence result holds even
-without the presence of function space. In the presence of function
+without the presence of function types. In the presence of function
 space, the same equivalence holds for the simply typed version of our
@@
 typing and conversion relations. A higher level representation is
-working in setoid hell \cite{chapman09eatitself}: terms are
+working in setoid hell \cite{chapman09eatitself,setoidhell}: terms are
 intrinsically well-typed, but conversion is still an explicit
@@
 Second-order models of SOGATs are not well-behaved, as there is no
-good notion of homomorphism between them. This is why Kaposi and Xie
+good notion of homomorphism between them \cite[bottom of page 5]{DBLP:conf/fscd/KaposiX24}. This is why Kaposi and Xie
 \cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
@@
 context. We would like to describe the operation providing the last
-variable in a context, the zero De Bruijn index \cite{debruijn}, we denote it by $\q :
-\Tm\,(\Gamma\ext A)\,A'$. It is a term in an extended context, and it
-should have type $A$, but the issue is that $A : \Ty\,\Gamma$, and $A'
-: \Ty\,(\Gamma\ext A)$. We need a way to weaken the type $A$ to obtain
-such an $A'$. For this, we will introduce an operation $\blank[\p] :
-\Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext A)\,i$ and define $A' := A[\p]$. Instead
-of introducing just $\blank[\p]$, we generalise a bit and add a new
-sort $\Sub$ which for now only has the single element $\p$ and the
-operation $\blank[\blank]$ is called \emph{instantiation}. We
-instantiate the type with the $\Sub\,\Delta\,\Gamma$ weakening.
-Elements of $\Sub$ will be later called
-substitutions, hence the name, but for now, we only have single end-of
-context weakenings in $\Sub$. $\Delta$ is called the domain, $\Gamma$
-the codomain of a $\Sub\,\Delta\,\Gamma$.
+variable in a context, the zero De Bruijn index \cite{debruijn}, we
+denote it by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
+extended context, and it should have type $A$, but the issue is that
+$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
+weaken the type $A$ to obtain such an $A'$. For this, we will
+introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
+A)\,i$ and define $A' := A[\p]$. Instead of introducing just
+$\blank[\p]$, we generalise a bit and add a new sort $\Sub$ which for
+now only has the single element $\p$ and the operation
+$\blank[\blank]$ is called \emph{instantiation}. We instantiate the
+type with the $\Sub\,\Delta\,\Gamma$ weakening.  Elements of $\Sub$
+will later be called substitutions, hence the name, but for now, we
+only have single end-of context weakenings in $\Sub$. The context
+$\Delta$ is called the domain, $\Gamma$ the codomain of a
+$\Sub\,\Delta\,\Gamma$.
 \begin{alignat*}{10}
@@
 the uniqueness rule again needs an extra equation on the codomain type
-of the function to make sense. This is the type version of the
-equation mentioned in the introduction (we don't have its term version).
+of the function to make sense (we mentioned the term version of this equation in the introduction).
 \begin{alignat*}{10}
@@
 equipped with a substitution rule. There is no need to add more
-structural (substitution calculus) rules. For example, we show how to
+structural (substitution calculus) rules.
+\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
+these equations are enough to describe the first-order theory of any SOGAT.}
+For example, we show how to
 extend Definition \ref{def:tt} with $\top$ and $\Sigma$ types.
@@
 $\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
-this is why we had to lift $\El\,\q$. Another subtlety is
-happening when deriving the third implicit argument of $\q$. We have
-to coerce $\q$ along the equality $\U[]$.
+which is why we had to lift $\El\,\q$. Another subtlety is
+happening when deriving the third implicit argument of $\q$, because we have
+to coerce $\q$ along the equality $\U[]$:
 \[
@@
   & (\lam\,\q[\langle\cd\,A\rangle])\cdot a && {=}(\lam[]) \\
-  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a && {=}([\q][^+]) \\
+  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a\, && {=}([\q][^+]) \\
   & (\lam\,\q)\cdot a && {=}(\Pi\beta) \\
@@
 \]
-(with different implicit arguments), so we are able apply $\cd\,A'$
+(with different implicit arguments for the $\lam$s on the left and right hand side), so we are able to apply $\cd\,A'$
 and $a'$ just as before. What we cannot do is to directly weaken a
@@
 
-As a sanity check for our notion of SSC-model, we defined the standard
-(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
-SSC-based calculus in Agda. We used the techniques of Kovács
-\cite{DBLP:conf/csl/Kovacs22} to define inductive-recursive universes
-to model the infinite hierarchy of types, and all the SSC-equations
-hold by reflexivity in this model. We defined the analogous CwF-based
-standard model.
+As a sanity check for our notion of SSC-model, we defined a variant of
+the standard (metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model
+of the SSC-based calculus in Agda. For convenience, we used an
+inductive-recursive universe to define the infinite hierarchy of types
+\cite{DBLP:journals/jsyml/Dybjer00,DBLP:conf/csl/Kovacs22}. All
+SSC-equations hold by reflexivity in this model. We defined the
+analogous CwF-based standard model.
 
@@
 \end{definition}
+In the rest of this section, we work with the SSC syntax denoted
+$\Con$, $\Sub$, $\Ty$, $\Tm$, and so on, towards the goal of defining
+all CwF operations and equations.
 
@@
   & \diamond && : \Tel\,\Gamma \\
-  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Omega
+  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma
 \end{alignat*}
@@
 $\alpha$-normal (except variables). It is important to truncate the
-$\alpha$-normal predicate to be propositions to ensure that it preserves the
+$\alpha$-normal predicate to be a proposition to ensure that it preserves the
 $\beta$/$\eta$ conversion rules.
@@
   \label{thm:lifting}
-  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if $A[\gamma_0] =
-  A[\gamma_1]$ and $x[\gamma_0] = x[\gamma_1]$ for any $A:\Ty\,\Gamma$ and
-  variable $x:\Tm\,\Gamma\,A$, then
+  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if
+  \begin{itemize}
+  \item for any $A:\Ty\,\Gamma$, $A[\gamma_0] = A[\gamma_1]$, and
+  \item for any variable $x:\Tm\,\Gamma\,A$, $x[\gamma_0] = x[\gamma_1]$,
+  \end{itemize}
+  then:
   \begin{itemize}
@@
 substitutions, however it does not satisfy the equations of parallel
-substitutions. Thus we define parallel substitutions as a list of terms with a
+substitutions. Thus we define parallel substitutions as lists of terms with a
 map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to convert it
@@
 
-\begin{theorem}
+\begin{problem}
   Contexts, types, and terms in the SSC syntax are isomorphic to the
@@
   substitutions.
-\end{theorem}
-\begin{proof}\leavevmode
+\end{problem}
+\begin{proof}[Construction]\leavevmode
   \begin{description}
@@
 \end{proof}
+This also implies that the initial SSC is the initial CwF.
 
@@
 
-In this section we show that if have we universes and $\Pi$ types
+In this section we show that if have universes and $\Pi$ types
 (which is the case in the theory (\ref{eq:tt})), we can decrease the
@@
 In our single substitution calculus, we have equation $[\p][^+]$
-stated both for types and terms (we know that the left one is about
-types because of our convention that type metavariables are uppercase
-Latin letters and term metavariables are lowercase):
+stated both for types and terms (here $B$ is a type and $b$ is a term of type $B$):
 \[
@@
 That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
-we have that for any $b$ we have the equation (suitably over the input
-equation). Thus this is a conditional equation.
+we have the equation (suitably over the input
+equation) for any $b$. Thus this is a conditional equation.
 
@@
 We were forced to introduce it so that we can state the substitution
-rule for function application $\cdot[]$. What if we made $\cdot[]$
-itself conditional?
+rule for function application $\cdot[]$. A conditional version of
+$\cdot[]$ looks as follows.
 \begin{alignat*}{10}
@@
 \end{alignat*}
-Now we derive the assumption of $\Pi\beta/\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
+Now we derive the assumption of $\Pi\beta'/\Pi\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
 \begin{alignat*}{10}
@@
 \end{definition}
-\begin{theorem}
+\begin{problem}
   The GATs of Definition \ref{def:min} and Definition \ref{def:tt} are
   isomorphic, in particular, all equations are interderivable.
-\end{theorem}
-\begin{proof}
+\end{problem}
+\begin{proof}[Construction]
   Clear from the construction in this section.
@@
 \begin{problem}
-  Every CwF (with type formers $\Pi, \dots, \Sigma$) is an SSC (with
+  Every CwF (with type formers $\Pi, \dots, \Sigma$) gives rise to an SSC (with
   the same type formers, see Definition \ref{def:sigma}).
@@
 CwF-morphism (pseudomorphism) which is bijective on types and
-terms. Note that a conextual isomorphism preserves all type formers
+terms. Note that a contextual isomorphism preserves all type formers
 specified by universal properties.
@@
   Given a CwF (with type formers) $M$, let $M'$ denote the CwF
-  obtained by first seeing it as an SSC and then termifying it. There
-  is a contextual isomorphism between $M'$ and $M$.
+  obtained by first seeing it as an SSC and then termifying it. We
+  construct a contextual isomorphism between $M'$ and $M$.
 \end{problem}
@@
   It is easy to check that $F$ preserves the CwF structure, i.e.\ it
-  is functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
+  is a functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
   isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
@@
   4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
-syntax of type theory can be defined using infinitely many coherene
+syntax of type theory can be defined using infinitely many coherence
 rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
@@
 set, hence they don't have decidable equality \cite{hedberg}. The
-solution is to add the pentagon law, but then this introduces new
+solution is to add the pentagon law stating that these two equalities are equal, but then this introduces new
 higher equalities, which require more coherence laws, so in the end we

### 2025-01-30 cpp/p.tex
lipics

@@
-\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\title{Type theory in type theory with single substitutions}
+
+\author{Ambrus Kaposi}{Eötvös Loránd University (ELTE), Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{}
+
+\author{Szumi Xie}{Eötvös Loránd University (ELTE), Hungary}{szumi@inf.elte.hu}{https://orcid.org/0009-0001-1355-1114}{}
+
+\authorrunning{A. Kaposi and Sz. Xie}
+
+\Copyright{Ambrus Kaposi and Szumi Xie}
+
+\ccsdesc[500]{Theory of computation~Type theory}
+
+\keywords{type theory, category with families, substitution calculus}
+
+%\acknowledgements{I want to thank \dots}%optional
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\maketitle
+
 \newtheorem{problem}[theorem]{Problem}
@@
 \newtheorem{construction}[theorem]{Construction}
-
-\title{Type theory in type theory with single substitutions}
-\author{Ambrus Kaposi and Szumi Xie}
-\date{\today}
 
@@
 \end{abstract}
-
-\maketitle
 
@@
 
-\paragraph{B-systems} B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
-an algebraic way of describing type theories close to the notations
-using typing judgements. B-systems are an intrinsic, essentially
+\subparagraph*{B-systems.} B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
+an algebraic description of type theory close to extrinsic definitions.
+B-systems are an intrinsic, essentially
 algebraic presentation of type theory using single
@@
 \begin{tikzpicture}
-  \node (b0) at (0,0) {$\top$};
-  \node (b1) at (1.5,0) {$B_1$};
-  \node (b2) at (4,0) {$B_2$};
-  \node (b3) at (7,0) {$\dots$};
-  \node (c1) at (2.5,1) {$\tilde{B}_1$};
-  \node (c2) at (5,1) {$\tilde{B}_2$};
+  \node (b0) at (4,0) {$\top$};
+  \node (b1) at (5,0) {$B_1$};
+  \node (b2) at (12,0) {$B_2$};
+  \node (b3) at (15,0) {$\dots$};
+  \node (c1) at (7,1) {$\tilde{B}_1$};
+  \node (c2) at (14,1) {$\tilde{B}_2$};
   \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
@@
 \end{tikzpicture} \\
-Using our notation, a B-frame contains the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
+Using our notation, this is the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
 \begin{tikzpicture}
-  \node (b0) at (0,0) {$\top$};
-  \node (b1) at (1.5,0) {$\Ty\,\diamond$};
-  \node (b2) at (4,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
-  \node (b3) at (7,0) {$\dots$};
-  \node (c1) at (2.5,1) [text width=3cm] {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
-  \node (c2) at (5.7,1.7) [text width=4cm,align=center] {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
+  \node (b0) at (4,0) {$\top$};
+  \node (b1) at (5,0) {$\Ty\,\diamond$};
+  \node (b2) at (12,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
+  \node (b3) at (15,0) {$\dots$};
+  \node (c1) at (7,1)  {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
+  \node (c2) at (14,1) {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
   \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
@@
 \end{tikzpicture} \\
-The substitution operation $\mathbb{S}$ for an element of
-$x:\tilde{B}_{n+1}$ is a homomorphism of the slice B-frames
-$\mathbb{B}/\partial(x) \ra \mathbb{B}/\mathsf{ft}(\partial(x))$. In
-our notation, $x = (\Gamma, A, a)$ where $a : \Tm\,\Gamma\,A$, and the
-homomorphism $\mathbb{S}$ corresponds to the following maps for $\Ty$
-(mapping between the bottom rows of the diagrams):
+The substitution operation $\mathbb{S}$ for an $x:\tilde{B}_{n+1}$ is
+a homomorphism of the slice B-frames $\mathbb{B}/\partial(x) \ra
+\mathbb{B}/\mathsf{ft}(\partial(x))$. In our notation, $x = (\Gamma,
+A, a)$ where $a : \Tm\,\Gamma\,A$, and $\mathbb{S}$ corresponds to the
+following functions for $\Ty$ (a map between the bottom rows of the
+diagrams):
 \begin{alignat*}{10}
@@
 for terms. The missing equations are admissible, see Section
-\ref{sec:admissible}. In the presence of universes and $\Pi$ types, we
+\ref{sec:admissible}. In the presence of $\U$ and $\Pi$, we
 reduce the needed equations even more, see Section
@@
 
-\paragraph{Ehrhard's calculus}
+\subparagraph*{Ehrhard's calculus.}
 The first generalised algebraic presentation of type theory was Thomas
@@
 
-\paragraph{Formalisations of type theory}
+\subparagraph*{Formalisations of type theory.}
 Most computer formalisations of type theory are extrinsic
@@
 
-\paragraph{SOGATs}
+\subparagraph*{SOGATs}
 We can define languages with binders as second-order theories
@@
 \begin{alignedat}{10}
-  & \Ty && : \N\ra\Set \\
-  & \Tm && : \Ty\,i\ra\Set \\
-  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
-  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
-  & \U && : (i:\N)\ra\Ty\,(1+i) \\
-  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
-  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
-  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
+  & \Ty && : \N\ra\Set                                                             && \U && : (i:\N)\ra\Ty\,(1+i) \\           
+  & \Tm && : \Ty\,i\ra\Set                                                         && \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
+  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i                               && \Lift && : \Ty\,i\ra\Ty\,(1+i) \\        
+  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \hspace{3em} && \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
 \end{alignedat}
@@
 \begin{alignedat}{10}
-  & \Ty && : \Set \\
-  & \Tm && : \Ty \ra\Set
+  & \Ty && : \Set \hspace{3em} \Tm && : \Ty \ra\Set
 \end{alignedat}
@@
 translation relates to the result of its parallel translation (the
-latter first-order theory is quite well-known). There is no general
-result about the relationship of the two translations.
+latter first-order theory is quite well-known).
+%There is no general result about the relationship of the two translations.
 
@@
 an implicit argument of $\Tm$.
-\begin{alignat*}{10}
-& \Con && : \Set \\
-& \Ty && : \Con\ra\N\ra\Set \\
-& \Tm && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
-\end{alignat*}
+\[
+\Con : \Set \hspace{3em} 
+\Ty : \Con\ra\N\ra\Set \hspace{3em} 
+\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
+\]
 Just as lists have two constructors, there are two ways to form
@@
 type which can have variables in the context preceeding the type.
-\begin{alignat*}{10}
-& \diamond && : \Con \\
-& \blank\ext\blank && : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
-\end{alignat*}
+\[
+\diamond : \Con \hspace{3em}
+\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
+\]
 We will refer to variables by their distance from the end of the
@@
 \begin{alignat*}{10}
-& \Sub && : \Con\ra\Con\ra\Set \\
-& \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
-& \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \\
-& \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
+  & \Sub && : \Con\ra\Con\ra\Set && \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
+  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{3em} && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
 \end{alignat*}
@@
 elements of $\Sub$ have the form
-\begin{alignat*}{10}
-  & \p^{+^n} : \Sub\, && (\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}]) \\
-  & && (\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
-\end{alignat*}
+\[
+\p^{+^n} : \Sub\,(\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}])\,(\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
+\]
 where $+^n$ means the $n$-times iteration of $\blank^+$.
@@
 equations.
-\begin{alignat*}{10}
-& [\p][^+] && : B[\p][\gamma^+] = B[\gamma][\p] \\
-& [\p][^+] && : b[\p][\gamma^+] = b[\gamma][\p] \\
-& \q[^+] && : \q[\gamma^+] = \q
-\end{alignat*}
+\[
+[\p][^+] : B[\p][\gamma^+] = B[\gamma][\p] \hspace{3em}
+[\p][^+] : b[\p][\gamma^+] = b[\gamma][\p] \hspace{3em}
+\q[^+] : \q[\gamma^+] = \q
+\]
 The second and third equations only make sense because of the first
-equation. The phenomenon that later equations (or operations) depend
-on previous equations is usual in the world of GATs. In fact, the
-later equations cannot even be stated without having some previous
-equations.
+equation. The phenomenon that later equations/operations depend on
+previous equations is common in GATs.
 
@@
 rule for nondependent function space as follows.
-\begin{alignat*}{10}
-  & (A\Ra B)[\gamma] && {=} \\
-  & (\Pi\,A\,(B[\p]))[\gamma] && {=}(\Pi[]) \\
-  & \Pi\,(A[\gamma])\,(B[\p][\gamma^+])\,\, && {=}([\p][^+]) \\
-  & \Pi\,(A[\gamma])\,(B[\gamma][\p]) && {=} \\
-  & (A[\gamma])\Ra(B[\gamma])
-\end{alignat*}
+\[
+(A\Ra B)[\gamma] =
+\Pi\,A\,(B[\p])[\gamma] \overset{\Pi[]}{=}
+\Pi\,(A[\gamma])\,(B[\p][\gamma^+]) \overset{[\p][^+]}{=}
+\Pi\,(A[\gamma])\,(B[\gamma][\p]) =
+A[\gamma]\Ra(B[\gamma])
+\]
 
@@
 for instantiation analogous to $\Pi[]$.
-\begin{alignat*}{10}
-& \lam && : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \\
-& \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
-\end{alignat*}
+\[
+\lam : \Tm\,(\Gamma\ext A)\,B\ra\Tm\,\Gamma\,(\Pi\,A\,B) \hspace{3em}
+\lam[] : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+])
+\]
 Note that the equation $\lam[]$ only makes sense because of the
@@
 $\blank[\blank]$ to substitute the last variable in $B$.
-\begin{alignat*}{10}
-& \langle\blank\rangle && : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \\
-& \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle])
-\end{alignat*}
+\[
+\langle\blank\rangle : \Tm\,\Gamma\,A\ra\Sub\,\Gamma\,(\Gamma\ext A) \hspace{3em}
+\blank\cdot\blank : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle])
+\]
 We could have introduced new, separate sorts and $\blank[\blank]$
@@
 and then substituting the last variable.
-\begin{alignat*}{10}
-& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
-& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
-\end{alignat*}
+\[
+[\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \hspace{2em}
+{\cdot}[] : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
+\]
 Following the introduction of the operation $\langle\blank\rangle$, we
@@
 if we have an equation for types.
-\begin{alignat*}{10}
-& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
-& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
-& \q[\langle\rangle] && : \q[\langle a\rangle] = a
-\end{alignat*}
+\[
+[\p][\langle\rangle] : B[\p][\langle a\rangle] = B \hspace{3em}
+[\p][\langle\rangle] : b[\p ][\langle a\rangle] = b \hspace{3em}
+\q[\langle\rangle] : \q[\langle a\rangle] = a
+\]
 We don't have a separate sort of variables, so the rule
@@
 of the function to make sense (we mentioned the term version of this equation in the introduction).
-\begin{alignat*}{10}
-& \Pi\beta && : \lam\,b\cdot a = b[\langle a\rangle] \\
-& [\p^+][\langle\q\rangle] && : B[\p^+][\langle\q\rangle] = B \\
-& \Pi\eta && : t = \lam\,(t[\p]\cdot\q)
-\end{alignat*}
+\[
+\Pi\beta : \lam\,b\cdot a = b[\langle a\rangle] \hspace{3em}
+[\p^+][\langle\q\rangle] : B[\p^+][\langle\q\rangle] = B \hspace{3em}
+\Pi\eta : t = \lam\,(t[\p]\cdot\q)
+\]
 
@@
 which make an isomorphism between terms of type $\U\,i$ and types of
-level $i$. All three operations come with substitution rules ($\El[]$
-and $\cd[]$ are interderivable, so it would be enough to assume one of
-the two, but we add both to completely align with the algorithm
-\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
-(\ref{eq:tt})).
-\begin{alignat*}{10}
-  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
-  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
-  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
-  & \U\beta && : \El\,(\cd\,A) = A \\
-  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
-\end{alignat*}
-Finally, we add the rules for moving types one level up. Because we
-defined function space only between types at the same level, without
-lifting, we cannot even define the polymorphic identity function.
-\begin{alignat*}{10}
-  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
-  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
-  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
-  & \Lift\beta && : \un\,(\mk\,a) = a \\
-  & \Lift\eta && : \mk\,(\un\,a) = a
+level $i$. All three operations come with substitution rules.
+% ($\El[]$
+% and $\cd[]$ are interderivable, so it would be enough to assume one of
+% the two, but we add both to completely align with the algorithm
+% \cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
+% (\ref{eq:tt})).
+\begin{alignat*}{10}
+  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && \U[] && : (\U\,i)[\gamma] = \U\,i                           && \U\beta && : \El\,(\cd\,A) = A \\        
+  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{3em} && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
+  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
+\end{alignat*}
+Finally, we add the rules for moving types one level up. Because $\Pi$
+needs two types at the same level, without lifting, we cannot even
+define the polymorphic identity function.
+\begin{alignat*}{10}
+  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])\hspace{3em} && \Lift\beta && : \un\,(\mk\,a) = a \\
+  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma])              && \Lift\eta && : \mk\,(\un\,a) = a \\
+  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma])
 \end{alignat*}
@@
 \begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
-This concludes the first-order single substitution semantics of the
-SOGAT (\ref{eq:tt}). For reference, we list the same rules again in
-Appendix \ref{app:tt}.
+This concludes the definition of our basic type theory. For
+reference, we list the same rules again in Appendix \ref{app:tt}.
 \end{definition}
@@
 equipped with a substitution rule. There is no need to add more
-structural (substitution calculus) rules.
-\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
+structural (substitution calculus) rules.\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
 these equations are enough to describe the first-order theory of any SOGAT.}
@@
   \begin{alignat*}{10}
-    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top \\
-    & \tt && : \Tm\,\Gamma\,\top && \tt[] && : \tt[\gamma] = \tt \\
-    & \top\eta && : t = \tt \\
-    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
-    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
-    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong \\
-    & && \rlap{$\hphantom{{}:{}} (a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : (\blank,\blank)$} \\
-    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
+    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top\hspace{6em} && \top\eta && : t = \tt \\
+    & \tt && : \Tm\,\Gamma\,\top \hspace{6em} && \tt[] && : \tt[\gamma] = \tt \\
+    & \Sigma && : \rlap{$(A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \hspace{4.4em} \Sigma[] : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+])$} \\
+    & \fst,\snd && : \rlap{$\Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \hspace{1.6em} {,}[] : (a,b)[\gamma] = (a[\gamma],b[\gamma])$}
   \end{alignat*}
   To save space, we compressed the introduction, elimination, $\beta$
-  and $\eta$ rules of $\Sigma$ into one isomorphism, and did not list
-  substitution laws for $\fst$ and $\snd$ as they are derivable (see
-  the examples below).
+  and $\eta$ rules of $\Sigma$ into an isomorphism, and did not list
+  substitution law for $\fst$, $\snd$ as they are derivable (see
+  below).
 \end{definition}
@@
 Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
-\begin{alignat*}{10}
-  & \lam\,(\lam\,\q)\cdot\cd\,A\cdot a && {=}(\Pi\beta) \\
-  & (\lam\,\q[\langle\cd\,A\rangle])\cdot a && {=}(\lam[]) \\
-  & (\lam\,(\q[\langle\cd\,A\rangle^+]))\cdot a\, && {=}([\q][^+]) \\
-  & (\lam\,\q)\cdot a && {=}(\Pi\beta) \\
-  & \q[\langle a\rangle] && {=}(\q[\langle\rangle]) \\
-  & a
-\end{alignat*}
+\[
+\lam\,(\lam\,\q)\cdot\cd\,A\cdot a \overset{\Pi\beta}{=}
+\lam\,\q[\langle\cd\,A\rangle]\cdot a \overset{\lam[]}{=}
+\lam\,(\q[\langle\cd\,A\rangle^+])\cdot a\, \overset{[\q][^+]}{=}
+\lam\,\q\cdot a \overset{\Pi\beta}{=}
+\q[\langle a\rangle] \overset{\q[\langle\rangle]}{=}
+a
+\]
 If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
@@
 accept them:
-\begin{alignat*}{10}
-  & \lam\,(\lam\,\q)[\p] : \Tm\,&& (\diamond\ext C)\,\\
-  & && \Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big)
-\end{alignat*}
+\[
+\lam\,(\lam\,\q)[\p] : \Tm\,(\diamond\ext C)\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big).
+\]
 But we have
-\[
+$
 \lam\,(\lam\,\q)[\p] \overset{\lam[]\,2\text{x}}{=} \lam\,(\lam\,(\q[\p^{++}])) \overset{\q[^+]}{=} \lam\,(\lam\,\q)
-\]
+$
 (with different implicit arguments for the $\lam$s on the left and right hand side), so we are able to apply $\cd\,A'$
@@
 
-The operations for $\U$ and $\Lift$ can be written more consicely as
+The operations for $\U$ and $\Lift$ can be written more concisely as
 the following isomorphisms:
@@
 (\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
-direction (${,}[]$), because naturalities in the other direction are
+direction (${,}[]$), naturalities in the other direction are
 derivable:
@@
   & (\fst\,w)[\gamma] && {=}(\Sigma\beta_1)                             && (\snd\,w)[\gamma] && {=}(\Sigma\beta_2)                             \\
-  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{4em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
+  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{6em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
   & \fst\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        && \snd\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        \\
@@
 \end{alignat*}
-For $\Pi$ types, we construct a similar isomorphism:
-\[
-(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam
-\]
-The roundtrips are proven as follows:
-\[
-\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t
-\]
-The last step is the admissible equation (\ref{eq:lifted4}) proven in
+For $\Pi$ types, we construct a similar isomorphism
+$
+(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
+$
+The first roundtrip equality is proven as
+$\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
+The last step is the admissible equation (\ref{eq:lifted4}) from
 Section \ref{sec:admissible}. As above with weakening into an
@@
 %and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
-The $\eta$ rountrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
-
-We show another isomorphism between terms depending on a variable and
+The otherroundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
+
+We construct another isomorphism between terms depending on a variable and
 terms depending on the lifted version of the same variable:
@@
 \end{alignat*}
-We hope that the above examples demonstrated the for equations
+We hope that the above examples demonstrated the need for equations
 (\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
@@
 
-As a sanity check for our notion of SSC-model, we defined a variant of
-the standard (metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model
-of the SSC-based calculus in Agda. For convenience, we used an
+As a sanity check for our notion of SSC-model, we defined the standard
+(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
+SSC-based calculus in Agda. For convenience, we used an
 inductive-recursive universe to define the infinite hierarchy of types
@@
 In this section we show that our SSC-based syntax is isomorphic to the
-CwF-based syntax with the same type formers.
-\footnote{Using the two translations from SOGATs to GATs
+CwF-based syntax with the same type formers.\footnote{Using the two translations from SOGATs to GATs
 \cite{DBLP:conf/fscd/KaposiX24}, we can rephrase the contents of this
@@
 \begin{alignat}{10}
-  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \label{eq:lifted1}\\
-  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2}\\
-  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3}\\
+  B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \label{eq:lifted1} \\
+  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2} \\
+  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3} \\
   B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B\label{eq:lifted4}
@@
 
-To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we define telescopes as an
-inductive type together with a recursive operation to append telescopes to
-contexts.
-\begin{alignat*}{10}
-  & \Tel && : \Con\ra\Set \\
-  & \blank+\blank && : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con \\
-  & \diamond && : \Tel\,\Gamma \\
-  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma
-\end{alignat*}
-The appending operation is defined recursively on telescopes.
-\begin{alignat*}{10}
-  & \Gamma+\diamond && := \Gamma \\
-  & \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
+To formally state equations (\ref{eq:lifted1})--(\ref{eq:lifted4}), we
+define telescopes as an inductive type together with a recursive
+operation to append telescopes to contexts.
+\begin{alignat*}{10}
+  & \Tel && : \Con\ra\Set   && \rlap{$\blank+\blank : (\Gamma:\Con)\ra\Tel\,\Gamma\ra\Con$} \\
+  & \diamond && : \Tel\,\Gamma && \Gamma+\diamond && := \Gamma \\
+  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma \hspace{3em} && \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
 \end{alignat*}
@@
 instantiation operation on telescopes by recursion on telescopes.
-\begin{alignat*}{10}
-  & \blank[\blank] && : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \\
-  & \blank^{+^\Omega} && : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
-\end{alignat*}
+\[
+\blank[\blank] : \Tel\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Tel\,\Delta \hspace{3em}
+\blank^{+^\Omega} : (\gamma:\Sub\,\Delta\,\Gamma)\ra\Sub\,(\Delta+\Omega[\gamma])\,(\Gamma+\Omega)
+\]
 
@@
   \label{thm:lifting}
-  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if
-  \begin{itemize}
-  \item for any $A:\Ty\,\Gamma$, $A[\gamma_0] = A[\gamma_1]$, and
-  \item for any variable $x:\Tm\,\Gamma\,A$, $x[\gamma_0] = x[\gamma_1]$,
-  \end{itemize}
+  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if for any $A:\Ty\,\Gamma$, $A[\gamma_0] = A[\gamma_1]$, and
+  for any variable $x:\Tm\,\Gamma\,A$, $x[\gamma_0] = x[\gamma_1]$,
   then:
@@
 \begin{alignat*}{10}
-  & \Tms && : \Con\ra\Con\ra\Con \\
-  & \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
-  & \epsilon && : \Tms\,\Gamma\,\diamond \\
-  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A) \\
-  & \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
-  & \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
+  & \Tms && : \Con\ra\Con\ra\Con && \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
+  & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
+  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{2em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
 \end{alignat*}
 The computation of instantiating with $\Tms$ is illustrated below.
-\begin{gather*}
-  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = \\
+\[
+  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = 
   B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
-\end{gather*}
+\]
 All CwF operations and equations can be defined with $\Tms$ by induction, using
@@
 code in the universe, so for a $B : \Ty\,\Gamma\,i$,
-\begin{alignat*}{10}
-  & B[\p][\gamma^+] && {=}(\U\beta) \\
-  & \El\,(\cd\,B)[\p][\gamma^+] && {=}(\El[]) \\
-  & \El\,((\cd\,B)[\p][\gamma^+])\,\, && {=}([\p][^+]'\,\U[]) \\
-  & \El\,((\cd\,B)[\gamma][\p]) && {=}(\El[]) \\
-  & \El\,(\cd\,B)[\gamma][\p] && {=}(\U\beta) \\
-  & B[\gamma][\p].
-\end{alignat*}
+\[
+B[\p][\gamma^+] \overset{\U\beta}{=}
+\El\,(\cd\,B)[\p][\gamma^+] \overset{\El[]}{=}
+\El\,((\cd\,B)[\p][\gamma^+]) \overset{[\p][^+]'\,\U[]}{=}
+\El\,((\cd\,B)[\gamma][\p]) \overset{\El[]}{=}
+\El\,(\cd\,B)[\gamma][\p] \overset{\U\beta}{=}
+B[\gamma][\p].
+\]
 Equation $[\q][^+] : \q[\gamma^+] = \q$ also only makes sense if we
@@
 $\cdot[]$ looks as follows.
-\begin{alignat*}{10}
-  & \cdot[]' :{} && (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra \\
-  & && (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma])
-\end{alignat*}
+\[
+\cdot[]' : (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra(t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma])
+\]
 Again, for $B = \U$ we have the assumption $e$. But then $t$ is in
@@
 
-\bibliography{b}{}
-\bibliographystyle{plain}
-
-\pagebreak
-
-\phantom{A}
-
-\pagebreak
+\bibliography{b}
 

### 2025-01-30 cpp/p.tex
stuff

@@
 \begin{definition}[Minimised single substitution calculus with $\Pi$ and $\U$]\label{def:min}
-  This section defined the minimised first-order version of the SOGAT
-  (\ref{eq:tt}), relying essentially on the presence of universes and
+  This section defined the minimised version of Definition
+  \ref{def:tt}, relying essentially on the presence of universes and
   $\Pi$ types. For reference, we list all the rules of the minimised
@@
   components refer to the old SSC-model):
-  \begin{alignat*}{10}
-    & \Con && := (i:\N)\times\Ty\,\diamond\,i \\
-    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Lift^{\Gamma-\Delta}\,\Delta\Ra\Lift^{\Delta-\Gamma}\,\Gamma)
-  \end{alignat*}
+  \[
+    \Con := (i:\N)\times\Ty\,\diamond\,i \hspace{6em}
+    \Sub\,\Delta\,\Gamma := \Tm\,\diamond\,(\Lift^{\Gamma-\Delta}\,\Delta\Ra\Lift^{\Delta-\Gamma}\,\Gamma)
+  \]
   In the definition of $\Sub$, we did not write the projections for
@@
   \begin{alignat*}{10}
-    & \gamma\circ\delta && := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\, \\[-1.5em]
-    & && \hspace{10em}(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
-    & \rlap{$\id := \lam\,\q \hspace{4.35em} \diamond := (0,\top) \hspace{4.35em} \epsilon := \lam\,(\mk^\Gamma\,\tt)$}
+    & \gamma\circ\delta := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\,(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
+    & \id := \lam\,\q \hspace{10em} \diamond := (0,\top) \hspace{10em} \epsilon := \lam\,(\mk^\Gamma\,\tt)
   \end{alignat*}
@@
     & \Ty\,\Gamma\,i && := \Tm\,\diamond\,\big(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)}\,(\U\,i)\big) \\
-    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\, \\[-1em]
-    & && \hspace{4em} \big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
+    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\,\big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
   \end{alignat*}
@@
   \begin{alignat*}{10}
-    & \Con && := \Ty\,\diamond \\
-    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma) \\
-    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \\
-    & \id && := \lam\,\q \\
-    & \diamond && := \top \\
-    & \epsilon && := \lam\,\tt \\
-    & \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U) \\
-    & A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) \\
-    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \\
-    & a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\
-    & \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\
-    & (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q) \\
-    & \p && := \lam\,(\fst\,\q) \\
-    & \q && := \lam\,(\snd\,\q)
+    & \Con && := \Ty\,\diamond                                           && \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U) && \p && := \lam\,(\fst\,\q) \\                     
+    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma)        && A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) && \q && := \lam\,(\snd\,\q) \\
+    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \hspace{1.5em} && \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q)))\hspace{1.5em} \\   
+    & \id && := \lam\,\q                                                 && a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\                      
+    & \diamond && := \top                                                && \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\                   
+    & \epsilon && := \lam\,\tt                                           && (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q)
   \end{alignat*}
@@
   by pairing/projections of $\Sigma$. All the CwF equations hold, for
-  example we derive the functor law for type substitution as follows.
+  example we derive the functor law for type substitution:
   \begin{equation}
@@
   by $\Sigma$ types, for example,
-  \[
+  $
   \Pi\,A\,B := \lam\,(\cd\,(\Pi\,(\El\,(A[\p]\cdot\q))\,(\El\,(B[\p][\p]\cdot(\q[\p],\q)))))
-  \]
+  $
   and $\lam\,b := \lam\,(\lam\,(b[\p][\p]\cdot(\q[\p],\q)))$. All
@@
   \begin{alignat*}{10}
-    & F : \Con_{M'}\ra\Con_M \\
-    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma) \\
-    & F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
-    & F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
-  \end{alignat*}
-  We define them in the same order as follows, omitting the $M$ subscripts.
-  \begin{alignat*}{10}
-    & F : \Ty\,\diamond\ra\Con \\
-    & F\,\Gamma := \diamond\ext\Gamma \\
-    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) \\
-    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \\
-    & F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
-    & F\,A := \El\,(A[\p]\cdot\q) \\
-    & F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
-    & F\,a := a[\p]\cdot\q
+    & F : \Con_{M'}\ra\Con_M && F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
+    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma)\hspace{3em} && F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
+  \end{alignat*}
+  We define them in the same order, omitting the $M$ subscripts:
+  \begin{alignat*}{10}
+%    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
+    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
+%    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
+    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
   \end{alignat*}

### 2025-02-04 cpp/p.tex
stuff

@@
 \keywords{type theory, category with families, substitution calculus}
+
+%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+%\supplementdetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
+Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newtheorem{problem}[theorem]{Problem}
+\newtheorem{construction}[theorem]{Construction}
 
@@
 \maketitle
-
-\newtheorem{problem}[theorem]{Problem}
-\theoremstyle{remark}
-\newtheorem{construction}[theorem]{Construction}
 
@@
 proved separately for weakenings and $\alpha$-normal single substitutions for
-the induction to be well-founded. We define predicates for weakening and
+the induction to be structural. We define predicates for weakening and
 $\alpha$-normal single substitutions as follows.
@@
   \end{itemize}
-  Note that the later equations depend the the earlier ones.
+  Note that the later equations depend on the the earlier ones.
 \end{lemma}
@@
 substitutions, however it does not satisfy the equations of parallel
-substitutions. Thus we define parallel substitutions as lists of terms with a
-map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to convert it
-into a sequence of instantiations of single substitutions.
+substitutions. Thus we define parallel substitutions as lists of terms. We
+cannot define the instatiation operation for $\Tms$ directly, so we also define
+a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
+convert it into a sequence of instantiations of single substitutions.
 \begin{alignat*}{10}
@@
 \end{alignat*}
-The computation of instantiating with $\Tms$ is illustrated below.
+The computation of instantiating with $\Tms$ is illustrated below. On the
+right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
+then replace the variables of $B$ one by one by $a_i$'s, which are well-typed
+because of the previous weakening.
 \[
@@
 
-\section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
+\section{\texorpdfstring{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}{CwF from SSC with Σ, Π and U}}
 \label{sec:cwf}

### 2025-02-06 cpp/p.tex
stuff

@@
 \begin{itemize}
-\item  In type theory, typing depends on conversion:
-  in an extrinsic presentation, this is expressed by a rule such as
-  the following.
-  \[
-  \infer{\Gamma\vdash t : B}{\Gamma\vdash A \sim B && \Gamma\vdash t : A}
-  \]
-  When conversion $\sim$ is equality, this rule holds by the general
+\item In type theory, typing depends on conversion: in an extrinsic
+  presentation, this is expressed by the rule deriving $\Gamma\vdash t
+  : B$ from $\Gamma\vdash A \sim B$ and $\Gamma\vdash t : A$.  When
+  conversion $\sim$ is equality, this rule holds by the general
   properties of equality; and we also avoid lots of boilerplate
@@
 $\alpha$-normalisation: every term admits an $\alpha$-normal form.
-Thus, when one wants to prove something about all terms, it is enough
-to prove it for $\alpha$-normal forms. Just as $\beta$-normal forms
+Thus, proving something for $\alpha$-normal forms implies it for all terms.
+Just as $\beta$-normal forms
 don't distinguish between $\beta$-convertible terms, $\alpha$-normal
-forms don't include explicit substitutions. $\alpha$-normal forms are
-not like ordinary $\beta/\eta$-normal forms because they are still
+forms don't include explicit instantiations. $\alpha$-normal forms are
+not like ordinary $\beta/\eta$-normal forms because they are 
 quotiented by the computation/uniqueness rules like $\beta$/$\eta$ for
@@
 
-In summary, our contributions are the following:
-\begin{itemize}
-\item A new generalised algebraic presentation of type theory in the
+In summary, our contributions are (in the same order as the structure of the paper):
+\begin{description}
+\item[Section \ref{sec:tt}] A new generalised algebraic presentation of type theory in the
   form of a minimalistic single substitution calculus. Our calculus
@@
   easy-to-understand way where all operations are well-motivated.
-\item The $\alpha$-normalisation technique which shows that the syntax
+\item[Section \ref{sec:admissible}] The $\alpha$-normalisation technique which shows that the syntax
   of our single substitution calculus is isomorphic to the syntax of
   the CwF-based theory.
-\item For type theory with $\Pi$, $\U$, a minimised presentation of
+\item[Section \ref{sec:minimisation}] For type theory with $\Pi$, $\U$, a minimised presentation of
   the equations which results in an isomorphic theory.
-\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
+\item[Section \ref{sec:cwf}] For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
   single and parallel models are equivalent in a weak sense.
-\end{itemize}
-
-\subsection{Structure of the paper}
-
-After discussing related work and our metatheory, in Section
-\ref{sec:tt}, we introduce our single substitution calculus (SSC) only
-relying on a working knowledge of an implementation of type theory
-such as Agda, no prior experience with the metatheory of type theory
-is required. Section \ref{sec:admissible} shows that in the syntax of
-SSC, several new equations are admissible, and that the syntax of SSC
-is actually isomorphic to the CwF-based syntax. In Section
-\ref{sec:minimisation}, we minimise our SSC obtaining a theory with
-fewer equations, relying on the presence of $\Pi$ and $\U$ in our
-theory. In Section \ref{sec:cwf}, we show that in the presence of
-$\Pi$, $\U$, $\Sigma$, SSC-models are actually equivalent to
-CwF-models in a weak sense. We conclude in Section
-\ref{sec:conclusion}.
+\end{description}
+
+% \subsection{Structure of the paper}
+% 
+% After discussing related work and our metatheory, in Section
+% \ref{sec:tt}, we introduce our single substitution calculus (SSC) only
+% relying on a working knowledge of an implementation of type theory
+% such as Agda, no prior experience with the metatheory of type theory
+% is required. Section \ref{sec:admissible} shows that in the syntax of
+% SSC, several new equations are admissible, and that the syntax of SSC
+% is actually isomorphic to the CwF-based syntax. In Section
+% \ref{sec:minimisation}, we minimise our SSC obtaining a theory with
+% fewer equations, relying on the presence of $\Pi$ and $\U$ in our
+% theory. In Section \ref{sec:cwf}, we show that in the presence of
+% $\Pi$, $\U$, $\Sigma$, SSC-models are actually equivalent to
+% CwF-models in a weak sense. We conclude in Section
+% \ref{sec:conclusion}.
 
@@
 
-\subparagraph*{B-systems.} B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
-an algebraic description of type theory close to extrinsic definitions.
-B-systems are an intrinsic, essentially
+\subparagraph*{B-systems.} B-systems introduced by Voevodsky \cite{voevodsky2014bsystems}
+are an intrinsic, essentially
 algebraic presentation of type theory using single
@@
 theory, all the rules of B-systems are admissible. We describe the
-relationship in more detail. B-systems are B-frames together with
-substitution, weakening and generic element operations. A B-frame is
-given by sets $B_i$, $\tilde{B}_i$ and functions between them as shown
-below \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. \\
-\begin{tikzpicture}
-  \node (b0) at (4,0) {$\top$};
-  \node (b1) at (5,0) {$B_1$};
-  \node (b2) at (12,0) {$B_2$};
-  \node (b3) at (15,0) {$\dots$};
-  \node (c1) at (7,1) {$\tilde{B}_1$};
-  \node (c2) at (14,1) {$\tilde{B}_2$};
-  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
-  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
-  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
-  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
-  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
-\end{tikzpicture} \\
-Using our notation, this is the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
-\begin{tikzpicture}
-  \node (b0) at (4,0) {$\top$};
-  \node (b1) at (5,0) {$\Ty\,\diamond$};
-  \node (b2) at (12,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
-  \node (b3) at (15,0) {$\dots$};
-  \node (c1) at (7,1)  {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
-  \node (c2) at (14,1) {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
-  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
-  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
-  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
-  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
-  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
-\end{tikzpicture} \\
-The substitution operation $\mathbb{S}$ for an $x:\tilde{B}_{n+1}$ is
-a homomorphism of the slice B-frames $\mathbb{B}/\partial(x) \ra
-\mathbb{B}/\mathsf{ft}(\partial(x))$. In our notation, $x = (\Gamma,
-A, a)$ where $a : \Tm\,\Gamma\,A$, and $\mathbb{S}$ corresponds to the
-following functions for $\Ty$ (a map between the bottom rows of the
-diagrams):
-\begin{alignat*}{10}
-  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
-  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
-  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
-  & \dots,
-\end{alignat*}
-and similarly $\mathbb{S}$ also includes all the (lifted) substitution
-operations for terms (top rows of the
-diagrams). Analogously, the weakening operation corresponds to
-$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
-our notation. The six groups of equations correspond to our
-$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
-$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
-order. However we don't include the lifted versions of these equations,
-equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
-stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
-for terms. The missing equations are admissible, see Section
-\ref{sec:admissible}. In the presence of $\U$ and $\Pi$, we
-reduce the needed equations even more, see Section
-\ref{sec:minimisation}.
+relationship in more detail in Appendix \ref{app:bsys}.
 
@@
 of their calculus is equivalent to the category of contextual simply
-typed CwFs with function space. Their equivalence result holds even
-without the presence of function types. In the presence of function
-space, the same equivalence holds for the simply typed version of our
-substitution calculus, see the formalisation accompanying this paper.
+typed CwFs with function space.
+%Their equivalence result holds even without the presence of function types.
+%In the presence of function space,
+The same equivalence holds for the simply typed version of our
+substitution calculus (TODO link).
 
 \subparagraph*{Ehrhard's calculus.}
-The first generalised algebraic presentation of type theory was Thomas
+The first generalised algebraic presentation of type theory was
 Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
@@
   \begin{alignat*}{10}
-    & \top && : \Ty\,\Gamma\,0 && \top[] && : \top[\gamma] = \top\hspace{6em} && \top\eta && : t = \tt \\
-    & \tt && : \Tm\,\Gamma\,\top \hspace{6em} && \tt[] && : \tt[\gamma] = \tt \\
-    & \Sigma && : \rlap{$(A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \hspace{4.4em} \Sigma[] : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+])$} \\
-    & \fst,\snd && : \rlap{$\Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \hspace{1.6em} {,}[] : (a,b)[\gamma] = (a[\gamma],b[\gamma])$}
+    & \rlap{$
+  \top : \Ty\,\Gamma\,0 \hspace{3em}
+  \top[] : \top[\gamma] = \top\hspace{3em}
+  \top\eta : t = \tt \hspace{3em}
+  \tt : \Tm\,\Gamma\,\top \hspace{3em}
+  \tt[] : \tt[\gamma] = \tt
+      $} \\
+    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
+    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
+    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
+    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
   \end{alignat*}
@@
 substitutions, however it does not satisfy the equations of parallel
-substitutions. Thus we define parallel substitutions as lists of terms. We
-cannot define the instatiation operation for $\Tms$ directly, so we also define
+substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
+difficult to define the instatiation operation for $\Tms$ directly, so we also define
 a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
@@
 
-\section{GAT presentation of (\ref{eq:tt})}
+\section{Detailed comparison with B-systems}
+\label{app:bsys}
+
+In this section, we describe the relationship between B-systems
+\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} and our Definition
+\ref{def:tt} in detail.
+
+B-systems are B-frames
+together with substitution, weakening and generic element
+operations. A B-frame is given by sets $B_i$, $\tilde{B}_i$ and
+functions between them as shown below . \\
+\begin{tikzpicture}
+  \node (b0) at (4,0) {$\top$};
+  \node (b1) at (5,0) {$B_1$};
+  \node (b2) at (12,0) {$B_2$};
+  \node (b3) at (15,0) {$\dots$};
+  \node (c1) at (7,1) {$\tilde{B}_1$};
+  \node (c2) at (14,1) {$\tilde{B}_2$};
+  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
+  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
+  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
+  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
+  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
+\end{tikzpicture} \\
+Using our notation, this is the following data ($\diamond$ is the empty context, $\ext$ is context extension). \\
+\begin{tikzpicture}
+  \node (b0) at (4,0) {$\top$};
+  \node (b1) at (5,0) {$\Ty\,\diamond$};
+  \node (b2) at (12,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
+  \node (b3) at (15,0) {$\dots$};
+  \node (c1) at (7,1)  {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
+  \node (c2) at (14,1) {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
+  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
+  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
+  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
+  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
+  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
+\end{tikzpicture} \\
+The substitution operation $\mathbb{S}$ for an $x:\tilde{B}_{n+1}$ is
+a homomorphism of the slice B-frames $\mathbb{B}/\partial(x) \ra
+\mathbb{B}/\mathsf{ft}(\partial(x))$. In our notation, $x = (\Gamma,
+A, a)$ where $a : \Tm\,\Gamma\,A$, and $\mathbb{S}$ corresponds to the
+following functions for $\Ty$ (a map between the bottom rows of the
+diagrams):
+\begin{alignat*}{10}
+  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
+  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
+  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
+  & \dots,
+\end{alignat*}
+and similarly $\mathbb{S}$ also includes all the (lifted) substitution
+operations for terms (top rows of the
+diagrams). Analogously, the weakening operation corresponds to
+$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
+our notation. The six groups of equations correspond to our
+$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
+$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
+order. However we don't include the lifted versions of these equations,
+equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
+stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
+for terms. The missing equations are admissible, see Section
+\ref{sec:admissible}. In the presence of $\U$ and $\Pi$, we
+reduce the needed equations even more, see Section
+\ref{sec:minimisation}.
+
+\section{GAT presentation of Definition \ref{def:tt}}
 \label{app:tt}
 
-Definition \ref{def:tt} in full detail: \\
 The core substitution calculus:
@@
   & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
-  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])
-\end{alignat*}
-\begin{alignat*}{10}
+  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
   & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
@@
 
-\section{Minimised GAT presentation of (\ref{eq:tt})}
+\section{Minimised GAT presentation of Definition \ref{def:min} (omitting $\Lift$)}
 \label{app:minimisation}
 
-Definition \ref{def:min} in full detail (omitting $\Lift$):
 \begin{alignat*}{10}

### 2025-02-20 cpp/p.tex
stugf

@@
 
+\vspace{-1.1em}
 \subparagraph*{Ehrhard's calculus.}
@@
 
+\vspace{-1.1em}
 \subparagraph*{Formalisations of type theory.}
@@
 
+\vspace{-1.1em}
 \subparagraph*{SOGATs}
@@
 \ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
+\begin{multicols}{2}
 \begin{alignat}{10}
@@
   B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2} \\
+\end{alignat}
+\
+\begin{alignat}{10}
   B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3} \\
@@
 \end{alignat}
+\end{multicols}
 The corresponding equations on terms are also needed. Note that we do not have
@@
 explicit). An alternative presentation of the second equation without
-requiring the first one is the following:
-\[
-[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p]
-\]
+requiring the first one is
+$
+[\p][^+]' : (e : B[\p][\gamma^+] = B[\gamma][\p])\ra b[\p][\gamma^+] =_{e} b[\gamma][\p].
+$
 That is, for any type $B$ for which $B[\p][\gamma^+] = B[\gamma][\p]$,
@@
 with the conditional equation
-\[
-[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b,
-\]
-notice that we have the input for $B = \U\,i$, and derive the input
+$
+[\p][\langle\rangle]' : (e:B[\p][\langle a\rangle] = B)\ra b[\p][\langle a\rangle] =_e b.
+$
+Notice that we have the input for $B = \U\,i$, and derive the input
 equation for all types. We make $\q[\langle\rangle]$ conditional as
@@
 Now we turn our attention to the equation
-\[
+$
   [\langle\rangle][] : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle].
-\]
+$
 We were forced to introduce it so that we can state the substitution
 rule for function application $\cdot[]$. A conditional version of
-$\cdot[]$ looks as follows.
-\[
-\cdot[]' : (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra(t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma])
-\]
+$\cdot[]$ is
+$
+\cdot[]' : (e : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle])\ra(t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]).
+$
 Again, for $B = \U$ we have the assumption $e$. But then $t$ is in
@@
 A)$. So we argue as follows for any $B : \Ty\,(\Gamma\ext A)$.
-\begin{alignat*}{10}
-  & B[\langle a\rangle][\gamma] && {=}(\U\beta) \\
-  & \El\,(\cd\,(B[\langle a\rangle][\gamma])) && {=}(\cd[]) \\
-  & \El\,((\cd\,B)[\langle a\rangle][\gamma]) && {=}(\Pi\beta) \\
-  & \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) && {=}(\cdot[]'\,\U[]) \\
-  & \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma]))\,\, && {=}(\lam[]) \\
-  & \El\,(\lam\,(\cd\,B[\gamma^+])\cdot (a[\gamma])) && {=}(\Pi\beta) \\
-  & \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) && {=}(\cd[]) \\
-  & \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] && {=}(\U\beta) \\
-  & B[\gamma^+][\langle a[\gamma]\rangle]
+\begin{alignat*}{20}
+  & B[\langle a\rangle][\gamma] \overset{\U\beta}{=}
+  \El\,(\cd\,(B[\langle a\rangle][\gamma])) \overset{\cd[]}{=}
+  \El\,((\cd\,B)[\langle a\rangle][\gamma]) \overset{\Pi\beta}{=}
+  \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) \overset{\cdot[]'\,\U[])}{=} \\
+  & \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma])) \overset{\lam[]}{=}
+  \El\,(\lam\,(\cd\,B[\gamma^+])\cdot (a[\gamma])) \overset{\Pi\beta}{=}
+  \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) \overset{\cd[]}{=} \\
+  & \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] \overset{\U\beta}{=}
+  B[\gamma^+][\langle a[\gamma]\rangle]
 \end{alignat*}
@@
 
-Finally, we will remove equation $[\p^+][\langle\q\rangle]$ by making
-$\Pi\eta$ conditional. This needs another change: we substitute
-$\Pi\beta$ for a more general variant which is also conditional on the
-same equation.
+Finally, we remove equation $[\p^+][\langle\q\rangle]$ by making
+$\Pi\eta$ conditional. This needs another change: replacing $\Pi\beta$
+with a more general variant conditional on the same equation.
 \begin{alignat*}{10}
@@
 \begin{alignat*}{10}
-  & \Pi\beta^\U :{} &&  \lam\,\hat{B}\cdot a && {=}(\q[\langle\rangle]) \\
-  & && \lam\,\hat{B}\cdot(\q[\langle a\rangle])\,\, && {=}([\p][\langle\rangle]) \\
-  & && ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle])\,\, && {=}(\cdot[]'\,\U[]) \\
-  & && ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] && {=}(\Pi\beta'\,\U[]) \\
-  & && \hat{B}[\langle a\rangle].
+  & \Pi\beta^\U :{} \lam\,\hat{B}\cdot a \overset{\q[\langle\rangle])}{=}
+  \lam\,\hat{B}\cdot(\q[\langle a\rangle]) \overset{[\p][\langle\rangle]}=
+  ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle]) \overset{\cdot[]'\,\U[]}= \\
+  & ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] \overset{\Pi\beta'\,\U[]}=
+  \hat{B}[\langle a\rangle].
 \end{alignat*}
@@
 \begin{alignat*}{10}
-  & B[\p^+][\langle\q\rangle] && {=}(\U\beta) \\
-  & \El\,(\cd\,B)[\p^+][\langle\q\rangle] && {=}(\El[]) \\
-  & \El\,(\cd\,B[\p^+][\langle\q\rangle]) && {=}(\Pi\beta^\U) \\
-  & \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) && {=}(\lam[]) \\
-  & \El\,(\lam\,(\cd\,B)[\p]\cdot\q) && {=}(\Pi\beta'\,\U[]) \\
-  & \El\,(\cd\,B) && {=}(\U\beta) \\
-  & B
+  & B[\p^+][\langle\q\rangle] \overset{\U\beta}=
+  \El\,(\cd\,B)[\p^+][\langle\q\rangle] \overset{\El[]}=
+  \El\,(\cd\,B[\p^+][\langle\q\rangle]) \overset{\Pi\beta^\U}=
+  \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) \overset{\lam[]}= \\
+  & \El\,(\lam\,(\cd\,B)[\p]\cdot\q) \overset{\Pi\beta'\,\U[]}=
+  \El\,(\cd\,B) \overset{\U\beta}=
+  B
 \end{alignat*}
@@
   \begin{alignat*}{10}
-    & \gamma\circ\delta := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\,(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
+%    & \gamma\circ\delta := \lam\Bigg(\mk^{\Theta-\Gamma}\,\bigg(\un^{\Delta-\Gamma}\,\Big(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,\big(\un^{\Theta-\Delta}\,(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q))\big)\Big)\bigg)\Bigg) \\
+    & \gamma\circ\delta := \lam\,(\mk^{\Theta-\Gamma}\,(\un^{\Delta-\Gamma}\,(\gamma[\p]\cdot\mk^{\Gamma-\Delta}\,(\un^{\Theta-\Delta}\,(\delta[\p]\cdot\mk^{\Delta-\Theta}\,(\un^{\Gamma-\Theta}\,\q)))))) \\
     & \id := \lam\,\q \hspace{10em} \diamond := (0,\top) \hspace{10em} \epsilon := \lam\,(\mk^\Gamma\,\tt)
@@
     & \Ty\,\Gamma\,i && := \Tm\,\diamond\,\big(\Lift^{1+i-\Gamma}\,\Gamma\Ra\Lift^{\Gamma-(1+i)}\,(\U\,i)\big) \\
-    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\,\big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
+%    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,\Bigg(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,\bigg(\Lift^{\Gamma-i}\,\Big(\El\,\big(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))\big)\Big)\bigg)\Bigg)
+    & \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,(\Lift^{i-\Gamma}\,\Gamma)\,(\Lift^{\Gamma-i}\,(\El\,(\un^{\Gamma-(1+i)}\,(A[\p]\cdot\mk^{1+i-\Gamma}\,(\un^{i-\Gamma}\,\q))))))
   \end{alignat*}
@@
   We define them in the same order, omitting the $M$ subscripts:
-  \begin{alignat*}{10}
+  \[
+  F\,\Gamma := \diamond\ext\Gamma  \hspace{3em}
+  F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em}
+  F\,A := \El\,(A[\p]\cdot\q)  \hspace{3em}
+  F\,a := a[\p]\cdot\q
+  \]
+%  \begin{alignat*}{10}
 %    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
-    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
+%    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
 %    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
-    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
-  \end{alignat*}
+%    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
+%  \end{alignat*}
   It is easy to check that $F$ preserves the CwF structure, i.e.\ it
@@
 
-\section{GAT presentation of Definition \ref{def:tt}}
+\section{Listing of Definition \ref{def:tt}}
 \label{app:tt}
+
+In Section \ref{sec:tt}, we presented the single substitution syntax
+interleaved with explanations, here we list all the rules in one
+place, for reference.
 
@@
 
-\section{Minimised GAT presentation of Definition \ref{def:min} (omitting $\Lift$)}
+\section{Listing of Definition \ref{def:min}}
 \label{app:minimisation}
 
+For reference, we list the rules for the minimised single substitution
+syntax defined in Section \ref{sec:minimisation}.
 \begin{alignat*}{10}
@@
   & \U\beta && : \El\,(\cd\,A) = A \\
-  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
+  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
+  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \\
+  & \Lift[] && : (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
+  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \\
+  & \mk[] && : (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
+  & \un[] && : (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
+  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \\
+  & \Lift\beta && : \un\,(\mk\,a) = a \\
+  & \Lift\eta && : \mk\,(\un\,a) = a
 \end{alignat*}

### 2025-02-21 cpp/p.tex
stuff

@@
 
-\newtheorem{problem}[theorem]{Problem}
-\newtheorem{construction}[theorem]{Construction}
-
 \usepackage{tikz}
@@
 \usepackage{proof}
+\usepackage{multicol}
+
+\newtheorem{problem}[theorem]{Problem}
+\newtheorem{construction}[theorem]{Construction}
+
+\newcommand{\Agda}{%
+  \raisebox{-.15em}{\includegraphics[height=.9em]{agda-logo.pdf}}%
+}
 
@@
 \ref{sec:admissible} of this paper are formalised in the proof
-assistant Agda. The formalisation is availble at TODO. We use a strict
+assistant Agda. The formalisation is availble at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
 Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
@@
 \ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
-\begin{multicols}{2}
+\begin{multicols}{2}\noindent
 \begin{alignat}{10}
   B[\p^{+^n}][(\gamma^+)^{+^n}] & = B[\gamma^{+^n}][\p^{+^n}] \label{eq:lifted1} \\
-  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2} \\
+  B[\p^{+^n}][\langle a \rangle^{+^n}] & = B \label{eq:lifted2}
 \end{alignat}
-\
+\columnbreak
 \begin{alignat}{10}
@@
 right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
-then replace the variables of $B$ one by one by $a_i$'s, which are well-typed
-because of the previous weakening.
+then replace the variables of $B$ one by one using the $a_i$'s, which are
+well-typed because of the previous weakening.
 \[

### 2025-02-21 cpp/p.tex
minimisation

@@
 
-\title{Type theory in type theory with single substitutions}
+\title{Type theory with single substitutions}
 
@@
 we show that SSC models are actually equivalent to CwF-based models in
-a weak sense. Sections 2 and 3 of this paper were formalised in Agda.
+a weak sense.
 \end{abstract}
@@
 over objects.
-
+%
 The GAT presentation of type theory is intrinsic
@@
 arbitrary model, but they hold in the syntax. In this paper, we show
-that the \emph{syntaxes} of our SSC and the corresponding parallel
+that the \emph{syntaxes} of SSC and the corresponding parallel
 substitution calculus (CwF-based calculus) are isomorphic. All
-CwF-based models are models of our SSC, but not
+CwF-based models are models of SSC, but not
 necessarily the other way. While B-systems are equivalent to CwFs (and
 C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
-SSC models than CwF-based models. This relationship is similar to that
+SSC models than CwF-based models. This relationship is like that
 of extensional combinatory calculus and lambda calculus, where the
 syntaxes are equivalent, and every lambda model is a combinatory
-model, but not the other way round
+model, but not the other way
 \cite{DBLP:conf/fscd/AltenkirchKSV23}. Another example is the
 relationship between monoids over a set $X$ and nil/cons algebras over
-$X$, where every monoid is a nil/cons-algebra, but not the other way;
-but it is well-known that free monoids over $X$ (syntax of monoid over
-$X$) are the same as lists of $X$ (the syntax of nil/cons algebras
+$X$: every monoid is a nil/cons-algebra, but not the other way;
+it is well-known that free monoids over $X$ (syntax of monoid over
+$X$) are the same as $X$-lists (the syntax of nil/cons algebras
 over $X$).
@@
 dependent types. Then the SSC is just there for bootstrapping
-purposes: we use the SSC to formulate $\Sigma$, $\Pi$, $\U$, and then
+purposes: SSC formulates $\Sigma$, $\Pi$, $\U$, and then
 we use these to define a parallel substitution calculus. In summary,
@@
 internal to extensional type theory, and most results were formalised
-in Agda. Some results in this paper were presented at the TYPES 2024
-conference \cite{singleTypes}.
+in Agda. %Some results in this paper were presented at the TYPES 2024
+%conference \cite{singleTypes}.
 
@@
 relationship in more detail in Appendix \ref{app:bsys}.
-
+%
 Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
@@
 
+\vspace{-0.7em}
 \subsection{Metatheory and formalisation}\label{sec:metatheory}
@@
 elements of $\Sub$ have the form
-\[
-\p^{+^n} : \Sub\,(\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}])\,(\Gamma\ext B_1\ext B_2\ext\dots\ext B_n).
-\]
-where $+^n$ means the $n$-times iteration of $\blank^+$.
+$
+\p^{+^n} : \Sub\,(\Gamma\ext A\ext B_1[\p]\ext B_2[\p^+]\ext\dots\ext B_n[\p^{+^{n-1}}])\,(\Gamma\ext B_1\ext B_2\ext\dots\ext B_n),
+$
+where $^{+^n}$ means the $n$-times iteration of $\blank^+$.
 
@@
 
-Now that we have $[\p][^+]$ for types, we can derive the instantiation
-rule for nondependent function space as follows.
-\[
+Now that we have $[\p][^+]$ for types, we can derive its instantiation rule by
+$
 (A\Ra B)[\gamma] =
@@
 \Pi\,(A[\gamma])\,(B[\gamma][\p]) =
-A[\gamma]\Ra(B[\gamma])
-\]
+(A[\gamma])\Ra(B[\gamma]).
+$
 
@@
 equipped with a substitution rule. There is no need to add more
-structural (substitution calculus) rules.\footnote{This is made formal in \cite{DBLP:conf/fscd/KaposiX24} where
-these equations are enough to describe the first-order theory of any SOGAT.}
+structural (substitution calculus) rules.\footnote{In \cite{DBLP:conf/fscd/KaposiX24},
+the same structural rules suffice to describe the first-order theory for any SOGAT.}
 For example, we show how to
@@
 \[
-\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
+\lam\,(\lam\,\q) : \Tm\,\diamond\,(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))).
 \]
@@
 argument, and these arguments themselves are terms written using
-implicit arguments. Inrinsically typed terms are the same as
+implicit arguments. Intrinsically typed terms are the same as
 derivation trees, we illustrate this by deriving this term as follows.
@@
 When working informally (or in extensional type theory), we don't
-write the coercion, but Agda requires it.
+write coercions.%, but Agda requires it.
 
 Given a type $A:\Ty\,\diamond\,0$, and $a : \Tm\,\diamond\,A$, we have
-\[
+$
 \lam\,(\lam\,\q)\cdot\cd\,A\cdot a \overset{\Pi\beta}{=}
@@
 \q[\langle a\rangle] \overset{\q[\langle\rangle]}{=}
-a
-\]
+a.
+$
 If we want to specify open inputs $A':\Ty\,(\diamond\ext C)\,0$, and
@@
 accept them:
-\[
-\lam\,(\lam\,\q)[\p] : \Tm\,(\diamond\ext C)\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)[\p]\Big).
-\]
+$
+\lam\,(\lam\,\q)[\p] : \Tm\,(\diamond\ext C)\,(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))[\p]).
+$
 But we have
@@
 term in context $\diamond$ to an arbitrary context $\Gamma$, where
-$\Gamma$ is a metavariable. If we know that $\Gamma$ has lengh $n$,
-then we can do the weakening by
-$\blank\underbrace{[\p][\p]\dots[\p]}_{\text{$n$ times}}$. If we work
+$\Gamma$ is a metavariable. If we know that $\Gamma$ has length $n$,
+then we can do the weakening by applying $\blank[\p]$ to it $n$ times.
+If we work
 in the syntax, then we can derive such weakenings (and more) using the
@@
 $\Sigma$ types were introduced by a similar isomorphism, and we give
-names to the roundtrips eqautions according to our convention introduced in
+names to the roundtrips equations according to our convention introduced in
 Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
@@
 direction (${,}[]$), naturalities in the other direction are
-derivable:
-\begin{alignat*}{10}
-  & (\fst\,w)[\gamma] && {=}(\Sigma\beta_1)                             && (\snd\,w)[\gamma] && {=}(\Sigma\beta_2)                             \\
-  & \fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \hspace{6em} && \snd\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big)\, && {=}({,}[]) \\
-  & \fst\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        && \snd\,\big((\fst\,w,\snd\,w)[\gamma]\big) && {=}(\Sigma\eta)        \\
-  & \fst\,(w[\gamma])                                                && && \snd\,(w[\gamma])                                                   
-\end{alignat*}
+derivable, for $\fst$ this is
+$
+(\fst\,w)[\gamma]  \overset{\Sigma\beta_1}{=}                          
+\fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big) \overset{{,}[]}{=} 
+\fst\,\big((\fst\,w,\snd\,w)[\gamma]\big)  \overset{\Sigma\eta}=        
+\fst\,(w[\gamma]).
+$
 For $\Pi$ types, we construct a similar isomorphism
@@
 %and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
-The otherroundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
+The other roundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
 
@@
 terms depending on the lifted version of the same variable:
-\[
+\begin{equation}\label{eq:liftvar}
 (\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
-\]
-To derive the roundtrips, we again need admissible equations. Here is
-the proof of the first roundtrip equality, the other is analogous:
-\begin{alignat*}{10}
-  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[]) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+]) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[]) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
-  & t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
-  & t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
-  & t
-\end{alignat*}
+\end{equation}
+To derive the roundtrips, we again need admissible equations. We prove
+the first roundtrip equality in Appendix \ref{app:liftvar}, the other
+is analogous.
 Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
 \begin{alignat*}{10}
-  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) && {\cong}\text{($\Lift$ iso)} \\
-  & \Tm\,\Gamma\,(\Pi\,A\,B) && {\cong}\text{($\Pi$ iso)} \\
-  & \Tm\,(\Gamma\ext A)\,B && {\cong}\text{($\Lift$ iso)} \\
-  & \Tm\,(\Gamma\ext A)\,(\Lift\,B) && {\cong}\text{(lifted variable iso)} \\
-  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) && {\cong}\text{($\Pi$ iso)} \\
-  & \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
+  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
+  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
+  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
+  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
+  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
+  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
 \end{alignat*}
@@
 
-\subsection{Standard model}
-\label{sec:standard}
-
-As a sanity check for our notion of SSC-model, we defined the standard
-(metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
-SSC-based calculus in Agda. For convenience, we used an
-inductive-recursive universe to define the infinite hierarchy of types
-\cite{DBLP:journals/jsyml/Dybjer00,DBLP:conf/csl/Kovacs22}. All
-SSC-equations hold by reflexivity in this model. We defined the
-analogous CwF-based standard model.
+% \subsection{Standard model}
+% \label{sec:standard}
+% 
+% As a sanity check for our notion of SSC-model, we defined the standard
+% (metacircular \cite{DBLP:conf/popl/AltenkirchK16}) model of the
+% SSC-based calculus in Agda. For convenience, we used an
+% inductive-recursive universe to define the infinite hierarchy of types
+% \cite{DBLP:journals/jsyml/Dybjer00,DBLP:conf/csl/Kovacs22}. All
+% SSC-equations hold by reflexivity in this model. We defined the
+% analogous CwF-based standard model.
 
@@
 \cite{DBLP:conf/fscd/KaposiX24}, we can rephrase the contents of this
-section as follows: the syntaxes of the single and parallel
-substitution first order versions of the SOGAT \ref{eq:tt} are
+section as follows: the parallel and single substitution syntax of the SOGAT (\ref{eq:tt}) are
 isomorphic.}
@@
 \ref{sec:tt} lifted over arbitrary amount of $\blank^+$s.
+\setlength{\columnsep}{4em}
 \begin{multicols}{2}\noindent
@@
 \begin{alignat}{10}
-  B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3} \\
-  B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B\label{eq:lifted4}
+  \hspace{-4em} B[\langle a \rangle^{+^n}][\gamma^{+^n}] & = B[(\gamma^+)^{+^n}][\langle a[\gamma] \rangle^{+^n}] \label{eq:lifted3} \\
+  \hspace{-4em} B[(\p^+)^{+^n}][\langle\q\rangle^{+^n}] & = B\label{eq:lifted4}
 \end{alignat}
 \end{multicols}
+\vspace{-2em}
 The corresponding equations on terms are also needed. Note that we do not have
@@
   & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
-  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{2em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
+  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{1.9em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
 \end{alignat*}
@@
 code in the universe, so for a $B : \Ty\,\Gamma\,i$,
-\[
+$
 B[\p][\gamma^+] \overset{\U\beta}{=}
@@
 B[\gamma][\p].
-\]
+$
 Equation $[\q][^+] : \q[\gamma^+] = \q$ also only makes sense if we
@@
 $\Tm\,\Gamma\,(\Pi\,A\,\U)$, which is isomorphic to $\Ty\,(\Gamma\ext
-A)$. So we argue as follows for any $B : \Ty\,(\Gamma\ext A)$.
-\begin{alignat*}{20}
-  & B[\langle a\rangle][\gamma] \overset{\U\beta}{=}
+A)$. So we argue as follows for any $B : \Ty\,(\Gamma\ext A)$:
+$
+  B[\langle a\rangle][\gamma] \overset{\U\beta}{=}
   \El\,(\cd\,(B[\langle a\rangle][\gamma])) \overset{\cd[]}{=}
   \El\,((\cd\,B)[\langle a\rangle][\gamma]) \overset{\Pi\beta}{=}
-  \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) \overset{\cdot[]'\,\U[])}{=} \\
-  & \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma])) \overset{\lam[]}{=}
+  \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) \overset{\cdot[]'\,\U[])}{=}
+  \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma])) \overset{\lam[]}{=}
   \El\,(\lam\,(\cd\,B[\gamma^+])\cdot (a[\gamma])) \overset{\Pi\beta}{=}
-  \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) \overset{\cd[]}{=} \\
-  & \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] \overset{\U\beta}{=}
-  B[\gamma^+][\langle a[\gamma]\rangle]
-\end{alignat*}
+  \El\,(\cd\,B[\gamma^+][\langle a[\gamma]\rangle]) \overset{\cd[]}{=}
+  \El\,(\cd\,B)[\gamma^+][\langle a[\gamma]\rangle] \overset{\U\beta}{=}
+  B[\gamma^+][\langle a[\gamma]\rangle].
+$
 Hence, the conditional $\cdot[]'$ equation implies the condition for
@@
   a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
-$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$,
-\begin{alignat*}{10}
-  & \Pi\beta^\U :{} \lam\,\hat{B}\cdot a \overset{\q[\langle\rangle])}{=}
+$\hat{B} : \Tm\,(\Gamma\ext A)\,\U$, we argue
+$
+  \Pi\beta^\U :{} \lam\,\hat{B}\cdot a \overset{\q[\langle\rangle])}{=}
   \lam\,\hat{B}\cdot(\q[\langle a\rangle]) \overset{[\p][\langle\rangle]}=
-  ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle]) \overset{\cdot[]'\,\U[]}= \\
-  & ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] \overset{\Pi\beta'\,\U[]}=
+  ((\lam\,\hat{B})[\p][\langle a\rangle])\cdot(\q[\langle a\rangle]) \overset{\cdot[]'\,\U[]}=
+  ((\lam\,\hat{B})[\p]\cdot\q)[\langle a\rangle] \overset{\Pi\beta'\,\U[]}=
   \hat{B}[\langle a\rangle].
-\end{alignat*}
-Now we derive the assumption of $\Pi\beta'/\Pi\eta'$ for any type $B : \Ty\,(\Gamma\ext A)$.
-\begin{alignat*}{10}
-  & B[\p^+][\langle\q\rangle] \overset{\U\beta}=
+$
+Now, for any type $B : \Ty\,(\Gamma\ext A)$, we derive the assumption of $\Pi\beta'/\Pi\eta'$:
+$
+  B[\p^+][\langle\q\rangle] \overset{\U\beta}=
   \El\,(\cd\,B)[\p^+][\langle\q\rangle] \overset{\El[]}=
   \El\,(\cd\,B[\p^+][\langle\q\rangle]) \overset{\Pi\beta^\U}=
-  \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) \overset{\lam[]}= \\
-  & \El\,(\lam\,(\cd\,B)[\p]\cdot\q) \overset{\Pi\beta'\,\U[]}=
+  \El\,(\lam\,(\cd\,B[\p^+])\cdot\q) \overset{\lam[]}= 
+  \El\,(\lam\,(\cd\,B)[\p]\cdot\q) \overset{\Pi\beta'\,\U[]}=
   \El\,(\cd\,B) \overset{\U\beta}=
-  B
-\end{alignat*}
+  B.
+$
 We summarise this section by formally stating what we defined.
@@
 new model are terms in the old model.
-\begin{problem}[Termification]
+\begin{problem}[Termification]\label{prob:termification}
   From a model of SSC (with $\Pi, \dots, \Sigma$, see Definition
@@
     & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma)        && A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) && \q && := \lam\,(\snd\,\q) \\
-    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \hspace{1.5em} && \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q)))\hspace{1.5em} \\   
+    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \hspace{1.5em} && \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q)))\hspace{1.1em} \\   
     & \id && := \lam\,\q                                                 && a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\                      
@@
   by pairing/projections of $\Sigma$. All the CwF equations hold, for
-  example we derive the functor law for type substitution:
-  \begin{equation}
-  \begin{alignedat}{10}\label{eq:functor}
-    & A[\gamma\circ\delta] && {=} \\
-    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q))[\p]\cdot\q)) && {=}(\lam[]) \\
-    & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta, {\cdot}[]) \\
-    & \lam\,(A[\p]\cdot(\gamma[\p][\p^+][\langle\q\rangle]\cdot(\delta[\p][\p^+][\langle\q\rangle]\cdot\q))) && {=}([\p][^+]) \\
-    & \lam\,(A[\p]\cdot(\gamma[\p][\p][\langle\q\rangle]\cdot(\delta[\p][\p][\langle\q\rangle]\cdot\q))) && {=}([\p][\langle\rangle]) \\
-    & \lam\,(A[\p]\cdot(\gamma[\p]\cdot(\delta[\p]\cdot\q))) && {=}(\Pi\beta) \\
-    & \lam\,(\lam\,(A[\p][\p^+]\cdot(\gamma[\p][\p^+]\cdot\q))\cdot(\delta[\p]\cdot\q))\,\, && {=}(\lam[]) \\
-    & \lam\,(\lam\,(A[\p]\cdot(\gamma[\p]\cdot\q))[\p]\cdot(\delta[\p]\cdot\q)) && {=} \\
-    & A[\gamma][\delta]
-  \end{alignedat}
-  \end{equation}
-  Type formers are added by adjusting them to handle contexts built up
+  example we derive the functor law for type substitution in Appendix
+  \ref{app:functor}. Type formers are added by adjusting them to handle contexts built up
   by $\Sigma$ types, for example,
@@
   We define them in the same order, omitting the $M$ subscripts:
-  \[
-  F\,\Gamma := \diamond\ext\Gamma  \hspace{3em}
-  F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em}
-  F\,A := \El\,(A[\p]\cdot\q)  \hspace{3em}
-  F\,a := a[\p]\cdot\q
-  \]
+  $F\,\Gamma := \diamond\ext\Gamma$, 
+  $F\,\gamma := (\p,\gamma[\p]\cdot\q)$, 
+  $F\,A := \El\,(A[\p]\cdot\q)$, 
+  $F\,a := a[\p]\cdot\q$.
 %  \begin{alignat*}{10}
@@
 
-Why are we still looking at alternative ways of defining type theory?
-New definitions might be closer to actual implementations, suggest new
-implementation techniques, they might be better pedagogically, have
-more models, or contribute to a coherent definition. We elaborate on
-the latter.
-
-Defining a useable syntax of type theory internally to homotopy type
-theory (HoTT \cite{HoTTbook}) is an open problem \cite{mike,nicolai}: the (h)set-truncated
-syntax can be formalised as a QIIT, but then we cannot even define the
-$\Type$-interpretation, that is, a function from the syntax which
-interprets closed types in the metatheoretic universe $\Type$ (rather
-than an inductive recursive universe as in Subsection
-\ref{sec:standard} or \cite[Section
-  4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
-syntax of type theory can be defined using infinitely many coherence
-rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
-$\Sub$, we have an equation in the syntax on the associativity of composition of substitutions.
-If we truncate our syntax to be a set, then this is
-enough. If we don't truncate, there are two different ways of proving
-$((\gamma\circ\delta)\circ\theta)\circ\xi =
-\gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
-be the same, so in such a (wild) syntax substitutions don't form a
-set, hence they don't have decidable equality \cite{hedberg}. The
-solution is to add the pentagon law stating that these two equalities are equal, but then this introduces new
-higher equalities, which require more coherence laws, so in the end we
-have an infinite sequence of higher coherence laws that need to be
-part of the syntax. Our single substitution calculus does not have
-composition for $\Sub$, so there is no need for higher coherence laws
-like associativity. However as it currently stands, our syntax is not
-coherent, it needs set-truncation when working in HoTT. For example,
-there are two different ways to prove the following equation:
-\begin{alignat*}{10}
-  & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
-  & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
-\end{alignat*}
-In a coherent version of our syntax we either need to state that these
-two proofs are equal, or we have to remove one of them. We can remove
-the first one by limiting $[\p][\langle\rangle]$ to only apply to
-variables (this needs the introduction of a new sort for
-variables). If we do this, the minimisation of Section
-\ref{sec:minimisation} does not work anymore, so we need the extra
-four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
-$[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
-types make the syntax non-coherent again. A possible way out of this
-would be to make the equations that depend on one of these 
- conditional. But then the methods of Section
-\ref{sec:admissible} don't work anymore for re-proving the needed
-admissible equations. Also, if we only have the rule $[\p][^+]$ for
-variables, we cannot redo the termification of Section
-\ref{sec:cwf}, e.g.\ it is not clear how to prove equation
-(\ref{eq:functor}).
-
-In this paper we unfolded a particular example of the SOGAT $\ra$
-single substitution GAT translation
-\cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
-syntax is isomorphic to the parallel syntax, and we conjecture that
-our method works for any SOGAT. For our particular SOGAT, we minimised the single
-substitution syntax, and we even have that any
-single substitution model is equivalent to a parallel model in a weak
-sense. This shows that if we have enough type formers, the
-substitution calculus can be very weak, and still enough to bootstrap
-the usual theory. It needs further investigation whether we can remove
-even more structural rules, and still obtain isomorphic syntaxes. We
-would also like to understand what a good notion of weak equivalence
-is for single substitution models.
-
-TODO: named CwFs would be even closer to usual syntax.
-
+We described type theory in a minimalistic way, without referring to
+categories or parallel substitutions. Our presentation has pedagogical
+value, as illustrated in Section \ref{sec:tt}.
+%
+It is amazing that any type theory can be described with such a
+minimal substitution calculus: the lifted equations
+(\ref{eq:lifted1})--(\ref{eq:lifted4}) are not required. We use them
+when proving properties of the syntax (such as normalisation), but
+they are admissible.
+
+In the future, we would like to investigate whether the calculus can
+be minimised even more, e.g.\ to the degree that there is only one
+equation between any two terms. This would be interesting for a
+possible coherent syntax of type theory avoiding the need for
+truncation in the setting of homotopy type theory.
+
+% Why are we still looking at alternative ways of defining type theory?
+% New definitions might be closer to actual implementations, suggest new
+% implementation techniques, they might be better pedagogically, have
+% more models, or contribute to a coherent definition. We elaborate on
+% the latter.
+% 
+% Defining a useable syntax of type theory internally to homotopy type
+% theory (HoTT \cite{HoTTbook}) is an open problem \cite{mike,nicolai}: the (h)set-truncated
+% syntax can be formalised as a QIIT, but then we cannot even define the
+% $\Type$-interpretation, that is, a function from the syntax which
+% interprets closed types in the metatheoretic universe $\Type$ (rather
+% than an inductive recursive universe as in Subsection
+% \ref{sec:standard} or \cite[Section
+%   4]{DBLP:conf/popl/AltenkirchK16}). In two-level type theory the
+% syntax of type theory can be defined using infinitely many coherence
+% rules \cite{DBLP:conf/lics/Kraus21}. For example, for substitution
+% $\Sub$, we have an equation in the syntax on the associativity of composition of substitutions.
+% If we truncate our syntax to be a set, then this is
+% enough. If we don't truncate, there are two different ways of proving
+% $((\gamma\circ\delta)\circ\theta)\circ\xi =
+% \gamma\circ(\delta\circ(\theta\circ\xi))$, and nothing forces these to
+% be the same, so in such a (wild) syntax substitutions don't form a
+% set, hence they don't have decidable equality \cite{hedberg}. The
+% solution is to add the pentagon law stating that these two equalities are equal, but then this introduces new
+% higher equalities, which require more coherence laws, so in the end we
+% have an infinite sequence of higher coherence laws that need to be
+% part of the syntax. Our single substitution calculus does not have
+% composition for $\Sub$, so there is no need for higher coherence laws
+% like associativity. However as it currently stands, our syntax is not
+% coherent, it needs set-truncation when working in HoTT. For example,
+% there are two different ways to prove the following equation:
+% \begin{alignat*}{10}
+%   & (\cd\,\U)[\p][\langle a\rangle] \overset{[\p][\langle\rangle]}{=} \cd\,\U \\
+%   & (\cd\,\U)[\p][\langle a\rangle] \overset{\cd[]}{=} \cd\,(\U[\p][\langle a\rangle]) \overset{\U[]}{=} \cd\,\U
+% \end{alignat*}
+% In a coherent version of our syntax we either need to state that these
+% two proofs are equal, or we have to remove one of them. We can remove
+% the first one by limiting $[\p][\langle\rangle]$ to only apply to
+% variables (this needs the introduction of a new sort for
+% variables). If we do this, the minimisation of Section
+% \ref{sec:minimisation} does not work anymore, so we need the extra
+% four equations for types ($[\p][^+]$, $[\p][\langle\rangle]$,
+% $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$). These equations on
+% types make the syntax non-coherent again. A possible way out of this
+% would be to make the equations that depend on one of these 
+%  conditional. But then the methods of Section
+% \ref{sec:admissible} don't work anymore for re-proving the needed
+% admissible equations. Also, if we only have the rule $[\p][^+]$ for
+% variables, we cannot redo the termification of Section
+% \ref{sec:cwf}, e.g.\ it is not clear how to prove equation
+% (\ref{eq:functor}).
+% 
+% In this paper we unfolded a particular example of the SOGAT $\ra$
+% single substitution GAT translation
+% \cite{DBLP:conf/fscd/KaposiX24}. We showed that for this theory, the
+% syntax is isomorphic to the parallel syntax, and we conjecture that
+% our method works for any SOGAT. For our particular SOGAT, we minimised the single
+% substitution syntax, and we even have that any
+% single substitution model is equivalent to a parallel model in a weak
+% sense. This shows that if we have enough type formers, the
+% substitution calculus can be very weak, and still enough to bootstrap
+% the usual theory. It needs further investigation whether we can remove
+% even more structural rules, and still obtain isomorphic syntaxes. We
+% would also like to understand what a good notion of weak equivalence
+% is for single substitution models.
+% 
+% TODO: named CwFs would be even closer to usual syntax.
+% 
 \bibliography{b}
@@
   & \Lift\eta && : \mk\,(\un\,a) = a
+\end{alignat*}
+
+\section{Equation for lifting variables}
+\label{app:liftvar}
+
+Here we prove one of the roundtrips in the isomorphism
+(\ref{eq:liftvar}), the other is analogous:
+\begin{alignat*}{10}
+  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[]) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+]) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[]) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
+  & t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
+  & t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
+  & t
 \end{alignat*}
@@
 
+\section{An equation in termification}
+\label{app:functor}
+
+As an example of the equations in the termification model construction
+(Problem \ref{prob:termification}), we derive one of the functor laws
+for types:
+\begin{equation*}
+\begin{alignedat}{10}
+  & A[\gamma\circ\delta] && {=} \\
+  & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q))[\p]\cdot\q)) && {=}(\lam[]) \\
+  & \lam\,(A[\p]\cdot(\lam\,(\gamma[\p][\p^+]\cdot(\delta[\p][\p^+]\cdot\q))\cdot\q)) && {=}(\Pi\beta, {\cdot}[]) \\
+  & \lam\,(A[\p]\cdot(\gamma[\p][\p^+][\langle\q\rangle]\cdot(\delta[\p][\p^+][\langle\q\rangle]\cdot\q))) && {=}([\p][^+]) \\
+  & \lam\,(A[\p]\cdot(\gamma[\p][\p][\langle\q\rangle]\cdot(\delta[\p][\p][\langle\q\rangle]\cdot\q))) && {=}([\p][\langle\rangle]) \\
+  & \lam\,(A[\p]\cdot(\gamma[\p]\cdot(\delta[\p]\cdot\q))) && {=}(\Pi\beta) \\
+  & \lam\,(\lam\,(A[\p][\p^+]\cdot(\gamma[\p][\p^+]\cdot\q))\cdot(\delta[\p]\cdot\q))\,\, && {=}(\lam[]) \\
+  & \lam\,(\lam\,(A[\p]\cdot(\gamma[\p]\cdot\q))[\p]\cdot(\delta[\p]\cdot\q)) && {=} \\
+  & A[\gamma][\delta]
+\end{alignedat}
+\end{equation*}
+
 \end{document}

### 2025-02-21 cpp/p.tex
spellchec

@@
 where objects are contexts (lists of types), then types/terms are
-preheaves, context extension is a representability condition, so the
+presheaves, context extension is a representability condition, so the
 equations of the algebraic theory are forced by the categorical
@@
 \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
-desription of a single substitution calculus. B-systems involve
+description of a single substitution calculus. B-systems involve
 complex rules describing telescopes, weakenings and substitutions
@@
 \ref{sec:admissible} of this paper are formalised in the proof
-assistant Agda. The formalisation is availble at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
+assistant Agda. The formalisation is available at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
 Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
@@
 which is like the snoc operation for lists. Context extension takes a
-type which can have variables in the context preceeding the type.
+type which can have variables in the context preceding the type.
 \[
@@
 theory, but makes stating our theory simpler, and includes more
-models. The situation is similar to the correspondance between
+models. The situation is similar to the correspondence between
 $\Ty\,\Gamma\,i$ and $\Tm\,\Gamma\,(\U\,i)$ which could have been
@@
 The $\U\beta$, $\U\eta$ and $\Lift\beta$, $\Lift\eta$ rules express
-that the roundtrips are identities. These isomorphisms are also
+that the round-trips are identities. These isomorphisms are also
 \emph{natural}, or compatible with substitutions: these are expressed
@@
 $\Sigma$ types were introduced by a similar isomorphism, and we give
-names to the roundtrips equations according to our convention introduced in
+names to the round-trips equations according to our convention introduced in
 Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
@@
 $
-The first roundtrip equality is proven as
+The first round-trip equality is proven as
 $\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
@@
 Section \ref{sec:admissible}. As above with weakening into an
-arbitrary context, this equation holds in the syntax, bot not in an
+arbitrary context, this equation holds in the syntax, but not in an
 arbitrary model.
 %and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
-The other roundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
+The other round-trip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.
 
@@
 \end{equation}
-To derive the roundtrips, we again need admissible equations. We prove
-the first roundtrip equality in Appendix \ref{app:liftvar}, the other
+To derive the round-trips, we again need admissible equations. We prove
+the first round-trip equality in Appendix \ref{app:liftvar}, the other
 is analogous.
@@
 substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
-difficult to define the instatiation operation for $\Tms$ directly, so we also define
+difficult to define the instantiation operation for $\Tms$ directly, so we also define
 a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
@@
   \end{description}
-  The roundtrips are proven by induction on the syntax.
+  The round-trips are proven by induction on the syntax.
 \end{proof}
@@
 \end{proof}
-The other roundtrip, that is, starting with an SSC-model, termifying
+The other round-trip, that is, starting with an SSC-model, termifying
 it and comparing the result with the original SSC-model does not provide an
@@
 substitutions. However, the set of terms and types are still
-isomorphic after the roundtrip. We leave formulating the right notion
+isomorphic after the round-trip. We leave formulating the right notion
 of contextual isomorphism for SSCs as future work.
@@
 
-Here we prove one of the roundtrips in the isomorphism
+Here we prove one of the round-trips in the isomorphism
 (\ref{eq:liftvar}), the other is analogous:

### 2025-02-21 cpp/p.tex
links

@@
 
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
+%\supplement{Formalisation: \url{https://szumixie.github.io/single-subst}} % TODO: add this back
 
@@
 %
-Kaposi and Luksa \cite{luksa} defined a telescopic single substitution
-calculus for simple type theory, it can be seen as a simply typed
+Kaposi and Luksa \cite{luksa} defined a telescopic SSC for simple type theory, it can be seen as a simply typed
 version of B-systems. They show that the category of contextual models
 of their calculus is equivalent to the category of contextual simply
-typed CwFs with function space.
+typed CwFs.
 %Their equivalence result holds even without the presence of function types.
 %In the presence of function space,
-The same equivalence holds for the simply typed version of our
-substitution calculus (TODO link).
+If we have function space, the same equivalence holds for the simply typed version of our
+substitution calculus (\href{https://szumixie.github.io/single-subst/STT.Contextual.html}{\Agda}).
 
@@
 (QIITs). On paper we usually omit writing coercions, so our notation
-is close to extensional type theory. Sections \ref{sec:tt} and
+is close to extensional type theory. Definition \ref{def:tt} and Section
 \ref{sec:admissible} of this paper are formalised in the proof
-assistant Agda. The formalisation is available at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
-Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
-with postulated coercion rule (transport rule) and postulated QIITs
-with computation rules added using rewrite rules
+assistant Agda. The formalisation is available online
+(\href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}). The Agda
+logos next to definitions/theorems are links to their formalised counterparts. We use a strict Prop-valued
+\cite{DBLP:journals/pacmpl/GilbertCST19} equality type with postulated
+coercion rule (transport rule) and postulated QIITs with computation
+rules added using rewrite rules
 \cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
@@
 
-\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$]\label{def:tt}
+\begin{definition}[Single substitution calculus with $\Pi$, $\U$, $\Lift$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Syntax.html}{\Agda})]\label{def:tt}
 This concludes the definition of our basic type theory. For
@@
 \end{definition}
+
+As a sanity check for our notion of SSC-model, we defined the standard (metacircular\slash set/type) model \cite{DBLP:conf/popl/AltenkirchK16} of the
+SSC-based calculus in Agda (\href{https://szumixie.github.io/single-subst/TT.SSC.Standard.html}{\Agda}).
+
 The theory can be extended with new type and term formers in the same
@@
 
-\begin{definition}[CwF]\label{def:cwf}
+\begin{definition}[CwF (see \href{https://szumixie.github.io/single-subst/TT.CwF.Syntax.html}{\Agda})]\label{def:cwf}
   A category with $\N$-many families is defined as a category (objects
-  denoted $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
+   $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
   ($\epsilon$ denotes the unique morphism into it); for each $i$ a
@@
 define telescopes as an inductive type together with a recursive
-operation to append telescopes to contexts.
+operation to append telescopes to contexts (\href{https://szumixie.github.io/single-subst/TT.SSC.Tel.html#Tel}{\Agda}).
 \begin{alignat*}{10}
@@
 We then define a lifting operation over any telescope mutually with an
-instantiation operation on telescopes by recursion on telescopes.
+instantiation operation on telescopes by recursion on telescopes (\href{https://szumixie.github.io/single-subst/TT.SSC.Tel.html#_[_]\%E1\%B5\%80\%CB\%A1}{\Agda}).
 \[
@@
 not need to do full normalization. We define variables and $\alpha$-normal forms
-as inductive predicates in $\Prop$ as follows.
+as inductive predicates in $\Prop$ as follows (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#Var}{\Agda}).
 \begin{itemize}
@@
 We prove that the $\alpha$-normal predicate holds for $\alpha$-normal types and
-terms instantiated with $\alpha$-normal substitutions. However, this needs to be
+terms instantiated with $\alpha$-normal substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html\#[]\%E1\%B4\%BA\%E1\%B4\%BE}{\Agda}). However, this needs to be
 proved separately for weakenings and $\alpha$-normal single substitutions for
-the induction to be structural. We define predicates for weakening and
-$\alpha$-normal single substitutions as follows.
+the induction to be structural. We define predicates for weakening (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#Wk}{\Agda}) and
+$\alpha$-normal single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#NSSub}{\Agda}) as follows.
 \begin{itemize}
@@
 We define $\alpha$-normal substitutions to be the disjoint union of weakenings
-and $\alpha$-normal single substitutions.
-
-\begin{lemma}
+and $\alpha$-normal single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html#NSub}{\Agda}).
+
+\begin{lemma}[\href{https://szumixie.github.io/single-subst/TT.SSC.AlphaNorm.html\#norm\%E1\%B5\%80}{\Agda}]
   \label{thm:alpha}
@@
 instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
-freely added identity and composition operations, we do not require it to
+freely added identity and composition operations (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html#Sub*}{\Agda}), we do not require it to
 satisfy the laws of a category as we will not compare $\Sub^*$s directly for
@@
 instantiation operations, lifting operations, and substitution rules are
-redefined for $\Sub^*$.
+redefined for $\Sub^*$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html}{\Agda}).
 \begin{lemma}
@@
   \begin{itemize}
-    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
-    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
-    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
+    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5574}{\Agda})} $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
+    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5713}{\Agda})} $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
+    \item \textnormal{(\href{https://szumixie.github.io/single-subst/TT.SSC.Lift.html\#5874}{\Agda})} $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
   \end{itemize}
@@
 a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
-convert it into a sequence of instantiations of single substitutions.
+convert it into a sequence of instantiations of single substitutions (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html\#Tms}{\Agda}).
 \begin{alignat*}{10}
@@
 \end{alignat*}
-The computation of instantiating with $\Tms$ is illustrated below. On the
+The computation of instantiating with $\Tms$ (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html\#559}{\Agda}) is illustrated below. On the
 right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
@@
 All CwF operations and equations can be defined with $\Tms$ by induction, using
-Lemma \ref{thm:lifting} to avoid further induction on the syntax.
-
-\begin{problem}
+Lemma \ref{thm:lifting} to avoid further induction on the syntax (\href{https://szumixie.github.io/single-subst/TT.SSC.Parallel.html}{\Agda}).
+
+\begin{problem}[\href{https://szumixie.github.io/single-subst/TT.Isomorphism.html}{\Agda}]
   Contexts, types, and terms in the SSC syntax are isomorphic to the
@@
 % 
-% TODO: named CwFs would be even closer to usual syntax.
+% named CwFs would be even closer to usual syntax.
 % 

### 2025-02-22 fscd/p.tex
small_fix

--- 2025-02-21 cpp/p.tex
@@
 In the future, we would like to investigate whether the calculus can
-be minimised even more, e.g.\ to the degree that there is only one
-equation between any two terms. This would be interesting for a
-possible coherent syntax of type theory avoiding the need for
-truncation in the setting of homotopy type theory.
+be minimised even more, e.g.\ to the degree that there is at most one
+proof for any equation. This would be interesting for a possible
+coherent syntax of type theory avoiding the need for truncation in the
+setting of homotopy type theory.
 

### 2025-04-07 fscd/response.txt
stuff

We thank the reviewers for their constructive criticism.

----------------------------------------------------------
Response to Reviewer 1
----------------------------------------------------------

> 2. In various places you speak of "equivalent" systems, but it is
> not always clear to me what you mean by that. E.g. line 93, in what
> sense B-systems are equivalent to CwF? Certainly it is not an
> equivalence of categories, that one exists between Bsys and Csys. If
> you are only referring to the syntax, then it would be good to
> specify so. This happens again in lines 149-150.

TODO

> 7. line 233. It is not clear to me what you mean by: "We instantiate
> the type with the Sub[...] weakening."

We will remove this unclear sentence.

> 13. line 568-569. The sentence starting "as most sorts..." is a bit
> obscure to me: in Section 2 you speak of a sort of types and one of
> terms, etc.

Sorry, we should have said that most sorts in the new model are given
by specific term-sorts in the old model. We will fix this.

----------------------------------------------------------
Response to Reviewer 2
----------------------------------------------------------

> - Problem 11: What is a contextual isomorphism? Why are `F_Con` and
> `F_Sub` not required to be invertible?

A contextual isomorphism is an isomorphism between the contextual
cores of the two models (that is, when context and substitutions are
inductively generated from types and terms). In a model of type
theory, the important structure is carried by types and terms, and
contexts/substitutions are only there as helpful infrastructure
(scaffolding). This is why the notion of contextual isomorphism makes
sense: it ignores the possible differences between
contexts/substitutions. We will add a definition and more explanation.

> As mentioned in the summary above, the claim that for type theory
> with Π-types, Σ-types and a universe hierarchy, the theory of CwFs
> is "weakly equivalent" to the SSC, is not well-defined and not fully
> substantiated.

We agree that this is problematic: as we mention in line 618--619,
types and terms are isomorphic after the other roundtrip, but because
of the limitations of the single substitution calculus, it is not
clear how to define contextual cores. We will acknowledge this more
directly and will only say (in the abstract and introduction) that a
CwF structure can be defined on an SSC with Σ,Π,U.

> I think the second half of §2.1 is a bit sloppy: - For Σ-types, it
> is suddenly mentioned that naturalities in one direction are
> derivable, but the argument used seems to apply to any type former
> with η.

Yes, this indeed works for any type former with η, if it is specified
as an isomorphism (Π is not specified as such, this is why we need
admissibility, see response to the next comment).

> - On ln 400, an admissible equation is suddenly used.

In the beginning of subsection 2.1, we say that this subsection will
motivate the need for admissible equations, and this is why we use
some equations defined later. This illustrates a limitation of the SSC
formulation: although the SSC theory is very concise, it is too weak
to even define the Π-isomorphism from the traditional version of
application in an arbitrary model.

> ## Minor comments
> 
> - Somewhat important in my opinion: a theorem environment labelled
> "Problem" gives the impression of an *open* problem.

We used Vladimir Voevodsky's convention from his paper "A C-system
defined by a universe in a category" where problem/construction refers
to a proof relevant theorem/proof pair. That is, when a construction
is a theorem, it means that it has at most one proof. We will clarify.

> - The word "instantiation" is used often but not really introduced;

A substitution is a list of terms. We use instantiation for the action
of a substitution. We will clarify.

> - ln 456: "It is important ...": [...] since αβη-equality is encoded
> as metatheoretic equality, any predicate will preserve the
> conversion rules.

It is true that any predicate preserves αβη-equality, but we cannot
define a map into a non-truncated version of α-normal forms, because
that would expose differences between βη-equal terms. This is the
reason for truncation.

> - ln 568-569: Do you mean "most sort inhabitants"?

See answer to reviewer 1.

> 1. A lot of the presented material is formalized in Agda. A
> recurring theme is admissibility vs. derivability.

We directly encode admissible equations as "provable in the syntax
(initial model)" and derivability as provable in any model. Similarly
for admissible operations/sorts. One can say that in Agda, pattern
matching corresponds to admissibility, and adding a definition to a
record corrsponds to derivability. We don't use pattern matching
because we simply postulate initial models and use their eliminator
instead.

> 2. To what extent can the central result in the paper (problem 6) be
> factored through the following result/conjecture: For any CwF, the
> category of parallel substitutions is exactly presented (†) by the
> graph of single-variable weakenings and substitutions, subject to
> the equations in def. 1 / def. 7 (expressed not as equations on
> type/term actions but as equations on morphisms in the free category
> over that graph).

We believe that the reviewer's conjecture does not hold: TODO

Ehrhard is exactly SSC+moving_equations_to_morphisms+ε+comp+id+⁺functoriality.

----------------------------------------------------------
Response to Reviewer 3
----------------------------------------------------------

We agree that we don't see a direct application of the SSC technique,
but we think that such a surprisingly concise and previously unknown
definition of type theory deserves a publication. Also, the general
phenomenon that for certain theories, the syntaxes are equivalent but
not the models, is interesting (like the relationship of lambda
calculus and SK-combinator calculus), and we work out a similar
example. This calculus is a kind of "minimisation" of type theory --
and it might be the first step in this direction: in the conclusions,
we speculate about its potential use as a coherent syntax, but this is
so far only fictional.

### 2025-04-08 fscd/response.txt
finalise response

@@
 > 2. In various places you speak of "equivalent" systems, but it is
-> not always clear to me what you mean by that. E.g. line 93, in what
-> sense B-systems are equivalent to CwF? Certainly it is not an
-> equivalence of categories, that one exists between Bsys and Csys. If
-> you are only referring to the syntax, then it would be good to
-> specify so. This happens again in lines 149-150.
+> not always clear to me what you mean by that. E.g. line 93 [...]
 
-TODO
-
-> 7. line 233. It is not clear to me what you mean by: "We instantiate
-> the type with the Sub[...] weakening."
-
-We will remove this unclear sentence.
+Sorry, we were too sloppy: in line 93, we should have said contextual
+CwFs are equivalent to B-systems, and the same in lines 149-150.
 
@@
 
-Sorry, we should have said that most sorts in the new model are given
-by specific term-sorts in the old model. We will fix this.
+We should have said that most sorts in the new model are given by
+specific term-sorts in the old model.
 
@@
 theory, the important structure is carried by types and terms, and
-contexts/substitutions are only there as helpful infrastructure
-(scaffolding). This is why the notion of contextual isomorphism makes
-sense: it ignores the possible differences between
-contexts/substitutions. We will add a definition and more explanation.
+contexts/substitutions are only there as scaffolding. This is why the
+notion of contextual isomorphism makes sense: it ignores the possible
+differences between contexts/substitutions.
 
@@
 
-We agree that this is problematic: as we mention in line 618--619,
-types and terms are isomorphic after the other roundtrip, but because
-of the limitations of the single substitution calculus, it is not
-clear how to define contextual cores. We will acknowledge this more
-directly and will only say (in the abstract and introduction) that a
+We agree. As we mention in line 618--619, types and terms are
+isomorphic after the other round-trip, but because of the limitations
+of the single substitution calculus, it is not clear how to define
+contextual cores. In the abstract and introduction, we will say that a
 CwF structure can be defined on an SSC with Σ,Π,U.
-
-> I think the second half of §2.1 is a bit sloppy: - For Σ-types, it
-> is suddenly mentioned that naturalities in one direction are
-> derivable, but the argument used seems to apply to any type former
-> with η.
-
-Yes, this indeed works for any type former with η, if it is specified
-as an isomorphism (Π is not specified as such, this is why we need
-admissibility, see response to the next comment).
-
-> - On ln 400, an admissible equation is suddenly used.
-
-In the beginning of subsection 2.1, we say that this subsection will
-motivate the need for admissible equations, and this is why we use
-some equations defined later. This illustrates a limitation of the SSC
-formulation: although the SSC theory is very concise, it is too weak
-to even define the Π-isomorphism from the traditional version of
-application in an arbitrary model.
 
@@
 defined by a universe in a category" where problem/construction refers
-to a proof relevant theorem/proof pair. That is, when a construction
-is a theorem, it means that it has at most one proof. We will clarify.
+to a proof relevant theorem/proof pair. When a construction is a
+theorem, it means that it has at most one proof. We will clarify.
 
@@
 A substitution is a list of terms. We use instantiation for the action
-of a substitution. We will clarify.
+of a substitution.
 
@@
 matching corresponds to admissibility, and adding a definition to a
-record corrsponds to derivability. We don't use pattern matching
-because we simply postulate initial models and use their eliminator
-instead.
+record corresponds to derivability. We don't use pattern matching
+because we simply postulate initial models and their eliminators.
 
@@
 > factored through the following result/conjecture: For any CwF, the
-> category of parallel substitutions is exactly presented (†) by the
-> graph of single-variable weakenings and substitutions, subject to
-> the equations in def. 1 / def. 7 (expressed not as equations on
+> category of parallel substitutions is exactly presented by the graph
+> of single-variable weakenings and substitutions, subject to the
+> equations in def. 1 / def. 7 (expressed not as equations on
 > type/term actions but as equations on morphisms in the free category
@@
 
-We believe that the reviewer's conjecture does not hold: TODO
-
-Ehrhard is exactly SSC+moving_equations_to_morphisms+ε+comp+id+⁺functoriality.
+We believe that the reviewer's conjecture does not hold: if we take
+the single weakenings and substitutions from a CwF, then we generate
+the free category and quotient by the morphism-versions of the
+equations in def.1/7, then e.g. p⁺∘⟨t⟩⁺ = id holds in the CwF, but not
+in the new category. If we add all lifted versions of our equations
+(Sec.3) in the theory, then it should work.
 
@@
 
-We agree that we don't see a direct application of the SSC technique,
-but we think that such a surprisingly concise and previously unknown
-definition of type theory deserves a publication. Also, the general
-phenomenon that for certain theories, the syntaxes are equivalent but
-not the models, is interesting (like the relationship of lambda
-calculus and SK-combinator calculus), and we work out a similar
-example. This calculus is a kind of "minimisation" of type theory --
-and it might be the first step in this direction: in the conclusions,
-we speculate about its potential use as a coherent syntax, but this is
-so far only fictional.
+We also don't see a direct application of the SSC technique, but we
+think that such a surprisingly concise and previously unknown
+definition of type theory deserves publication. Also, the phenomenon
+that for certain theories, the syntaxes are equivalent but not the
+models (e.g. lambda calculus and SK-combinator calculus), is
+interesting and here we work out another example. Our calculus is a
+"minimisation" of type theory -- it might be the first step in the
+direction of a completely minimal calculus where there is only one
+proof for each equation, and thus coherent (this is so far science
+fiction).

### 2025-05-06 lfmtp/p.tex
start lfmtp submission

--- 2025-02-22 fscd/p.tex
@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[a4paper,UKenglish]{article}
 
@@
 
-\author{Ambrus Kaposi}{Eötvös Loránd University (ELTE), Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{}
-
-\author{Szumi Xie}{Eötvös Loránd University (ELTE), Hungary}{szumi@inf.elte.hu}{https://orcid.org/0009-0001-1355-1114}{}
-
-\authorrunning{A. Kaposi and Sz. Xie}
-
-\Copyright{Ambrus Kaposi and Szumi Xie}
-
-\ccsdesc[500]{Theory of computation~Type theory}
-
-\keywords{type theory, category with families, substitution calculus}
+%\author{Ambrus Kaposi}{Eötvös Loránd University (ELTE), Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{}
+
+%\author{Szumi Xie}{Eötvös Loránd University (ELTE), Hungary}{szumi@inf.elte.hu}{https://orcid.org/0009-0001-1355-1114}{}
+
+%\ccsdesc[500]{Theory of computation~Type theory}
+
+%\keywords{type theory, category with families, substitution calculus}
 
 %\supplement{Formalisation: \url{https://szumixie.github.io/single-subst}} % TODO: add this back
-
-%\acknowledgements{I want to thank \dots}%optional
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \usepackage{multicol}
-
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{amssymb}
+\usepackage{hyperref}
+
+\newtheorem{theorem}{Theorem}
 \newtheorem{problem}[theorem]{Problem}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{construction}[theorem]{Construction}
@@
 the syntaxes are equivalent. If we have some additional type formers,
-we show that SSC models are actually equivalent to CwF-based models in
-a weak sense.
+we show that an SSC model gives rise to a CwF which is a retract.
 \end{abstract}

### 2025-05-14 lfmtp/p.tex
use EPTCS style
stuff

@@
-\documentclass[a4paper,UKenglish]{article}
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
+\documentclass[submission,copyright,creativecommons]{eptcs}
+\providecommand{\event}{LFMTP 2025} % Name of the event you are submitting to
+
+\usepackage{iftex}
+
+\ifpdf
+  \usepackage{underscore}         % Only needed if you use pdflatex.
+  \usepackage[T1]{fontenc}        % Recommended with pdflatex
+\else
+  \usepackage{breakurl}           % Not needed if you use pdflatex only.
+\fi
 
 \title{Type theory with single substitutions}
-
-%\author{Ambrus Kaposi}{Eötvös Loránd University (ELTE), Hungary}{akaposi@inf.elte.hu}{https://orcid.org/0000-0001-9897-8936}{}
-
-%\author{Szumi Xie}{Eötvös Loránd University (ELTE), Hungary}{szumi@inf.elte.hu}{https://orcid.org/0009-0001-1355-1114}{}
+\author{Ambrus Kaposi
+\institute{Eötvös Loránd University (ELTE)\\
+Budapest, Hungary}
+\email{akaposi@inf.elte.hu}
+\and
+Szumi Xie
+\institute{Eötvös Loránd University (ELTE)\\
+Budapest, Hungary}
+\email{szumi@inf.elte.hu}
+}
+\def\titlerunning{Type theory with single substitutions}
+\def\authorrunning{A. Kaposi \& Sz. Xie}
 
@@
 \usepackage{amssymb}
-\usepackage{hyperref}
 
@@
 \begin{document}
-
 \maketitle
@@
 CwF-based models are models of SSC, but not
-necessarily the other way. While B-systems are equivalent to CwFs (and
+necessarily the other way. While B-systems are equivalent to contextual CwFs (and
 C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
@@
 that we have fewer and less general equations, resulting in the fact
-that we have more models than B-systems. However in the syntax of our
+that we have more models than B-systems. A further difference is that contexts
+in B-systems are inductively defined (we call such models \emph{contextual}).
+However in the syntax of our
 theory, all the rules of B-systems are admissible. We describe the
@@
 substitutions are lists of terms. CwFs are equivalent to Ehrhard's
-calculus, and so are contextual categories/C-systems
-\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
-the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
+calculus and the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18}. Contextual versions of these are equivalent to contextual categories/C-systems
+\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18}, and
 B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.
@@
 \cite{DBLP:conf/fscd/KaposiX24} translate SOGATs to GATs: a model of a
-SOGAT then is a model of the GAT the we obtain by their
+SOGAT then is a model of the GAT that we obtain by their
 translation. GATs come with well-behaved metatheory: a category of
@@
 context. We would like to describe the operation providing the last
-variable in a context, the zero De Bruijn index \cite{debruijn}, we
-denote it by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
+variable in a context, first of all we need an operation providing the last
+variable in the context. This is the zero De Bruijn index \cite{debruijn}, which we
+denote by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
 extended context, and it should have type $A$, but the issue is that
@@
 A)\,i$ and define $A' := A[\p]$. Instead of introducing just
-$\blank[\p]$, we generalise a bit and add a new sort $\Sub$ which for
-now only has the single element $\p$ and the operation
-$\blank[\blank]$ is called \emph{instantiation}. We instantiate the
-type with the $\Sub\,\Delta\,\Gamma$ weakening.  Elements of $\Sub$
+$\blank[\p]$, we generalise a bit and add a new sort $\Sub$. For
+now, $\Sub$ only has the single element $\p$ and we introduce the operation
+$\blank[\blank]$ called \emph{instantiation}. Elements of $\Sub$
 will later be called substitutions, hence the name, but for now, we
@@
 $\Sub$, and there are no equations on $\Sub$. An element of $\Sub$ is
-either a single weakening $\p$ lifted a couple of times, or a single
-substitution lifted a couple of times. We call elements of $\Sub$
+either a single weakening $\p$ lifted a finite number of times, or a single
+substitution lifted a finite number of times. We call elements of $\Sub$
 substitutions for simplicity.
@@
 \begin{definition}[CwF (see \href{https://szumixie.github.io/single-subst/TT.CwF.Syntax.html}{\Agda})]\label{def:cwf}
-  A category with $\N$-many families is defined as a category (objects
+  A \emph{category with $\N$-many families} is defined as a category (objects
    $\Con$, morphisms $\Sub$) with a terminal object $\diamond$
@@
 
-A type or term is $\alpha$-normal if it does not contain instantiation
+A type or term is \emph{$\alpha$-normal} if it does not contain instantiation
 operations, except at the leaves of the syntax tree as variables. However,
@@
   \end{itemize}
-  Note that the later equations depend on the the earlier ones.
+  Note that the later equations depend on the earlier ones.
 \end{lemma}
@@
 $
+Thus, as the input of equation $[\p][^+]'$ holds all the time, we get this equation for every term.
 Equation $[\q][^+] : \q[\gamma^+] = \q$ also only makes sense if we
@@
   \El\,((\cd\,B)[\langle a\rangle][\gamma]) \overset{\Pi\beta}{=}
-  \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) \overset{\cdot[]'\,\U[])}{=}
+  \El\,((\lam\,(\cd\,B)\cdot a)[\gamma]) \overset{\cdot[]'\,\U[]}{=}
   \El\,((\lam\,(\cd\,B)[\gamma])\cdot (a[\gamma])) \overset{\lam[]}{=}
@@
 \cite{kaposi-shallow} we call it termification, as most sorts in the
-new model are terms in the old model.
+new model are given by the sort of terms in the old model.
 \begin{problem}[Termification]\label{prob:termification}
@@
 substitutions. Such a map is not definable only using single
-substitutions. However, the set of terms and types are still
+substitutions (a $\Sub$ is either a lifted weaking or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$).
+However, the set of terms and types are still
 isomorphic after the round-trip. We leave formulating the right notion
@@
 % 
+\bibliographystyle{eptcs}
 \bibliography{b}

### 2025-05-16 lfmtp/p.tex
Use more readable Agda logo
stuff

@@
 \newcommand{\Agda}{%
-  \raisebox{-.15em}{\includegraphics[height=.9em]{agda-logo.pdf}}%
+  \raisebox{-.05em}{\includegraphics{agda-logo-small.pdf}}%
 }
@@
   stating that e.g.\ conversion is a congruence with respect to all
-  operations (for an intrinsic presentation with explicit conversion
+  operations (for an alternative intrinsic presentation with explicit conversion
   relation see \cite{chapman09eatitself}).
@@
   u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
-in the context of $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is
-called a parallel substitution, where the terms $u_0,\dots,u_n$ are
+in the context of $t$. We call $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$
+a parallel substitution, where the terms $u_0,\dots,u_n$ are
 all in the same context. Parallel substitutions are very natural in
@@
 theory have been formalised \cite{cohtt}. Without QIITs, tricks such
-as shallow embedding the syntax can be used to formalise metatheoretic
+as shallowly embedding the syntax can be used to formalise metatheoretic
 results about type theory such as gluing
@@
 with only two sorts and no operations or equations, we obtain the GAT
-of category with families (CwF). If we apply their single substitution
+of categories with families (CwFs). If we apply their single substitution
 translation to the SOGAT (\ref{eq:tt}), we obtain Definition
@@
 substitutions. We won't have more ways to introduce elements of
-$\Sub$, and there are no equations on $\Sub$. An element of $\Sub$ is
+$\Sub$, and there are no equations on $\Sub$,
+as the equations relating substitutions are defined for their action
+on types and terms. An element of $\Sub$ is
 either a single weakening $\p$ lifted a finite number of times, or a single
@@
 (with different implicit arguments for the $\lam$s on the left and right hand side), so we are able to apply $\cd\,A'$
-and $a'$ just as before. What we cannot do is to directly weaken a
+and $a'$ just as before. What we cannot derive is to directly weaken a
 term in context $\diamond$ to an arbitrary context $\Gamma$, where
@@
 The corresponding equations on terms are also needed. Note that we do not have
-the non-lifted term versions of the last two equations in SSC.
+the non-lifted \emph{term} versions of the last two equations in SSC.
 
@@
 \end{proof}
-This also implies that the initial SSC is the initial CwF.
+This also implies that the initial SSC model is the initial CwF.
 
@@
   Given a CwF (with type formers) $M$, let $M'$ denote the CwF
-  obtained by first seeing it as an SSC and then termifying it. We
+  obtained by first seeing it as an SSC model and then termifying it. We
   construct a contextual isomorphism between $M'$ and $M$.

### 2025-05-16 lfmtp/p.tex
fix build on nixos

@@
 extrinsic single substitution world, Voevodsky introduced B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
+\cite{bc} which are an algebraic
 description of a single substitution calculus. B-systems involve
@@
 necessarily the other way. While B-systems are equivalent to contextual CwFs (and
-C-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more
+C-systems \cite{bc}), there are more
 SSC models than CwF-based models. This relationship is like that
@@
 \cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18}, and
-B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.
+B-systems \cite{bc}.
 
@@
 results about type theory such as gluing
-\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
+\cite{glue}, and its special cases
 canonicity and parametricity \cite{kaposi-shallow}.
@@
 and logical frameworks
-\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}. Second-order
+\cite{DBLP:journals/jacm/HarperHP93,beluga}. Second-order
 generalised algebraic theories (SOGATs
@@
 %
+
 \bibliographystyle{eptcs}
@@
 In this section, we describe the relationship between B-systems
-\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} and our Definition
+\cite{bc} and our Definition
 \ref{def:tt} in detail.

### 2025-05-16 lfmtp/p.tex
squashing

@@
 
-\title{Type theory with single substitutions}
+\title{Type theory with single substitutions\thanks{This research was supported by the the HOTT ERC Grant 101170308.}}
 \author{Ambrus Kaposi
@@
 %\keywords{type theory, category with families, substitution calculus}
-
-%\supplement{Formalisation: \url{https://szumixie.github.io/single-subst}} % TODO: add this back
 
@@
 the syntaxes are equivalent. If we have some additional type formers,
-we show that an SSC model gives rise to a CwF which is a retract.
+we show that an SSC model gives rise to a CwF.
 \end{abstract}
@@
   the equations. For example, it is not possible to define a printing
-  function which distinguishes the terms $(\lambda x.x)\,1$ and $1$,
+  function which distinguishes the terms $(\lambda x\ldotp x)\,(\lambda y\ldotp y)$ and $\lambda y\ldotp y$,
   as these are convertible. However, it is still possible to define
@@
 with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
-Awodey's natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
+natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
 categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
@@
 
-We introduce a technology for proving admissible rules in our
-theory. We define $\alpha$-normal forms and show
-$\alpha$-normalisation: every term admits an $\alpha$-normal form.
+We introduce a technology for proving admissible rules. We define $\alpha$-normal forms and show that every term admits an $\alpha$-normal form.
 Thus, proving something for $\alpha$-normal forms implies it for all terms.
@@
 function space. We use induction on $\alpha$-normal forms to define
-parallel substitutions and prove all the rules of CwFs in the syntax
+parallel substitutions and prove all CwF rules in the syntax
 of SSC.
@@
 type theory with a Coquand-universe and $\Pi$ types, we can do better:
-the 4 (conditional) equations are enough, because we can derive the
+4 (conditional) equations are enough, because we can derive the
 other 4 equations using the fact that every type is represented by a
@@
 
-If in addition to having a universe and $\Pi$ types, our type theory
+If in addition to having a universe and $\Pi$ types, our theory
 also supports unit and $\Sigma$ types, we can actually derive a
@@
 dependent types. Then the SSC is just there for bootstrapping
-purposes: SSC formulates $\Sigma$, $\Pi$, $\U$, and then
-we use these to define a parallel substitution calculus. In summary,
-although SSC models and CwFs without any type formers are not
-equivalent, if we equip them with enough type formers, they become
-(weakly) equivalent.
+purposes: SSC formulates $\Sigma$, $\Pi$, $\U$, and then we use these
+to define a parallel substitution calculus. In summary, although SSC
+models without any type formers are weaker than CwFs, if we equip them
+with enough type formers, the SSC model gives rise to a CwF. Doing a
+roundtrip CwF $\longrightarrow$ SSC $\longrightarrow$ CwF results in a
+contextually isomorphic CwF.
 
@@
 internal to extensional type theory, and most results were formalised
-in Agda. %Some results in this paper were presented at the TYPES 2024
-%conference \cite{singleTypes}.
+in Agda (see Subsection \ref{sec:metatheory}). Some results in this
+paper were presented at the TYPES 2024 conference \cite{singleTypes}.
 
@@
   the equations which results in an isomorphic theory.
-\item[Section \ref{sec:cwf}] For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
-  single and parallel models are equivalent in a weak sense.
+\item[Section \ref{sec:cwf}] For type theory with $\Pi$, $\U$, $\top$,
+  $\Sigma$, we show that a CwF structure is derivable.
 \end{description}
@@
 
-\vspace{-1.1em}
+\vspace{-1em}
 \subparagraph*{Ehrhard's calculus.}
@@
 \cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
-$\blank,\blank$ operations which make it more apparent that
+$(\blank,\blank)$ operations which make it more apparent that
 substitutions are lists of terms. CwFs are equivalent to Ehrhard's
@@
 \cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18}, and
-B-systems \cite{bc}.
-
-\vspace{-1.1em}
+B-systems \cite{bc}. For the precise statements of equivalences, see e.g.\ \cite{DBLP:conf/aplas/AhrensLN24}.
+
+\vspace{-1em}
 \subparagraph*{Formalisations of type theory.}
@@
 
-\vspace{-1.1em}
+\vspace{-1em}
 \subparagraph*{SOGATs}
@@
 modelled by the metatheoretic function space. For example, a type
-theory with $\Pi$, Coquand-universes and a universe level lifting
+theory with $\Pi$, Coquand-universes \cite{coquandUniverse} and a universe level lifting
 operation is described by the following SOGAT (where $\cong$ means
@@
 substitutions. If we apply their parallel translation to the SOGAT
+\vspace{-0.5em}
 \begin{equation}\label{eq:tytm}
@@
 
-\vspace{-0.7em}
+\vspace{-0.5em}
 \subsection{Metatheory and formalisation}\label{sec:metatheory}
@@
 saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
-and $g$ as a constructor).
-
+and $g$ as a constructor). Following Voevodsky \cite{voevodsky2015csystemdefineduniversecategory}, we call
+a proof relevant theorem a \emph{problem} and its proof a \emph{construction}.
+The words theorem, lemma, proof refer to propositions.
+
+\vspace{-0.5em}
 \section{Single substitution syntax}
@@
 an implicit argument of $\Tm$.
-\[
+\begin{equation*}
 \Con : \Set \hspace{3em} 
@@
 \Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
-\]
+\end{equation*}
 Just as lists have two constructors, there are two ways to form
@@
 \begin{alignat*}{10}
-  & \Sub && : \Con\ra\Con\ra\Set && \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
-  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{3em} && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
+  & \Sub && : \Con\ra\Con\ra\Set &&\hspace{1.5em}
+  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{1.5em}
+  \p && : \Sub\,(\Gamma\ext A)\,\Gamma \hspace{1.5em}
+  && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
 \end{alignat*}
@@
 to be instantiated by the weakening $\gamma$ in the domain context in
-order to fit in. Now we can explain how instantiation acts on $\Pi$ by
+order to fit in. Now we explain how instantiation acts on $\Pi$ by
 the following equation.
@@
 
-The functions $\lambda x.x$ and $\lambda x\,y.x$ can be defined in our
+The functions $\lambda x\ldotp x$ and $\lambda x\,y\ldotp x$ can be defined in our
 syntax as $\lam\,\q : \Tm\,\Gamma\,(A\Ra A)$ and
@@
 % (\ref{eq:tt})).
-\begin{alignat*}{10}
-  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && \U[] && : (\U\,i)[\gamma] = \U\,i                           && \U\beta && : \El\,(\cd\,A) = A \\        
-  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{3em} && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
-  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
+% \begin{alignat*}{10}
+%   & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && \U[] && : (\U\,i)[\gamma] = \U\,i                           && \U\beta && : \El\,(\cd\,A) = A \\        
+%   & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{3em} && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
+%   & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
+% \end{alignat*}
+\begin{alignat*}{10}
+  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \hspace{1.5em} && \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i                          && \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \hspace{1.5em} && \U\beta && : \El\,(\cd\,A) = A \\
+  & \U[] && : (\U\,i)[\gamma] = \U\,i                  && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{1.5em} && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])                && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
 \end{alignat*}
@@
 define the polymorphic identity function.
-\begin{alignat*}{10}
-  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])\hspace{3em} && \Lift\beta && : \un\,(\mk\,a) = a \\
-  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma])              && \Lift\eta && : \mk\,(\un\,a) = a \\
-  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma])
+% \begin{alignat*}{10}
+%   & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])\hspace{3em} && \Lift\beta && : \un\,(\mk\,a) = a \\
+%   & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma])              && \Lift\eta && : \mk\,(\un\,a) = a \\
+%   & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma])
+% \end{alignat*}
+\begin{alignat*}{10}
+  & \Lift : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i) \hspace{.9em} && \mk : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) \hspace{.9em} && \un : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A \hspace{.9em} && \un\,(\mk\,a) = a \\
+  & (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])                   && (\mk\,a)[\gamma] = \mk\,(a[\gamma])                            && (\un\,a)[\gamma] = \un\,(a[\gamma])                            && \mk\,(\un\,a) = a
 \end{alignat*}
@@
 The theory can be extended with new type and term formers in the same
-fashion: each operation has to be indexed over arbitrary contexts and come
-equipped with a substitution rule. There is no need to add more
+fashion: each operation has to be indexed over contexts and come
+ with a substitution rule. There is no need to add more
 structural (substitution calculus) rules.\footnote{In \cite{DBLP:conf/fscd/KaposiX24},
@@
     & \rlap{$
-  \top : \Ty\,\Gamma\,0 \hspace{3em}
-  \top[] : \top[\gamma] = \top\hspace{3em}
-  \top\eta : t = \tt \hspace{3em}
-  \tt : \Tm\,\Gamma\,\top \hspace{3em}
+  \top : \Ty\,\Gamma\,0 \hspace{3.9em}
+  \top[] : \top[\gamma] = \top\hspace{3.9em}
+  \top\eta : t = \tt \hspace{3.9em}
+  \tt : \Tm\,\Gamma\,\top \hspace{3.9em}
   \tt[] : \tt[\gamma] = \tt
-      $} \\
-    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
-    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \\
-    & \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
-    & {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
+      $} \\[0.5em]
+    & \Sigma && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i && \Sigma[] && : (\Sigma\,A\,B)[\gamma] = \Sigma\,(A[\gamma])\,(B[\gamma^+]) \\
+    & \fst,\snd && : \Tm\,\Gamma\,(\Sigma\,A\,B)\cong(a:\Tm\,\Gamma\,A)\times\Tm\,\Gamma\,(B[\langle a\rangle]) : \blank{,}\blank \hspace{3.2em} && {,}[] && : (a,b)[\gamma] = (a[\gamma],b[\gamma])
   \end{alignat*}
@@
 In this subsection, we show how to use the above defined calculus, and
-motivate the next section on admissible equations.
+illustrate its deficiencies: some important equations are not
+derivable in any model, only admissible in the syntax. These equations
+will be proven in Section \ref{sec:admissible}.
 
@@
 $
-(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
+(\lambda t\ldotp t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
 $
@@
 \begin{equation}\label{eq:liftvar}
-(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
+(\lambda t\ldotp t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t\ldotp t[\p^+][\langle\mk\,\q\rangle])
 \end{equation}
@@
   & \diamond && : \Tel\,\Gamma && \Gamma+\diamond && := \Gamma \\
-  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma \hspace{3em} && \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A
+  & \blank\ext\blank && : (\Omega:\Tel\,\Gamma)\ra\Ty\,(\Gamma+\Omega)\ra\Tel\,\Gamma \hspace{3em} && \Gamma+(\Omega\ext A) && := (\Gamma+\Omega)\ext A \hspace{3em}
 \end{alignat*}
@@
 $\lam$. Notably, we do not state that instantiated types/terms are
-$\alpha$-normal (except variables). It is important to truncate the
-$\alpha$-normal predicate to be a proposition to ensure that it preserves the
-$\beta$/$\eta$ conversion rules.
+$\alpha$-normal (except variables). We truncate the
+$\alpha$-normal predicate to be a proposition to allow interpreting into it from the
+quotiented syntax. Without truncation, interpretation into $\alpha$-normal forms
+would expose differences between $\beta$/$\eta$-equal terms (and such an interpretation is not definable).
 
@@
 instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
-freely added identity and composition operations (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html#Sub*}{\Agda}), we do not require it to
-satisfy the laws of a category as we will not compare $\Sub^*$s directly for
-equality, only between types and terms instantiated with $\Sub^*$s. All
+freely added identity and composition operations (\href{https://szumixie.github.io/single-subst/TT.SSC.Path.html#Sub*}{\Agda}). We do not require it to
+satisfy the category laws as we will not compare $\Sub^*$s directly for
+equality. All
 instantiation operations, lifting operations, and substitution rules are
@@
   x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
-  the variable. Then the two latter equations can be proven by mutual induction
+  the variable. Then the two latter equations are proven by mutual induction
   on $\alpha$-normalized types and terms, still assuming the first equation.
@@
 stated both for types and terms (here $B$ is a type and $b$ is a term of type $B$):
-\[
-B[\p][\gamma^+] = B[\gamma][\p] \hspace{4em} b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]
-\]
+$B[\p][\gamma^+] = B[\gamma][\p]$ and $b[\p][\gamma^+] =_{[\p][^+]} b[\gamma][\p]$. 
 We even need the first equation to typecheck the second one. We made
@@
 \begin{alignat*}{10}
-  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
+  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \hspace{3em}
   & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b
 \end{alignat*}
-But first we need to check that the usage of $\Pi\beta$ above when
+But first we need to verify that the instance of $\Pi\beta$ used when
 proving $B[\langle a\rangle][\gamma] = B[\gamma^+][\langle
-  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming a
+  a[\gamma]\rangle]$ is derivable from $\Pi\beta'$. Assuming
 $\hat{B} : \Tm\,(\Gamma\ext A)\,\U$, we argue
@@
 $
-Now, for any type $B : \Ty\,(\Gamma\ext A)$, we derive the assumption of $\Pi\beta'/\Pi\eta'$:
+Now, for any type $B : \Ty\,(\Gamma\ext A)$, we derive the input of $\Pi\beta'/\Pi\eta'$ by
 $
@@
 $
-We summarise this section by formally stating what we defined.
+We summarise this section by formally stating the above.
 
@@
 In this section we show that if the single substitution calculus has
-certain type formers, then it is actually weakly equivalent to CwFs
-with the same type formers. The idea is that
+certain type formers, then a CwF structure is derivable in any model.
+The idea is that
 \begin{itemize}
@@
   Every CwF (with type formers $\Pi, \dots, \Sigma$) gives rise to an SSC (with
-  the same type formers, see Definition \ref{def:sigma}).
+  the same type formers).
 \end{problem}
@@
   \begin{alignat*}{10}
-    & \Con && := \Ty\,\diamond                                           && \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U) && \p && := \lam\,(\fst\,\q) \\                     
-    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma)        && A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q)) && \q && := \lam\,(\snd\,\q) \\
-    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \hspace{1.5em} && \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q)))\hspace{1.1em} \\   
-    & \id && := \lam\,\q                                                 && a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q)) \\                      
-    & \diamond && := \top                                                && \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\                   
-    & \epsilon && := \lam\,\tt                                           && (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q)
+    & \Con && := \Ty\,\diamond                                                          && \epsilon && := \lam\,\tt                                                                      && \Gamma\ext A && := \Sigma\,\Gamma\,(\El\,(A[\p]\cdot\q)) \\
+    & \Sub\,\Delta\,\Gamma && := \Tm\,\diamond\,(\Delta\Ra\Gamma)                       && \Ty\,\Gamma && := \Tm\,\diamond\,(\Gamma\Ra\U)                                                && (\gamma,a) && := \lam\,(\gamma[\p]\cdot\q,a[\p]\cdot\q) \\
+    & \gamma\circ\delta && := \lam\,(\gamma[\p]\cdot(\delta[\p]\cdot\q)) \hspace{1.4em} && A[\gamma] && := \lam\,(A[\p]\cdot(\gamma[\p]\cdot\q))                                         && \p && := \lam\,(\fst\,\q) \\
+    & \id && := \lam\,\q                                                                && \Tm\,\Gamma\,A && := \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \hspace{1.4em}       && \q && := \lam\,(\snd\,\q) \\
+    & \diamond && := \top                                                               && a[\gamma] && := \lam\,(a[\p]\cdot(\gamma[\p]\cdot\q))
   \end{alignat*}
@@
 \end{proof}
-The notion of equivalence between CwFs that we use is contextual
-isomorphism \cite{DBLP:journals/corr/abs-2211-07487}: a weak
-CwF-morphism (pseudomorphism) which is bijective on types and
-terms. Note that a contextual isomorphism preserves all type formers
-specified by universal properties.
-\begin{problem}
-  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
-  obtained by first seeing it as an SSC model and then termifying it. We
-  construct a contextual isomorphism between $M'$ and $M$.
-\end{problem}
-\begin{proof}[Construction]
-  We denote the components of the contextual isomorphism $F$ as follows.
-  \begin{alignat*}{10}
-    & F : \Con_{M'}\ra\Con_M && F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
-    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma)\hspace{3em} && F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
-  \end{alignat*}
-  We define them in the same order, omitting the $M$ subscripts:
-  $F\,\Gamma := \diamond\ext\Gamma$, 
-  $F\,\gamma := (\p,\gamma[\p]\cdot\q)$, 
-  $F\,A := \El\,(A[\p]\cdot\q)$, 
-  $F\,a := a[\p]\cdot\q$.
-%  \begin{alignat*}{10}
-%    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
-%    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
-%    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
-%    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
-%  \end{alignat*}
-  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
-  is a functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
-  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
-  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
-  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack
-\end{proof}
-The other round-trip, that is, starting with an SSC-model, termifying
-it and comparing the result with the original SSC-model does not provide an
-isomorphism because we can't define the map $F :
-\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
-\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
-substitutions. Such a map is not definable only using single
-substitutions (a $\Sub$ is either a lifted weaking or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$).
-However, the set of terms and types are still
-isomorphic after the round-trip. We leave formulating the right notion
-of contextual isomorphism for SSCs as future work.
+In Appendix \ref{sec:functor}, we show that the roundtrip CwF
+$\longrightarrow$ SSC
+$\xrightarrow{\text{termification}}$ CwF results in a
+contextually isomorphic CwF.
 
@@
 
+\newpage
+
 \appendix
 
-\section{Detailed comparison with B-systems}
+\section{Detailed comparison of our SSC-calculus and B-systems}
 \label{app:bsys}
@@
 
-\section{Listing of Definition \ref{def:tt}}
+\newpage
+
+\section{Listings for Section \ref{sec:tt}}
 \label{app:tt}
@@
 In Section \ref{sec:tt}, we presented the single substitution syntax
-interleaved with explanations, here we list all the rules in one
-place, for reference.
-
-The core substitution calculus:
+interleaved with explanations (Definition \ref{def:tt}), here we list
+all the rules in one place, for reference.
+\vspace{-0.5em}
+\begin{multicols}{2}
+\noindent The core substitution calculus:\vspace{-.5em}
 \begin{alignat*}{10}
@@
 \end{alignat*}
-Rules for individual type formers.
-\begin{alignat*}{10}
-  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \\
+
+\newcolumn
+\noindent Rules for individual type formers.\vspace{-.5em}
+\begin{alignat*}{10}
+  & \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i \hspace{10em} \\
   & \Pi[] && : (\Pi\,A\,B)[\gamma] = \Pi\,(A[\gamma])\,(B[\gamma^+]) \\
@@
   & \lam[] && : (\lam\,b)[\gamma] = \lam\,(b[\gamma^+]) \\
-  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra\Tm\,\Gamma\,(B[\langle a\rangle]) \\
+  & \blank\cdot\blank && : \Tm\,\Gamma\,(\Pi\,A\,B)\ra(a:\Tm\,\Gamma\,A)\ra \\
+  & && \hphantom{{}:{}} \Tm\,\Gamma\,(B[\langle a\rangle]) \\
   & {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma]) \\
@@
 \end{alignat*}
-
-\section{Equation for lifting variables}
-\label{app:liftvar}
-
-Here we prove one of the round-trips in the isomorphism
-(\ref{eq:liftvar}), the other is analogous:
-\begin{alignat*}{10}
-  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[]) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+]) \\
-  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3}) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[]) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
-  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
-  & t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
-  & t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
-  & t
-\end{alignat*}
-
-\section{Listing of Definition \ref{def:min}}
+\end{multicols}
+
+\vspace{-0.5em}
+\noindent We prove one of the round-trips in the isomorphism for lifting
+variables (\ref{eq:liftvar}), the other is analogous:
+\begin{alignat*}{10}
+  & t[\p^+][\langle\un\,\q\rangle][\p^+][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3})                                  &&    t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\mk\,\q)\rangle] && {=}(\Lift\beta) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q[\p^+]\rangle][\langle\mk\,\q\rangle] && {=}(\un[])                                   &&    t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted1}) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,(\q[\p^+])\rangle][\langle\mk\,\q\rangle] && {=}(\q[^+])                                &&    t[\p^+][\p^+][\langle\mk\,\q\rangle^+][\langle\q\rangle] && {=}(\ref{eq:lifted2}) \\
+  & t[\p^+][{\p^+}^+][\langle\un\,\q\rangle][\langle\mk\,\q\rangle] && {=}(\ref{eq:lifted3})                              &&    t[\p^+][\langle\q\rangle] && {=}(\ref{eq:lifted4}) \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,\q[\langle\mk\,\q\rangle]\rangle] && {=}(\un[])                &&    t                                                                                             \\
+  & t[\p^+][{\p^+}^+][\langle\mk\,\q\rangle^+][\langle\un\,(\q[\langle\mk\,\q\rangle])\rangle] && {=}(\q[\langle\rangle]) \hspace{4em} && 
+\end{alignat*}
+
+\section{Listing for Section \ref{sec:minimisation}}
 \label{app:minimisation}
@@
 For reference, we list the rules for the minimised single substitution
-syntax defined in Section \ref{sec:minimisation}.
+syntax (Definition \ref{def:min}). \\
+\begin{minipage}{0.6\textwidth}
 \begin{alignat*}{10}
@@
   & && \hphantom{{}:{}} (t\cdot a)[\gamma] =_e (t[\gamma])\cdot(a[\gamma]) \\
-  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra t =_e \lam\,(t[\p]\cdot\q) \\
-  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra (\lam\,b)[\p]\cdot\q =_e b \\
+\end{alignat*}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\begin{alignat*}{10}
+  & \Pi\eta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra \\
+  & && \hphantom{{}:{}} t =_e \lam\,(t[\p]\cdot\q) \\
+  & \Pi\beta' && : (e:B[\p^+][\langle\q\rangle] = B)\ra \\
+  & && \hphantom{{}:{}} (\lam\,b)[\p]\cdot\q =_e b \\
   & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i) \\
@@
   & \Lift\beta && : \un\,(\mk\,a) = a \\
-  & \Lift\eta && : \mk\,(\un\,a) = a
-\end{alignat*}
-
-\section{An equation in termification}
+  & \Lift\eta && : \mk\,(\un\,a) = a \\
+  & \\
+  & \\
+  & \\
+\end{alignat*}
+\end{minipage}
+
+\newpage
+
+\section{Listings for Section \ref{sec:cwf}}
 \label{app:functor}
@@
 
+We show that the roundtrip CwF $\longrightarrow$ SSC
+$\xrightarrow{\text{termification}}$ CwF results in a contextually
+isomorphic CwF. A contextual isomorphism
+\cite{DBLP:journals/corr/abs-2211-07487} is a weak CwF-morphism
+(pseudomorphism, i.e.\ context extension and the empty context are
+only preserved up to isomorphism) which is bijective on types and
+terms. Note that a contextual isomorphism preserves all type formers
+specified by universal properties.
+\begin{problem}
+  Given a CwF (with type formers) $M$, let $M'$ denote the CwF
+  obtained by first seeing it as an SSC model and then termifying it. We
+  construct a contextual isomorphism between $M'$ and $M$.
+\end{problem}
+\begin{proof}[Construction]
+  We denote the components of the contextual isomorphism $F$ as follows.
+  \begin{alignat*}{10}
+    & F : \Con_{M'}\ra\Con_M && F : \Ty_{M'}\,\Gamma \cong \Ty_M\,(F\,\Gamma) \\
+    & F : \Sub_{M'}\,\Delta\,\Gamma\ra\Sub_M\,(F\,\Delta)\,(F\,\Gamma)\hspace{3em} && F : \Tm_{M'}\,\Gamma\,A \cong \Tm_M\,(F\,\Gamma)\,(F\,A)
+  \end{alignat*}
+  We define them in the same order, omitting the $M$ subscripts:
+  $F\,\Gamma := \diamond\ext\Gamma$, 
+  $F\,\gamma := (\p,\gamma[\p]\cdot\q)$, 
+  $F\,A := \El\,(A[\p]\cdot\q)$, 
+  $F\,a := a[\p]\cdot\q$.
+%  \begin{alignat*}{10}
+%    & F : \Ty\,\diamond\ra\Con && F : \Tm\,\diamond\,(\Gamma\Ra\U) \cong \Tm\,(\diamond\ext\Gamma)\,\U \cong \Ty\,(\diamond\ext\Gamma) \\
+%    & F\,\Gamma := \diamond\ext\Gamma && F\,A := \El\,(A[\p]\cdot\q) \\
+%    & F : \Tm\,\diamond\,(\Delta\Ra\Gamma) \ra \Sub\,(\diamond\ext\Delta) (\diamond\ext\Gamma) && F : \Tm\,\diamond\,(\Pi\,\Gamma\,(\El\,(A[\p]\cdot\q))) \cong \Tm\,(\diamond\ext\Gamma)\,(\El\,(A[\p]\cdot\q)) \\
+%    & F\,\gamma := (\p,\gamma[\p]\cdot\q) \hspace{3em} && F\,a := a[\p]\cdot\q
+%  \end{alignat*}
+  It is easy to check that $F$ preserves the CwF structure, i.e.\ it
+  is a functor, $\epsilon : \Sub\,(F\,\diamond)\,\diamond$ is an
+  isomorphism, $F\,(A[\gamma]) = F\,A[F\,\gamma]$, $F\,(a[\gamma]) =
+  F\,a[F\,\gamma]$ and $(F\,\p,F\,\q) : \Sub\,(F\,(\Gamma\ext
+  A))$ $(F\,\Gamma\ext F\,A)$ is an isomorphism. % hack
+\end{proof}
+The other round-trip, that is, starting with an SSC-model, termifying
+it and comparing the result with the original SSC-model does not provide an
+isomorphism because we can't define the map $F :
+\Tm\,\diamond\,(\Delta\Ra\Gamma) \ra
+\Sub\,(\diamond\ext\Delta)\,(\diamond\ext\Gamma)$ on
+substitutions. Such a map is not definable only using single
+substitutions (a $\Sub$ is either a lifted weaking or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$).
+However, the set of terms and types are still
+isomorphic after the round-trip. We leave formulating the right notion
+of contextual isomorphism for SSCs as future work.
+
 \end{document}

### 2025-05-16 lfmtp/p.tex
finish

@@
 \newcommand{\Type}{\mathsf{Type}}
+% \NewCommandCopy{\oldGamma}{\Gamma}\renewcommand{\Gamma}{{\mathit{\oldGamma}}}
+% \NewCommandCopy{\oldDelta}{\Delta}\renewcommand{\Delta}{{\mathit{\oldDelta}}}
+% \NewCommandCopy{\oldTheta}{\Theta}\renewcommand{\Theta}{{\mathit{\oldTheta}}}
 
@@
 To derive the round-trips, we again need admissible equations. We prove
-the first round-trip equality in Appendix \ref{app:liftvar}, the other
+the first round-trip equality in Appendix \ref{app:tt}, the other
 is analogous.
@@
 certain type formers, then a CwF structure is derivable in any model.
-The idea is that
-\begin{itemize}
-\item using $\Sigma$ types we emulate contexts,
-\item using functions between these $\Sigma$ types we emulate parallel
-  substitutions,
-\item using the functions into the universe we emulate dependent types,
-\end{itemize}
-and then single substitutions are just there to set up $\Sigma$, $\Pi$
-and $\U$, and these types are enough to bootstrap the parallel
-substitution calculus.
+The idea is that (i) using $\Sigma$ types we emulate contexts; (ii)
+using functions between these $\Sigma$ types we emulate parallel
+substitutions; (iii) using the functions into the universe we emulate
+dependent types. Then single substitutions are just there to set up
+$\Sigma$, $\Pi$ and $\U$, and these types are enough to bootstrap the
+parallel substitution calculus.
 
@@
 \end{proof}
-In Appendix \ref{sec:functor}, we show that the roundtrip CwF
+In Appendix \ref{app:functor}, we show that the roundtrip CwF
 $\longrightarrow$ SSC

### 2025-06-08 lfmtp/p.tex
typo

@@
   $\alpha$-normal types.
-  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
+  \item $f\cdot a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
   $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and

### 2025-07-01 lfmtp/p.tex
stuff

@@
 
-\title{Type theory with single substitutions\thanks{This research was supported by the the HOTT ERC Grant 101170308.}}
+\title{Type theory with single substitutions\thanks{This research was supported by the HOTT ERC Grant 101170308.}}
 \author{Ambrus Kaposi
@@
 formal system based on Martin-Löf's original definition
-\cite{martinlof73predicative}, and not to the study of type systems
+\cite{MARTINLOF197573}, and not to the study of type systems
 (e.g.\ \cite{DBLP:books/daglib/0005958}).
@@
 categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
-\cite{10.1145/3204492}.
+\cite{DBLP:journals/tocl/Berg18}.
 
@@
 \cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
-theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
+theory \cite{DBLP:conf/mpc/AltenkirchBKT19,DBLP:phd/hal/Pujet22} support QIITs. For example,
 in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
@@
 \cite{glue}, and its special cases
-canonicity and parametricity \cite{kaposi-shallow}.
+canonicity and parametricity \cite{DBLP:conf/mpc/KaposiKK19}.
 
@@
 variable in a context, first of all we need an operation providing the last
-variable in the context. This is the zero De Bruijn index \cite{debruijn}, which we
+variable in the context. This is the zero De Bruijn index \cite{DEBRUIJN1972381}, which we
 denote by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
@@
 contextualisation \cite{DBLP:conf/fscd/BocquetKS23}. Following
-\cite{kaposi-shallow} we call it termification, as most sorts in the
+\cite{DBLP:conf/mpc/KaposiKK19} we call it termification, as most sorts in the
 new model are given by the sort of terms in the old model.
@@
 operations. A B-frame is given by sets $B_i$, $\tilde{B}_i$ and
-functions between them as shown below . \\
+functions between them as shown below. \\
 \begin{tikzpicture}
@@
 substitutions. Such a map is not definable only using single
-substitutions (a $\Sub$ is either a lifted weaking or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$).
+substitutions (a $\Sub$ is either a lifted weakening or a lifted single substitution, but we don't know anything about the relationship between $\Delta$ and $\Gamma$).
 However, the set of terms and types are still
